\section{Inverses of \(2\times2\) matrices}

The first non-trivial case is a \(2\times2\) matrix. It is small enough that the inverse can be written explicitly, but rich enough to show the main ideas: the determinant appears, the order of entries matters, and the inverse exists only under a clear condition.

Let
\[
A=\begin{pmatrix}a&b\\c&d\end{pmatrix}.
\]
We want to find a matrix
\[
A^{-1}=\begin{pmatrix}p&q\\r&s\end{pmatrix}
\]
such that
\[
AA^{-1}=I_2.
\]
Multiplying, we get
\[
\begin{pmatrix}a&b\\c&d\end{pmatrix}
\begin{pmatrix}p&q\\r&s\end{pmatrix}
=
\begin{pmatrix}
ap+br&aq+bs\\
cp+dr&cq+ds
\end{pmatrix}.
\]
For this to be the identity matrix, we need
\[
\begin{cases}
ap+br=1,\\
cp+dr=0,
\end{cases}
\qquad
\begin{cases}
aq+bs=0,\\
cq+ds=1.
\end{cases}
\]
Thus finding the inverse is already a problem about solving systems of linear equations. Instead of solving these systems every time, we use the known formula.

\begin{theorembox}
Let
\[
A=\begin{pmatrix}a&b\\c&d\end{pmatrix}.
\]
If
\[
\det A=ad-bc\ne0,
\]
then
\[
A^{-1}=\frac{1}{ad-bc}
\begin{pmatrix}
d&-b\\
-c&a
\end{pmatrix}.
\]
If \(ad-bc=0\), then \(A\) has no inverse.
\end{theorembox}

The formula has two visible parts. First, the determinant appears in the denominator. This is not accidental: if the determinant is zero, the inverse cannot exist. Second, the matrix
\[
\begin{pmatrix}
d&-b\\-c&a
\end{pmatrix}
\]
is obtained by switching the diagonal entries \(a,d\) and changing signs of the off-diagonal entries \(b,c\).

\begin{examplebox}
Find the inverse of
\[
A=\begin{pmatrix}2&1\\5&3\end{pmatrix}.
\]
First compute the determinant:
\[
\det A=2\cdot3-1\cdot5=6-5=1.
\]
Since the determinant is non-zero, the inverse exists. Therefore
\[
A^{-1}=\frac{1}{1}
\begin{pmatrix}
3&-1\\
-5&2
\end{pmatrix}
=
\begin{pmatrix}
3&-1\\
-5&2
\end{pmatrix}.
\]
We should verify the result:
\[
AA^{-1}=\begin{pmatrix}2&1\\5&3\end{pmatrix}
\begin{pmatrix}3&-1\\-5&2\end{pmatrix}
=
\begin{pmatrix}
6-5&-2+2\\
15-15&-5+6
\end{pmatrix}
=
\begin{pmatrix}1&0\\0&1\end{pmatrix}.
\]
\end{examplebox}

Verification is not a decorative step. It helps us remember the meaning of the inverse. We are not only producing a matrix; we are producing a matrix that must multiply with the original one to give the identity.

\begin{examplebox}
Consider
\[
B=\begin{pmatrix}1&2\\3&6\end{pmatrix}.
\]
The determinant is
\[
\det B=1\cdot6-2\cdot3=6-6=0.
\]
The formula cannot be used because it would require division by zero. More importantly, the inverse does not exist. The second row is three times the first row, so the matrix has lost independent information.
\end{examplebox}

This example shows why the determinant test is not just a technical condition. A zero determinant indicates structural collapse. There is no inverse because there is no unique way to undo the action of the matrix.

\subsection{A formula should still be read}

It is tempting to memorize the \(2\times2\) inverse formula mechanically. But it should be read as a statement with meaning:
\[
A^{-1}=\frac{1}{\det A}\begin{pmatrix}d&-b\\-c&a\end{pmatrix}.
\]
The determinant controls whether the inverse exists and how strongly the inverse scales the result. The second matrix rearranges the entries in the only way that makes multiplication return the identity.

\begin{summarybox}
For a \(2\times2\) inverse:
\begin{itemize}
\item compute \(\det A=ad-bc\),
\item if \(\det A=0\), stop: the matrix is not invertible,
\item if \(\det A\ne0\), use the inverse formula,
\item verify the result by multiplying with the original matrix.
\end{itemize}
\end{summarybox}
