\chapter{Quantization Dictionary}

\begin{remarkbox}
This appendix records the translation rules that are used repeatedly in the
main text. They should not be treated as mechanical replacements valid in every
context. They are starting points that must be checked against constraints,
symmetries, domains of operators, and gauge redundancies.
\end{remarkbox}

\section{Classical Mode to Quantum Oscillator}

A classical normal mode with amplitude \(q_{\vect{k}}(t)\) and frequency
\(\omega_{\vect{k}}\) is quantized like a harmonic oscillator. Schematically,
\[
    q_{\vect{k}},p_{\vect{k}}
    \quad \longrightarrow \quad
    \hat q_{\vect{k}},\hat p_{\vect{k}},
    \qquad
    [\hat q_{\vect{k}},\hat p_{\vect{k}'}]=i\delta_{\vect{k}\vect{k}'}.
\]
The oscillator is then rewritten in terms of ladder operators
\(\hat a_{\vect{k}}\) and \(\hat a_{\vect{k}}^\dagger\).

\section{Poisson Bracket to Commutator}

The canonical rule is
\[
    \{A,B\}_{\mathrm{PB}}
    \quad \longrightarrow \quad
    \frac{1}{i}[\hat A,\hat B].
\]
For fields, the equal-time canonical relation is
\[
    [\hat\phi(t,\vect{x}),\hat\pi(t,\vect{y})]
    = i\delta^{(3)}(\vect{x}-\vect{y}).
\]

\section{Classical Field to Field Operator}

A classical field \(\phi(x)\) becomes an operator-valued distribution
\(\hat\phi(x)\). The word ``distribution'' is important: products of field
operators at the same spacetime point are generally singular and require a
careful definition.

\section{Classical Green's Function to Propagator}

A classical Green's function solves a differential equation with a delta-function
source. A quantum propagator is a vacuum expectation value of a time-ordered
product,
\[
    \Delta_F(x-y)
    = \langle 0|T\{\hat\phi(x)\hat\phi(y)\}|0\rangle.
\]
It is related to Green's functions, but its physical meaning is quantum
correlation, not merely classical signal propagation.

\section{Classical Symmetry to Quantum Ward Identity}

A classical continuous symmetry gives a conserved current. In the quantum
theory, the same symmetry constrains correlation functions and scattering
amplitudes through Ward identities. If the symmetry fails after quantization,
the failure is called an anomaly.
