\chapter{Path Integral for Quantum Mechanics}

\begin{remarkbox}
The operator formalism led from fields to propagators, diagrams, and scattering
amplitudes. The path integral gives a different but equivalent viewpoint: a
transition amplitude is computed as a sum over histories, weighted by
\(e^{iS}\). We first develop this idea in ordinary quantum mechanics before
extending it to fields.
\end{remarkbox}

\section{Transition Amplitudes as Sums Over Histories}

In the operator formulation of quantum mechanics, the transition amplitude from
position \(q_i\) at time \(t_i\) to position \(q_f\) at time \(t_f\) is
\[
    \langle q_f,t_f|q_i,t_i\rangle
    = \langle q_f|e^{-i\hat H(t_f-t_i)}|q_i\rangle.
\]
The path-integral idea is that the same amplitude can be written as a sum over
all paths connecting the endpoints:
\[
    \langle q_f,t_f|q_i,t_i\rangle
    = \int_{q(t_i)=q_i}^{q(t_f)=q_f}\mathcal{D}q(t)\,
      e^{iS[q]}.
\]
Here
\[
    S[q]=\int_{t_i}^{t_f}\dd t\,L(q,\dot q)
\]
is the classical action evaluated on the path \(q(t)\). The notation
\(\mathcal{D}q(t)\) means integration over all intermediate paths.

This formula should not be interpreted as saying that the particle secretly
chooses one path. Rather, the transition amplitude is built by coherently
summing contributions from all paths.

\section{Time Slicing}

The path integral can be derived by dividing the time interval into many small
steps. Let
\[
    t_f-t_i=N\epsilon.
\]
Insert resolutions of the identity in the position basis:
\[
    1=\int\dd q_j\,|q_j\rangle\langle q_j|.
\]
Then
\[
    \langle q_f|e^{-i\hat H(t_f-t_i)}|q_i\rangle
    = \int\prod_{j=1}^{N-1}\dd q_j\,
      \prod_{j=0}^{N-1}\langle q_{j+1}|e^{-i\epsilon\hat H}|q_j\rangle.
\]
For a Hamiltonian
\[
    H=\frac{p^2}{2m}+V(q),
\]
one also inserts momentum states and evaluates the short-time matrix element.
In the limit \(N\to\infty\), this produces
\[
    \int\mathcal{D}q\,e^{i\int\dd t\,[\frac{1}{2}m\dot q^2-V(q)]}.
\]
Thus the path integral is not a new postulate here; it is a reorganization of
operator time evolution.

\section{Classical Limit and Stationary Phase}

The factor \(e^{iS[q]}\) oscillates rapidly when the action is large compared
with \(\hbar\). Restoring \(\hbar\), the weight is
\[
    e^{iS[q]/\hbar}.
\]
In the limit \(\hbar\to0\), most paths interfere destructively. The dominant
contribution comes from paths for which the action is stationary:
\[
    \delta S=0.
\]
This condition gives the classical equations of motion.

Thus the classical path is not the only path included in the quantum amplitude.
It is the path around which neighbouring contributions add coherently in the
classical limit. Quantum mechanics is the sum over all histories; classical
mechanics is the stationary-phase approximation.

\section{Gaussian Integrals}

Gaussian integrals are the technical backbone of free path integrals. The
ordinary one-dimensional formula is
\[
    \int_{-\infty}^{\infty}\dd x\,
    e^{-\frac{1}{2}ax^2+Jx}
    = \sqrt{\frac{2\pi}{a}}\,
      e^{\frac{J^2}{2a}},
\]
for \(a>0\). In oscillatory form one uses an \(i\epsilon\) prescription to make
the integral well defined.

The multidimensional version is
\[
    \int \dd^n x\,
    \exp\left[-\frac{1}{2}x^TAx+J^Tx\right]
    = \left(\frac{(2\pi)^n}{\det A}\right)^{1/2}
      \exp\left(\frac{1}{2}J^TA^{-1}J\right).
\]
Free quantum theories lead to infinite-dimensional versions of this formula.
The inverse operator \(A^{-1}\) becomes the propagator.

\section{The Free Particle}

For a free particle,
\[
    L=\frac{1}{2}m\dot q^2.
\]
The path integral is Gaussian. The result is
\[
    K(q_f,t_f;q_i,t_i)
    = \left(\frac{m}{2\pi iT}\right)^{1/2}
      \exp\left[\frac{im(q_f-q_i)^2}{2T}\right],
\]
where \(T=t_f-t_i\). The exponent is the classical action evaluated on the
straight-line path connecting the endpoints.

This example illustrates a common pattern. For a quadratic action, the path
integral can be evaluated exactly. The answer is a normalization determinant
multiplying \(e^{iS_{\mathrm{cl}}}\).

\section{The Harmonic Oscillator}

For the harmonic oscillator,
\[
    L=\frac{1}{2}m\dot q^2-\frac{1}{2}m\omega^2q^2.
\]
The action is still quadratic, so the path integral remains Gaussian. The
transition amplitude again has the form
\[
    K(q_f,t_f;q_i,t_i)
    = \mathcal{N}(T)e^{iS_{\mathrm{cl}}(q_f,t_f;q_i,t_i)}.
\]
The exact normalization \(\mathcal{N}(T)\) comes from the determinant of the
quadratic fluctuation operator.

The oscillator is important because a free field is a collection of oscillators.
The free-field path integral will therefore be an infinite product of Gaussian
oscillator path integrals.

\section{Generating Functionals in Quantum Mechanics}

To compute correlation functions, introduce a source \(J(t)\) coupled to
\(q(t)\):
\[
    Z[J]
    = \int\mathcal{D}q\,
      \exp\left(iS[q]+i\int\dd t\,J(t)q(t)\right).
\]
The source is an auxiliary function. It is set to zero after differentiating.
Correlation functions are obtained by functional derivatives:
\[
    \langle T q(t_1)q(t_2)\rangle
    = \frac{1}{Z[0]}
      \left.\frac{1}{i^2}\frac{\delta^2 Z[J]}{\delta J(t_1)\delta J(t_2)}\right|_{J=0}.
\]
This is the prototype of the generating functional method used in QFT.

\section{Sources and Correlation Functions}

The source term keeps track of insertions. Differentiating once gives
\[
    \frac{1}{i}\frac{\delta Z[J]}{\delta J(t)}
    = \int\mathcal{D}q\,q(t)
      \exp\left(iS[q]+i\int\dd t'\,J(t')q(t')\right).
\]
Differentiating twice inserts two factors of \(q\), and so on. Therefore
\[
    \left.\frac{1}{i^n}\frac{\delta^n Z[J]}{\delta J(t_1)\cdots\delta J(t_n)}\right|_{J=0}
\]
generates the \(n\)-point time-ordered correlation function.

In operator language these are vacuum expectation values of time-ordered
products. In the path-integral language they are moments of a functional
integral with an oscillatory weight.

\section{Why the Path Integral Is Useful for QFT}

The path integral is especially useful in field theory for several reasons.
First, it makes spacetime symmetries more visible because the action appears
directly. Second, it gives a compact way to derive Feynman rules. Third, it
handles correlation functions and sources naturally. Fourth, it generalizes to
gauge fixing, ghosts, effective actions, and nonperturbative approximations.

For a field, the single coordinate \(q(t)\) is replaced by a field configuration
\(\phi(x)\). The ordinary source \(J(t)\) is replaced by a spacetime source
\(J(x)\). The path integral becomes
\[
    \int\mathcal{D}\phi\,e^{iS[\phi]}.
\]
This is the subject of the next chapter.

\section{Relation to Canonical Quantization}

The path integral and canonical quantization are not separate quantum theories.
They are two formulations of the same theory. Canonical quantization emphasizes
Hilbert spaces, operators, commutators, and states. The path integral emphasizes
action functionals, histories, sources, and correlation functions.

The connection can be summarized as
\[
    \langle q_f|e^{-i\hat H T}|q_i\rangle
    = \int_{q_i}^{q_f}\mathcal{D}q\,e^{iS[q]}.
\]
The left-hand side is canonical. The right-hand side is the path integral. In
field theory, this equivalence underlies the use of functional integrals to
compute the same time-ordered correlators that appear in the operator formalism.

\begin{summarybox}
The path integral expresses a transition amplitude as a sum over histories
weighted by \(e^{iS}\). Time slicing connects this formula to operator evolution.
The classical equations arise by stationary phase. Quadratic actions lead to
Gaussian integrals, whose inverses become propagators. Adding sources produces a
generating functional whose derivatives give time-ordered correlation functions.
\end{summarybox}

\section{Chapter Checklist}

After reading this chapter, one should be able to:
\begin{itemize}
    \item write a transition amplitude as a path integral over histories;
    \item explain the time-slicing derivation;
    \item explain how the classical limit follows from stationary phase;
    \item evaluate the role of Gaussian integrals in free theories;
    \item state the path integral for the free particle and harmonic oscillator;
    \item define a generating functional with a source;
    \item obtain correlation functions by differentiating with respect to the
    source;
    \item explain why the path integral is useful for QFT;
    \item relate the path integral to canonical quantization.
\end{itemize}
