\chapter{BRST Symmetry}

\begin{remarkbox}
Gauge fixing is necessary for quantizing gauge theories, but it hides the
original gauge symmetry. BRST symmetry is the remnant symmetry that survives
gauge fixing. It organizes ghosts, selects physical states, and leads to the
Slavnov--Taylor identities that keep non-Abelian gauge theory consistent.
\end{remarkbox}

\section{Why BRST Symmetry Is Needed}

In a gauge theory, many field configurations represent the same physical state.
Gauge fixing removes this redundancy so that propagators and Feynman rules are
well defined. But gauge fixing also introduces unphysical ingredients:
longitudinal gauge polarizations, timelike components, auxiliary fields, and
Faddeev--Popov ghosts.

The central question is therefore:
\[
    \text{How can a gauge-fixed theory with unphysical fields describe physical
    gauge-invariant observables?}
\]
BRST symmetry gives the answer. It replaces the original local gauge redundancy
by a global fermionic symmetry acting on gauge fields, matter fields, ghosts,
antighosts, and auxiliary fields.

The power of BRST symmetry lies in one algebraic property:
\[
    s^2=0.
\]
The BRST operator is nilpotent. This nilpotency makes cohomology possible, and
cohomology is the mathematical language used to define physical states.

\section{The BRST Operator}

Denote the BRST transformation by \(s\). It is an anticommuting operation, which
means that it changes Grassmann parity. Acting on a product, it obeys a graded
Leibniz rule:
\[
    s(XY)=(sX)Y+(-1)^{|X|}X(sY),
\]
where \(|X|=0\) for bosonic \(X\) and \(|X|=1\) for fermionic \(X\).

For a non-Abelian gauge theory, the basic fields are
\[
    A_\mu^a,
    \qquad
    c^a,
    \qquad
    \bar c^a,
    \qquad
    B^a.
\]
Here \(c^a\) is the ghost, \(\bar c^a\) is the antighost, and \(B^a\) is the
Nakanishi--Lautrup auxiliary field.

The BRST transformations are
\[
    sA_\mu^a=D_\mu^{ab}c^b,
\]
\[
    sc^a=-\frac{g}{2}f^{abc}c^b c^c,
\]
\[
    s\bar c^a=B^a,
    \qquad
    sB^a=0.
\]
For matter fields in a representation with generators \(T^a\), one writes
schematically
\[
    s\psi=igc^aT^a\psi,
    \qquad
    s\bar\psi=-ig\bar\psi c^aT^a.
\]
These transformations look like gauge transformations whose parameter has been
replaced by the ghost field.

\section{Nilpotency}

The defining property of BRST symmetry is nilpotency:
\[
    s^2=0.
\]
For the antighost and auxiliary field this is immediate:
\[
    s^2\bar c^a=sB^a=0,
    \qquad
    s^2B^a=0.
\]
For the gauge field and ghost, nilpotency follows from the Lie algebra and the
Jacobi identity. The ghost transformation
\[
    sc^a=-\frac{g}{2}f^{abc}c^b c^c
\]
is precisely what is needed to compensate the noncommutativity of two gauge
transformations.

Nilpotency means that BRST transformations behave like an exterior derivative
on the space of fields. This is why the language of cohomology appears naturally.

\section{Gauge Fixing as a BRST-Exact Term}

The gauge-fixing and ghost terms can be written as a BRST variation. Choose a
gauge-fixing fermion
\[
    \Psi=\int\dd^4x\,\bar c^a
    \left(\partial^\mu A_\mu^a+\frac{\xi}{2}B^a\right).
\]
Then define
\[
    S_{\mathrm{gf+gh}}=s\Psi.
\]
Using the BRST transformations gives
\[
    \mathcal{L}_{\mathrm{gf+gh}}
    = B^a\partial^\mu A_\mu^a+\frac{\xi}{2}B^aB^a
      -\bar c^a\partial^\mu D_\mu^{ab}c^b.
\]
Eliminating \(B^a\) by its algebraic equation of motion gives the familiar
covariant gauge-fixing term and the ghost Lagrangian:
\[
    \mathcal{L}_{\mathrm{gf}}
    = -\frac{1}{2\xi}(\partial^\mu A_\mu^a)^2,
\]
\[
    \mathcal{L}_{\mathrm{gh}}
    = -\bar c^a\partial^\mu D_\mu^{ab}c^b,
\]
up to sign conventions.

Because \(s^2=0\), the BRST variation of \(S_{\mathrm{gf+gh}}\) vanishes:
\[
    sS_{\mathrm{gf+gh}}=s^2\Psi=0.
\]
The gauge-fixed action is therefore BRST invariant.

\section{Ghost Number}

BRST symmetry comes with a grading called ghost number. The assignments are
\[
    \operatorname{gh}(A_\mu^a)=0,
    \qquad
    \operatorname{gh}(\psi)=0,
\]
\[
    \operatorname{gh}(c^a)=1,
    \qquad
    \operatorname{gh}(\bar c^a)=-1,
    \qquad
    \operatorname{gh}(B^a)=0.
\]
The BRST operator increases ghost number by one:
\[
    \operatorname{gh}(sX)=\operatorname{gh}(X)+1.
\]
Physical observables usually have ghost number zero. Ghost number helps organize
which operators can appear in identities and which terms can mix under
renormalization.

\section{The BRST Charge}

In canonical language, a continuous symmetry is generated by a conserved charge.
BRST symmetry is generated by a fermionic charge \(Q_{\mathrm{BRST}}\). Acting on
an operator \(\mathcal{O}\), one writes schematically
\[
    s\mathcal{O}=i[Q_{\mathrm{BRST}},\mathcal{O}\}_{\pm},
\]
where the bracket is graded: a commutator for bosonic operators and an
anticommutator for fermionic operators.

Nilpotency becomes
\[
    Q_{\mathrm{BRST}}^2=0.
\]
This equation is the operator version of \(s^2=0\). It is the basis for defining
physical states through cohomology.

\section{Physical States as Cohomology}

A physical state is required to be BRST closed:
\[
    Q_{\mathrm{BRST}}|\mathrm{phys}\rangle=0.
\]
However, a state of the form
\[
    |\chi\rangle=Q_{\mathrm{BRST}}|\lambda\rangle
\]
should be regarded as physically trivial. Such a state is BRST exact. Since
\[
    Q_{\mathrm{BRST}}|\chi\rangle=Q_{\mathrm{BRST}}^2|\lambda\rangle=0,
\]
BRST-exact states are automatically BRST closed.

The physical Hilbert space is therefore the cohomology:
\[
    \mathcal{H}_{\mathrm{phys}}
    = \frac{\ker Q_{\mathrm{BRST}}}{\operatorname{im}Q_{\mathrm{BRST}}}.
\]
This quotient removes unphysical gauge and ghost degrees of freedom while
retaining gauge-invariant physical content.

\section{Gauge-Invariant Operators}

Gauge-invariant operators are BRST closed. If \(\mathcal{O}\) is gauge
invariant, then
\[
    s\mathcal{O}=0.
\]
Examples include traces of field strengths, Wilson loops, and color-singlet
operators in non-Abelian gauge theory.

BRST-exact operators have vanishing matrix elements between physical states under
appropriate assumptions. If
\[
    \mathcal{O}=sX,
\]
then its physical matrix elements vanish because \(Q_{\mathrm{BRST}}\) kills
physical states and exact contributions are treated as gauge artifacts.

Thus BRST cohomology is not only a statement about states. It also classifies
physical observables.

\section{Slavnov--Taylor Identities}

In QED, Ward identities follow from gauge invariance. In non-Abelian gauge
theory, the corresponding identities are Slavnov--Taylor identities. They follow
from BRST invariance of the gauge-fixed path integral.

Schematically, if the action and functional measure are BRST invariant, then a
change of variables by a BRST transformation gives
\[
    \langle s\mathcal{O}\rangle=0
\]
for suitable operators \(\mathcal{O}\). Choosing \(\mathcal{O}\) appropriately
relates different Green functions.

These identities constrain the renormalization of gauge theories. They ensure
that counterterms preserve the gauge structure and that unphysical degrees of
freedom cancel in physical amplitudes.

\section{Gauge-Parameter Independence}

The gauge parameter \(\xi\) appears in propagators and intermediate Green
functions. Physical quantities must not depend on it. BRST symmetry explains why.
Changing \(\xi\) changes the gauge-fixing fermion \(\Psi\). Since the
gauge-fixing action is BRST exact,
\[
    S_{\mathrm{gf+gh}}=s\Psi,
\]
a change in \(\Psi\) changes the action by a BRST-exact term.

Matrix elements of BRST-exact terms vanish between physical states. Therefore
physical observables are independent of the gauge-fixing choice, even though
intermediate expressions are not.

\section{Unitarity and Negative-Norm States}

Covariant gauges include unphysical polarizations. Some of these modes would
lead to negative-norm states if treated as physical. Ghosts also appear in the
formalism but cannot be external particles.

BRST symmetry organizes cancellations among these unphysical contributions. The
physical \(S\)-matrix is unitary on the BRST cohomology, not on the enlarged
auxiliary state space.

This distinction is crucial. The gauge-fixed theory may contain unphysical
fields, but the physical sector selected by BRST cohomology contains only the
allowed gauge-invariant degrees of freedom.

\section{BRST and Renormalization}

Renormalization must preserve BRST symmetry. Counterterms cannot be chosen
arbitrarily; they must respect the Slavnov--Taylor identities. This requirement
ensures that the renormalized theory remains a quantum gauge theory rather than
a generic theory of vector fields and ghosts.

In practice, BRST symmetry strongly restricts the form of allowed counterterms.
It relates field renormalizations, coupling renormalization, gauge-fixing terms,
and ghost interactions. This is why non-Abelian gauge theory remains
renormalizable despite its complicated interactions.

If BRST symmetry is broken by the regulator or by quantum effects, one must check
whether it can be restored by counterterms. If it cannot, the theory has an
anomaly and is inconsistent as a gauge theory.

\section{BRST Anomalies}

An anomaly is a failure of a classical symmetry to survive quantization. In the
BRST language, a gauge anomaly is an obstruction to maintaining BRST invariance
of the quantum effective action.

If the BRST symmetry is anomalous, the Slavnov--Taylor identities fail. Then
unphysical modes may not decouple, and the theory may lose unitarity or gauge
independence.

This is why anomaly cancellation is a consistency condition in gauge theories.
The Standard Model matter content is arranged so that gauge anomalies cancel.
Anomalies will be studied more explicitly in a later chapter.

\section{Conceptual Summary}

BRST symmetry can be summarized in one line:
\[
    \text{gauge symmetry after gauge fixing} = \text{nilpotent BRST symmetry}.
\]
The original gauge redundancy is no longer manifest in the same way after gauge
fixing. Instead, it is encoded in a global fermionic symmetry involving ghosts.

The nilpotent BRST operator lets us separate physical information from gauge
artifacts. Closed but non-exact objects represent physical content. Exact objects
represent redundancy. This is the cohomological structure at the heart of
quantum gauge theory.

\begin{summarybox}
BRST symmetry is the nilpotent remnant of gauge symmetry after gauge fixing. It
acts on gauge fields, ghosts, antighosts, auxiliary fields, and matter fields.
Gauge-fixing and ghost terms can be written as BRST-exact terms. Physical states
are BRST cohomology classes: closed states modulo exact states. Slavnov--Taylor
identities, gauge-parameter independence, unitarity of the physical \(S\)-matrix,
and consistency of renormalization all follow from BRST symmetry.
\end{summarybox}

\section{Chapter Checklist}

After reading this chapter, one should be able to:
\begin{itemize}
    \item explain why gauge fixing requires a residual organizing symmetry;
    \item write the basic BRST transformations for Yang--Mills theory;
    \item explain the nilpotency condition \(s^2=0\);
    \item write gauge fixing as a BRST-exact term;
    \item define ghost number;
    \item explain the BRST charge and its nilpotency;
    \item define physical states as BRST cohomology;
    \item explain why gauge-invariant operators are BRST closed;
    \item state the origin of Slavnov--Taylor identities;
    \item explain why physical observables are gauge-parameter independent;
    \item describe the relation between BRST symmetry, unitarity, and anomalies.
\end{itemize}
