\chapter{Scattering Theory and the \texorpdfstring{\(S\)-Matrix}{S-Matrix}}

\begin{remarkbox}
The previous chapters built the machinery of perturbative QFT: free fields,
propagators, the interaction picture, Wick's theorem, and Feynman diagrams.
Scattering theory explains how this machinery becomes physics. It connects
correlation functions and diagrams to transition amplitudes, cross sections, and
decay rates.
\end{remarkbox}

\section{In-States and Out-States}

Scattering theory compares what comes in from the far past with what goes out in
the far future. The basic idealization is that particles are well separated at
very early and very late times, so that they may be described by free Fock
states.

An initial state is called an in-state and is denoted
\[
    |\alpha,\mathrm{in}\rangle.
\]
A final state is called an out-state and is denoted
\[
    |\beta,\mathrm{out}\rangle.
\]
For example, a two-particle initial state may be labelled by momenta and other
quantum numbers,
\[
    |p_1,p_2,\mathrm{in}\rangle,
\]
while a final state may contain different momenta or even a different number of
particles.

The central physical question is: what is the amplitude for the initial state
\(\alpha\) to become the final state \(\beta\)?

\section{The \texorpdfstring{\(S\)-Matrix}{S-Matrix}}

The scattering matrix is defined by the overlap
\[
    S_{\beta\alpha}
    = \langle\beta,\mathrm{out}|\alpha,\mathrm{in}\rangle.
\]
Equivalently, one introduces an operator \(\hat S\) acting on free asymptotic
states:
\[
    S_{\beta\alpha}
    = \langle\beta|\hat S|\alpha\rangle.
\]
In the interaction picture,
\[
    \hat S
    = T\exp\left(-i\int_{-\infty}^{\infty}\dd t\,\hat H_I(t)\right).
\]
This is the same operator introduced at the end of Chapter 7. The novelty here
is that its matrix elements are interpreted as scattering amplitudes.

It is useful to separate the identity contribution from the interacting part:
\[
    \hat S = 1+i\hat T.
\]
The identity term describes no scattering. The operator \(\hat T\) contains the
nontrivial transition information.

\section{Transition Amplitudes}

For relativistic scattering, momentum conservation is factored out. One writes
\[
    \langle\beta|i\hat T|\alpha\rangle
    = i(2\pi)^4\delta^{(4)}(P_\beta-P_\alpha)\mathcal{M}_{\beta\alpha}.
\]
The quantity \(\mathcal{M}_{\beta\alpha}\) is the invariant matrix element, or
scattering amplitude. It is what Feynman rules are designed to compute.

The delta function expresses conservation of total four-momentum. The amplitude
\(\mathcal{M}\) contains the dynamical information: coupling constants,
propagators, spinor factors, polarization vectors, and momentum dependence.

For example, in a tree-level scalar process, \(\mathcal{M}\) may be a simple
polynomial in the coupling. In QED it includes spinor bilinears and photon
propagators. In loop calculations it includes integrals over internal momenta.

\section{Matrix Elements and Momentum Conservation}

Translation invariance implies four-momentum conservation. If the theory is
invariant under spacetime translations, then the total four-momentum operator
commutes with the scattering operator:
\[
    [\hat P^\mu,\hat S]=0.
\]
Therefore only transitions satisfying
\[
    P_\alpha^\mu=P_\beta^\mu
\]
can have nonzero amplitude.

This is why every Feynman diagram in momentum space carries a momentum-conserving
delta function at each vertex. After all internal vertex integrations are done,
one overall delta function remains. It expresses total energy-momentum
conservation for the full scattering process.

\section{Lorentz-Invariant Phase Space}

Probabilities are obtained by squaring amplitudes and summing or integrating
over final states. The Lorentz-invariant one-particle phase-space measure is
\[
    \frac{\dd^3p}{(2\pi)^3}\frac{1}{2E_{\vect{p}}}.
\]
For an \(n\)-particle final state with total incoming momentum \(P\), the
Lorentz-invariant phase space is
\[
    \dd\Pi_n(P;p_1,\ldots,p_n)
    = (2\pi)^4\delta^{(4)}\left(P-\sum_{i=1}^{n}p_i\right)
      \prod_{i=1}^{n}\frac{\dd^3p_i}{(2\pi)^3 2E_i}.
\]
This measure combines the on-shell conditions for final particles with total
four-momentum conservation.

The same measure appears in cross sections and decay rates. It is one of the
standard bridges between amplitudes and measurable predictions.

\section{Cross Sections}

For a two-particle scattering process
\[
    a+b\longrightarrow 1+2+\cdots+n,
\]
the differential cross section has the schematic form
\[
    \dd\sigma
    = \frac{1}{\text{flux}}\,|\mathcal{M}|^2\,\dd\Pi_n.
\]
The flux factor depends on the normalization of the incoming states. In a
Lorentz-invariant form for two incoming particles with momenta \(p_a,p_b\), it
is built from
\[
    4\sqrt{(p_a\cdot p_b)^2-m_a^2m_b^2}.
\]
Thus
\[
    \dd\sigma
    = \frac{|\mathcal{M}|^2}{4\sqrt{(p_a\cdot p_b)^2-m_a^2m_b^2}}\,\dd\Pi_n.
\]
For massless or center-of-mass examples this expression simplifies, but the
structure is always the same: amplitude squared times phase space divided by
incoming flux.

\section{Decay Rates}

For a decay process
\[
    A\longrightarrow 1+2+\cdots+n,
\]
there is only one initial particle. The differential decay rate is
\[
    \dd\Gamma
    = \frac{1}{2M_A}|\mathcal{M}|^2\,\dd\Pi_n,
\]
where \(M_A\) is the mass of the decaying particle. The total decay rate is
obtained by integrating over final-state phase space:
\[
    \Gamma = \int \dd\Gamma.
\]
The lifetime is
\[
    \tau = \frac{1}{\Gamma}.
\]
The same invariant amplitude technology therefore describes both scattering
experiments and unstable-particle decays.

\section{The LSZ Reduction Formula}

Perturbation theory naturally computes time-ordered correlation functions. The
LSZ reduction formula explains how to extract scattering amplitudes from those
correlators.

For a scalar field, the rough idea is this: external particles correspond to
poles of the Fourier-transformed correlation function. To obtain an amplitude,
one multiplies by inverse propagator factors and takes the external momenta on
shell. Schematically,
\[
    \mathcal{M}
    \sim \prod_{i}\left(p_i^2-m^2\right)
    \widetilde G(p_1,\ldots,p_n)\bigg|_{p_i^2=m^2},
\]
with normalization factors suppressed.

The exact LSZ formula includes wave-function renormalization factors and signs
that depend on conventions. Its conceptual message is more important at this
stage: scattering amplitudes are obtained from the pole structure of
correlation functions.

\section{From Correlation Functions to Scattering Amplitudes}

The chain of ideas is now complete:
\[
    \text{Lagrangian}
    \longrightarrow
    \text{interaction Hamiltonian}
    \longrightarrow
    \text{Dyson series}
    \longrightarrow
    \text{Wick contractions}
    \longrightarrow
    \text{Feynman diagrams}
    \longrightarrow
    \mathcal{M}
    \longrightarrow
    \text{observables}.
\]
The first part of the chain is formal and operator-theoretic. The middle part is
diagrammatic. The last part converts amplitudes into cross sections and decay
rates.

This is why propagators and vertices are not merely mathematical decorations.
They are the intermediate objects that connect a Lagrangian to measurable
numbers.

\section{Tree-Level \texorpdfstring{\(2\to2\)}{2 to 2} Scattering}

A simple example is tree-level \(2\to2\) scattering in \(\phi^4\) theory. The
interaction is
\[
    \mathcal{L}_{\mathrm{int}}
    = -\frac{\lambda}{4!}\phi^4.
\]
At lowest order, there is one four-point vertex and no internal propagator. The
amplitude is
\[
    i\mathcal{M}=-i\lambda,
    \qquad
    \mathcal{M}=-\lambda.
\]
This example is deliberately simple. It shows that a contact interaction gives a
constant tree-level amplitude. More complicated theories produce amplitudes with
propagators, spinor factors, and polarization vectors.

In momentum space, the full matrix element is
\[
    \langle p_3,p_4|i\hat T|p_1,p_2\rangle
    = i(2\pi)^4\delta^{(4)}(p_1+p_2-p_3-p_4)\mathcal{M}.
\]
The delta function enforces momentum conservation; \(\mathcal{M}\) is the part
used in the cross-section formula.

\section{Decay of an Unstable Particle}

Suppose a particle \(A\) can decay into two particles \(1\) and \(2\). If the
amplitude is \(\mathcal{M}(A\to1+2)\), then
\[
    \dd\Gamma
    = \frac{1}{2M_A}|\mathcal{M}|^2
      (2\pi)^4\delta^{(4)}(p_A-p_1-p_2)
      \frac{\dd^3p_1}{(2\pi)^3 2E_1}
      \frac{\dd^3p_2}{(2\pi)^3 2E_2}.
\]
This formula is the decay analogue of the scattering cross section. It says that
rates are determined by amplitude squared times available phase space.

The phase space can strongly affect the physical rate. A large amplitude may
produce a small rate if very little phase space is available, while a modest
amplitude can produce a large rate if many final states are accessible.

\section{From Amplitudes to Measurable Predictions}

Experiments do not measure Feynman diagrams directly. They measure cross
sections, angular distributions, branching ratios, decay widths, and event
rates. QFT predicts these quantities by computing amplitudes and integrating them
over the appropriate phase space.

A realistic prediction usually requires additional steps: spin sums, averages
over initial polarizations, sums over final polarizations, symmetry factors for
identical particles, detector cuts, and sometimes radiative corrections. The
basic structure, however, is already visible:
\[
    \text{theory} \longrightarrow \mathcal{M}
    \longrightarrow |\mathcal{M}|^2
    \longrightarrow \text{phase-space integral}
    \longrightarrow \text{observable}.
\]

\begin{summarybox}
Scattering theory turns the operator and diagrammatic machinery of perturbative
QFT into measurable predictions. The \(S\)-matrix connects in-states to
out-states. Momentum conservation is factored out to define the invariant
amplitude \(\mathcal{M}\). Cross sections and decay rates are obtained from
\(|\mathcal{M}|^2\) integrated over Lorentz-invariant phase space. LSZ explains
why scattering amplitudes are encoded in the pole structure of time-ordered
correlation functions.
\end{summarybox}

\section{Chapter Checklist}

After reading this chapter, one should be able to:
\begin{itemize}
    \item define in-states, out-states, and the \(S\)-matrix;
    \item distinguish \(S\), \(T\), and the invariant amplitude \(\mathcal{M}\);
    \item explain the origin of the momentum-conserving delta function;
    \item write the Lorentz-invariant phase-space measure;
    \item state the schematic cross-section formula;
    \item state the decay-rate formula;
    \item explain the conceptual role of LSZ reduction;
    \item describe how correlation functions lead to amplitudes;
    \item compute the tree-level \(\phi^4\) amplitude at lowest order;
    \item explain how amplitudes become measurable predictions.
\end{itemize}
