\chapter{Quantum Electrodynamics}

\begin{remarkbox}
Quantum electrodynamics is the quantum field theory of charged Dirac matter and
the electromagnetic field. It combines the Dirac field, gauge invariance,
photons, fermion spinor chains, and perturbation theory into the first realistic
relativistic quantum field theory of the course.
\end{remarkbox}

\section{QED as a Gauge Theory}

The free Dirac Lagrangian is
\[
    \mathcal{L}_{\mathrm{Dirac}}
    = \bar\psi(i\gamma^\mu\partial_\mu-m)\psi.
\]
It is invariant under the global phase transformation
\[
    \psi\longrightarrow e^{i\alpha}\psi.
\]
QED is obtained by promoting this symmetry to a local one,
\[
    \psi(x)\longrightarrow e^{i\alpha(x)}\psi(x).
\]
Ordinary derivatives do not transform covariantly under a local phase rotation,
so one introduces the covariant derivative
\[
    D_\mu=\partial_\mu+ieA_\mu.
\]
The gauge field transforms as
\[
    A_\mu\longrightarrow A_\mu-\frac{1}{e}\partial_\mu\alpha.
\]
The QED Lagrangian is
\[
    \mathcal{L}_{\mathrm{QED}}
    = -\frac{1}{4}F_{\mu\nu}F^{\mu\nu}
      +\bar\psi(i\gamma^\mu D_\mu-m)\psi.
\]
Expanding the covariant derivative gives
\[
    \mathcal{L}_{\mathrm{QED}}
    = -\frac{1}{4}F_{\mu\nu}F^{\mu\nu}
      +\bar\psi(i\gamma^\mu\partial_\mu-m)\psi
      -e\bar\psi\gamma^\mu\psi A_\mu.
\]
The last term is the interaction between charged matter and photons.

\section{The Electromagnetic Current}

The interaction can be written as
\[
    \mathcal{L}_{\mathrm{int}}
    = -j^\mu A_\mu,
\]
where
\[
    j^\mu=e\bar\psi\gamma^\mu\psi.
\]
The current is conserved:
\[
    \partial_\mu j^\mu=0,
\]
using the Dirac equation. Current conservation is deeply connected to gauge
invariance. In perturbation theory it leads to Ward identities, which enforce
cancellations of unphysical photon polarizations.

The electric charge operator counts electrons and positrons with opposite signs.
With the convention that the electron has charge \(-e\), one may write
schematically
\[
    \hat Q_{\mathrm{em}}
    = -e\sum_s\int\frac{\dd^3p}{(2\pi)^3}
      \left(
          \hat b_s^\dagger(\vect{p})\hat b_s(\vect{p})
          -\hat d_s^\dagger(\vect{p})\hat d_s(\vect{p})
      \right).
\]

\section{Gauge Fixing and the Photon Propagator}

To define a photon propagator, one must fix a gauge. In covariant gauges, add
\[
    \mathcal{L}_{\mathrm{gf}}
    = -\frac{1}{2\xi}(\partial_\mu A^\mu)^2.
\]
The photon propagator is then
\[
    D_{\mu\nu}(k)
    = \frac{-i}{k^2+i\epsilon}
      \left(\eta_{\mu\nu}-(1-\xi)\frac{k_\mu k_\nu}{k^2+i\epsilon}\right).
\]
In Feynman gauge, \(\xi=1\), this becomes
\[
    D_{\mu\nu}(k)=\frac{-i\eta_{\mu\nu}}{k^2+i\epsilon}.
\]
The gauge parameter \(\xi\) may appear in intermediate expressions, but physical
observables must be independent of it.

In Abelian QED, the Faddeev--Popov ghosts decouple in linear covariant gauges,
so they do not appear in ordinary QED Feynman rules. This is special to Abelian
gauge theory.

\section{QED Feynman Rules}

The basic free propagators are the electron propagator
\[
    \frac{i(\gamma^\mu p_\mu+m)}{p^2-m^2+i\epsilon}
\]
and the photon propagator, for example in Feynman gauge,
\[
    \frac{-i\eta_{\mu\nu}}{k^2+i\epsilon}.
\]
The electron--photon vertex is
\[
    -ie\gamma^\mu.
\]
Momentum is conserved at each vertex. External electron and positron lines are
represented by spinors \(u,\bar u,v,\bar v\). External photons are represented by
polarization vectors \(\epsilon_\mu(k)\).

The fine-structure constant is
\[
    \alpha=\frac{e^2}{4\pi}.
\]
Perturbation theory in QED is an expansion in powers of \(e\), or equivalently
in powers of \(\alpha\) for probabilities.

\section{Spinor Chains}

QED amplitudes contain spinor chains. A typical factor has the form
\[
    \bar u(p')\gamma^\mu u(p).
\]
This is the matrix element of the electromagnetic current between electron
states. For processes with internal fermion propagators, longer chains appear,
for example
\[
    \bar u(p')\gamma^\mu
    \frac{i(\gamma^\rho q_\rho+m)}{q^2-m^2+i\epsilon}
    \gamma^\nu u(p).
\]
Spinor chains are evaluated using gamma-matrix algebra and spin sums. For
unpolarized scattering, one averages over initial spins and sums over final
spins. The completeness relations are
\[
    \sum_s u_s(p)\bar u_s(p)=\gamma^\mu p_\mu+m,
\]
and
\[
    \sum_s v_s(p)\bar v_s(p)=\gamma^\mu p_\mu-m.
\]
These identities turn squared amplitudes into traces of gamma matrices.

\section{Ward Identities}

Gauge invariance implies Ward identities. The simplest tree-level statement is
that replacing an external photon polarization vector by its momentum gives zero
for a physical amplitude:
\[
    \epsilon_\mu(k)\longrightarrow k_\mu
    \qquad\Longrightarrow\qquad
    \mathcal{M}\longrightarrow0.
\]
This expresses the fact that unphysical longitudinal photon polarizations do not
contribute to physical processes.

At the level of correlation functions, Ward identities relate current
conservation to identities among Green functions. At loop level, they constrain
renormalization constants and protect gauge invariance.

A practical lesson is that gauge invariance is not optional bookkeeping. It is a
powerful consistency condition on amplitudes.

\section{Electron--Muon Scattering}

A clean tree-level QED process is
\[
    e^-\mu^-\longrightarrow e^-\mu^-.
\]
At leading order, the particles exchange one virtual photon. If the momentum
transfer is \(q\), the amplitude in Feynman gauge is
\[
    i\mathcal{M}
    = [\bar u_e(p_3)(-ie\gamma^\mu)u_e(p_1)]
      \frac{-i\eta_{\mu\nu}}{q^2+i\epsilon}
      [\bar u_\mu(p_4)(-ie\gamma^\nu)u_\mu(p_2)].
\]
This example shows the basic QED structure:
\[
    \text{current}\times\text{photon propagator}\times\text{current}.
\]
In the nonrelativistic limit, this exchange reproduces the Coulomb interaction.
The photon propagator is therefore the quantum-field-theoretic origin of the
electromagnetic force between charged particles.

\section{Electron--Positron Annihilation}

Another fundamental process is
\[
    e^-e^+\longrightarrow \mu^-\mu^+.
\]
At tree level, the electron and positron annihilate into a virtual photon, which
then creates a muon pair. The amplitude has the form
\[
    i\mathcal{M}
    = [\bar v_e(p_2)(-ie\gamma^\mu)u_e(p_1)]
      \frac{-i\eta_{\mu\nu}}{q^2+i\epsilon}
      [\bar u_\mu(p_3)(-ie\gamma^\nu)v_\mu(p_4)].
\]
Here
\[
    q=p_1+p_2=p_3+p_4.
\]
This process is a prototype for collider physics: an initial particle--antiparticle
pair produces an intermediate virtual gauge boson, which then decays into a new
pair of particles.

\section{Compton Scattering}

Compton scattering is
\[
    e^-\gamma\longrightarrow e^-\gamma.
\]
At tree level, two diagrams contribute. They correspond to the two possible
orderings of photon absorption and emission along the electron line. The amplitude
contains two electron propagators in different channels.

This example is important because gauge invariance requires the sum of both
diagrams. Replacing an external photon polarization by its momentum gives a
cancellation only after both diagrams are included. A single diagram by itself is
not gauge invariant.

Compton scattering therefore illustrates a recurring lesson: physical amplitudes
are often gauge invariant only after summing all diagrams required at a given
order.

\section{Pair Creation and Crossing}

QED processes are related by crossing symmetry. A particle moved from the initial
state to the final state is reinterpreted as an antiparticle with reversed
momentum, and vice versa. For example, electron--positron annihilation and pair
creation by a virtual photon are crossing-related.

Crossing symmetry is not a new Feynman rule. It follows from the analytic
structure of relativistic amplitudes and from the fact that fields contain both
particle annihilation and antiparticle creation operators.

This idea was already visible in the mode expansion of the Dirac field:
\[
    \hat\psi\sim \hat b+\hat d^\dagger.
\]
QED makes this structure operational in scattering calculations.

\section{Loop Corrections in QED}

Beyond tree level, QED has several important loop corrections:
\begin{itemize}
    \item electron self-energy corrections to the fermion propagator;
    \item photon self-energy corrections from fermion loops;
    \item vertex corrections to the electron--photon interaction.
\end{itemize}
These corrections are ultraviolet divergent and require renormalization. They
also produce finite physical effects, such as anomalous magnetic moments and the
running of the electromagnetic coupling.

The next chapter studies loop effects in QED in more detail. For now, the key
point is that QED is not just tree-level photon exchange. Quantum fluctuations of
electrons, positrons, and photons modify the parameters and predictions of the
theory.

\section{Renormalization in QED}

The renormalized QED Lagrangian is written with field and parameter
renormalization constants:
\[
    \mathcal{L}
    = -\frac{1}{4}Z_3 F_{\mu\nu}F^{\mu\nu}
      +Z_2\bar\psi i\gamma^\mu\partial_\mu\psi
      -Z_m m\bar\psi\psi
      -Z_1 e\bar\psi\gamma^\mu\psi A_\mu.
\]
Gauge invariance imposes relations among renormalization constants. In
particular, Ward identities imply a relation between the vertex and fermion
wave-function renormalization.

The physical electric charge is defined by a renormalization condition. At
higher orders, the effective strength of the electromagnetic interaction depends
on the energy scale. This scale dependence is the running of the coupling.

\section{Infrared Issues}

Because the photon is massless, QED also has infrared subtleties. Soft photons
with very small energy can be emitted at little energy cost. As a result,
exclusive amplitudes with a fixed number of photons can contain infrared
divergences.

Physical detectors have finite energy resolution. Observable cross sections must
include processes with unresolved soft photons. When this is done correctly,
infrared divergences cancel between virtual and real soft-photon contributions.

This is one reason the particle interpretation in QED is more subtle than in a
massive scalar theory. Charged particles are accompanied by long-range
electromagnetic fields and soft photon clouds.

\section{What QED Teaches}

QED is the first fully realistic QFT in the course. It teaches several general
lessons:
\begin{itemize}
    \item local gauge symmetry determines the interaction;
    \item photons couple to conserved currents;
    \item amplitudes are built from spinor chains and photon propagators;
    \item gauge invariance appears as Ward identities;
    \item loop corrections require renormalization;
    \item massless gauge bosons create infrared subtleties.
\end{itemize}
These lessons reappear in non-Abelian gauge theory and the Standard Model.

\begin{summarybox}
QED describes charged Dirac matter interacting with the electromagnetic gauge
field. Local \(U(1)\) gauge invariance requires the covariant derivative and
produces the vertex \(-ie\gamma^\mu\). Tree-level amplitudes are built from
spinor currents and photon propagators. Ward identities enforce gauge invariance
and remove unphysical photon polarizations. Loop corrections lead to
renormalization and infrared effects, making QED the first realistic example of
perturbative quantum field theory.
\end{summarybox}

\section{Chapter Checklist}

After reading this chapter, one should be able to:
\begin{itemize}
    \item derive QED from local \(U(1)\) gauge invariance;
    \item write the QED Lagrangian and identify the interaction term;
    \item state the photon and electron propagators;
    \item state the QED vertex rule;
    \item construct spinor chains for simple amplitudes;
    \item explain the meaning of Ward identities;
    \item write the tree-level structure of electron--muon scattering;
    \item write the tree-level structure of electron--positron annihilation;
    \item explain why Compton scattering requires a sum of diagrams;
    \item describe the main loop corrections and renormalization issues in QED;
    \item explain why massless photons produce infrared subtleties.
\end{itemize}
