\chapter{Fermionic Path Integrals}

\begin{remarkbox}
Scalar path integrals integrate over ordinary commuting fields. Fermionic path
integrals require anticommuting variables. The price is algebraic unfamiliarity;
the reward is a compact functional-integral description of Dirac fields,
fermion propagators, fermion loops, and determinants.
\end{remarkbox}

\section{Why Fermions Need a Different Integral}

The scalar path integral sums over ordinary field configurations:
\[
    Z[J]=\int\mathcal{D}\phi\,e^{iS[\phi]+i\int J\phi}.
\]
This works because scalar fields are bosonic. Their quantum operators obey
commutation relations, and the corresponding functional variables commute.

Dirac fields are different. Their operators obey anticommutation relations, and
fermionic many-particle states are antisymmetric. A path integral for fermions
must reproduce this algebra. The appropriate variables are therefore not
ordinary complex numbers, but Grassmann variables: objects that anticommute.

The transition is
\[
    \text{bosons} \longrightarrow \text{commuting variables},
    \qquad
    \text{fermions} \longrightarrow \text{anticommuting variables}.
\]
This chapter develops the basic calculus needed for fermionic functional
integrals.

\section{Grassmann Variables}

A Grassmann variable \(\theta\) anticommutes with another Grassmann variable
\(\chi\):
\[
    \theta\chi=-\chi\theta.
\]
In particular,
\[
    \theta^2=0.
\]
This nilpotency makes every function of a single Grassmann variable finite:
\[
    f(\theta)=a+b\theta.
\]
For several Grassmann variables, polynomials terminate because each variable can
appear at most once.

Grassmann variables also anticommute with fermionic sources and fermionic fields,
but commute with ordinary numbers and bosonic variables. They are bookkeeping
objects that encode fermionic antisymmetry inside a functional integral.

\section{Berezin Integration}

Integration over Grassmann variables is called Berezin integration. It is defined
by the rules
\[
    \int\dd\theta\,1=0,
    \qquad
    \int\dd\theta\,\theta=1.
\]
Thus Grassmann integration behaves like differentiation. For a function
\(f(\theta)=a+b\theta\),
\[
    \int\dd\theta\,f(\theta)=b.
\]
For two variables, the order matters:
\[
    \int\dd\theta_2\dd\theta_1\,\theta_1\theta_2=1.
\]
Changing the order of variables or integration measures can introduce minus
signs. This is the algebraic origin of many sign factors in fermionic path
integrals.

\section{Gaussian Integrals for Grassmann Variables}

For one pair of independent Grassmann variables \(\bar\theta,\theta\),
\[
    \int\dd\bar\theta\dd\theta\,
    e^{-\bar\theta a\theta}=a.
\]
For many variables, the general formula is
\[
    \int\prod_i\dd\bar\theta_i\dd\theta_i\,
    \exp\left(-\bar\theta_i A_{ij}\theta_j\right)
    = \det A.
\]
This differs from the bosonic Gaussian integral, which produces
\((\det A)^{-1/2}\) for real variables or \((\det A)^{-1}\) for complex
commuting variables. Fermions produce determinants in the numerator.

With sources, one has
\[
    \int\dd\bar\theta\dd\theta\,
    \exp\left(-\bar\theta A\theta+\bar\eta\theta+\bar\theta\eta\right)
    = \det A\,\exp\left(\bar\eta A^{-1}\eta\right),
\]
up to conventions for signs and ordering. The inverse matrix \(A^{-1}\) will
become the fermion propagator.

\section{The Dirac Generating Functional}

The free Dirac action is
\[
    S_0[\bar\psi,\psi]
    = \int\dd^4x\,\bar\psi(i\gamma^\mu\partial_\mu-m)\psi.
\]
In the path integral, \(\psi\) and \(\bar\psi\) are independent Grassmann-valued
fields. Introduce Grassmann-valued sources \(\eta\) and \(\bar\eta\):
\[
    Z_0[\bar\eta,\eta]
    = \int\mathcal{D}\bar\psi\mathcal{D}\psi\,
      \exp\left(iS_0[\bar\psi,\psi]
      +i\int\dd^4x\,[\bar\eta\psi+\bar\psi\eta]\right).
\]
This is a Gaussian Grassmann functional integral. It gives
\[
    Z_0[\bar\eta,\eta]
    = Z_0[0,0]\exp\left(
        -i\int\dd^4x\dd^4y\,
        \bar\eta(x)S_F(x-y)\eta(y)
      \right),
\]
where \(S_F\) is the Dirac Feynman propagator.

The sign in the exponent depends on source-ordering conventions. The essential
point is invariant: differentiating with respect to fermionic sources produces
Dirac propagators and fermionic correlation functions.

\section{Dirac Propagator from the Path Integral}

The Dirac propagator is the inverse of the Dirac operator:
\[
    (i\gamma^\mu\partial_\mu-m)S_F(x-y)=i\delta^{(4)}(x-y).
\]
In momentum space,
\[
    S_F(p)=\frac{i(\gamma^\mu p_\mu+m)}{p^2-m^2+i\epsilon}.
\]
From the generating functional,
\[
    \langle0|T\hat\psi_\alpha(x)\hat{\bar\psi}_\beta(y)|0\rangle
    = S_{F,\alpha\beta}(x-y).
\]
Thus the same propagator obtained by canonical quantization appears as the
inverse operator in the fermionic Gaussian integral.

The path-integral perspective makes the parallel with scalar fields clear:
\[
    \text{quadratic operator in the action}^{-1}
    = \text{free propagator}.
\]
For scalars the operator is Klein--Gordon. For fermions it is Dirac.

\section{Generating Fermionic Correlation Functions}

Fermionic correlation functions are generated by differentiating with respect to
Grassmann sources. Schematically,
\[
    \langle T\psi(x_1)\bar\psi(y_1)\cdots\psi(x_n)\bar\psi(y_n)\rangle
\]
is obtained by applying derivatives with respect to \(\bar\eta(x_i)\) and
\(\eta(y_i)\) to \(Z[\bar\eta,\eta]\), then setting the sources to zero.

The order of the derivatives matters. Since the sources anticommute, exchanging
two source derivatives can produce a minus sign. These signs are the
path-integral version of the fermionic signs in Wick's theorem.

A useful practical rule is that the path integral automatically keeps track of
fermionic antisymmetry if the source ordering is used consistently.

\section{Interactions with Fermions}

For QED, the interaction term is
\[
    S_{\mathrm{int}}
    = -e\int\dd^4x\,\bar\psi\gamma^\mu\psi A_\mu.
\]
The full generating functional is
\[
    Z[\bar\eta,\eta,J]
    = \int\mathcal{D}A\mathcal{D}\bar\psi\mathcal{D}\psi\,
      \exp\left(iS[A,\bar\psi,\psi]
      +i\int\dd^4x\,[J_\mu A^\mu+\bar\eta\psi+\bar\psi\eta]\right).
\]
Perturbation theory is obtained by expanding the interaction exponential or by
rewriting interactions as functional derivatives with respect to sources.

The QED vertex contains one photon field, one \(\psi\), and one \(\bar\psi\).
Therefore each vertex joins one photon line and two fermion lines. This is the
path-integral origin of the usual QED diagrammatic vertex.

\section{Fermion Loops and Minus Signs}

Closed fermion loops carry an extra minus sign. In the operator formalism this
comes from permuting fermionic operators in Wick contractions. In the path
integral it comes from the algebra of Grassmann variables and the determinant
produced by integrating out fermions.

A closed fermion loop is associated with a trace over spinor indices and an
integral over loop momentum. The additional sign is part of the Feynman rule:
\[
    \text{each closed fermion loop contributes a factor } -1.
\]
This rule is not an arbitrary convention. It is a direct consequence of
fermionic anticommutation.

\section{Determinants from Integrating Out Fermions}

Because the fermionic path integral is Gaussian when the action is quadratic in
\(\psi,\bar\psi\), one can formally integrate out fermions. For a Dirac operator
\(D\),
\[
    \int\mathcal{D}\bar\psi\mathcal{D}\psi\,
    e^{i\int\bar\psi D\psi}
    \propto \det D.
\]
If the Dirac operator depends on a background gauge field, the determinant also
depends on that background:
\[
    \det(i\gamma^\mu D_\mu-m).
\]
This determinant encodes the effect of virtual fermion loops on the gauge-field
dynamics.

For bosonic Gaussian integrals, one typically gets inverse determinants. This
difference between bosonic and fermionic determinants is one of the key
structural distinctions between commuting and anticommuting fields.

\section{Majorana Fermions in the Path Integral}

A Majorana fermion is constrained to be related to its charge conjugate. In the
path integral this means that the independent fermionic variables are reduced
relative to a Dirac field. The Gaussian integral no longer gives an ordinary
determinant in the same way; it gives a Pfaffian.

Schematically, for an antisymmetric operator \(A\),
\[
    \int\mathcal{D}\chi\,e^{-\frac{1}{2}\chi A\chi}
    \propto \operatorname{Pf}(A),
\]
and
\[
    \operatorname{Pf}(A)^2=\det A.
\]
The details require a careful treatment of charge conjugation and spinor
conventions. At this stage, the main lesson is that Majorana path integrals are
fermionic Gaussian integrals with fewer independent variables than Dirac path
integrals.

\section{Physical Interpretation and Common Pitfalls}

Grassmann variables are not classical values of a fermion field. They are
anticommuting integration variables designed to reproduce fermionic quantum
statistics. One should not imagine a fermionic path integral as a sum over
ordinary classical spinor configurations.

Another common pitfall is to treat \(\psi\) and \(\bar\psi\) as complex
conjugates inside the functional integral. In the fermionic path integral they
are independent Grassmann variables. Their relation as adjoint operators belongs
to the canonical operator formulation, while the functional integral uses an
independent-variable representation.

Finally, fermionic determinants and loop signs are not optional decorations.
They are the mechanism by which the Pauli principle and virtual fermion effects
enter the path-integral formalism.

\begin{summarybox}
Fermionic path integrals use Grassmann variables. Berezin integration turns
fermionic Gaussian integrals into determinants, and source derivatives generate
fermionic time-ordered correlation functions. The inverse Dirac operator is the
Dirac propagator. Fermion loops acquire minus signs, and integrating out
fermions produces determinants that encode virtual fermion effects. Majorana
fermions require a related Pfaffian structure.
\end{summarybox}

\section{Chapter Checklist}

After reading this chapter, one should be able to:
\begin{itemize}
    \item define Grassmann variables and state their anticommutation property;
    \item state the basic rules of Berezin integration;
    \item evaluate the structure of a Gaussian Grassmann integral;
    \item write the free Dirac generating functional with Grassmann sources;
    \item identify the Dirac propagator as the inverse Dirac operator;
    \item explain how fermionic correlation functions are generated by source
    derivatives;
    \item explain why closed fermion loops carry a minus sign;
    \item describe how integrating out fermions produces determinants;
    \item explain the basic path-integral difference between Dirac and Majorana
    fermions;
    \item avoid interpreting Grassmann fields as ordinary classical spinor
    configurations.
\end{itemize}
