\chapter{From Classical Normal Modes to Quantum Oscillators}

\begin{remarkbox}
The first bridge from classical field theory to quantum field theory is the
normal-mode decomposition of a free field. A free field is not quantized as one
large indivisible object. It is first decomposed into independent oscillating
modes, and each mode is then quantized as a harmonic oscillator.
\end{remarkbox}

\section{The Aim of the Chapter}

Classical field theory describes a field by a function such as
\(\phi(t,\vect{x})\). Quantum field theory replaces this classical function by
an operator acting on a Hilbert space. That replacement can look abrupt if one
tries to perform it directly. The purpose of this chapter is to make the first
step gradual.

The key observation is that the free-field problem is already familiar from
classical mechanics. A free field can be decomposed into normal modes. Each
normal mode behaves like a harmonic oscillator. Therefore the first version of
field quantization is:
\[
    \text{free field}
    \longrightarrow
    \text{many independent classical oscillators}
    \longrightarrow
    \text{many independent quantum oscillators}.
\]
The excitations of these quantum oscillators will be interpreted as particles.
This is the origin of the particle language in perturbative quantum field
theory.

\section{A Finite Chain as the Warm-Up}

Begin with a finite chain of \(N\) identical masses connected by springs. Let
\(q_j(t)\) be the displacement of the \(j\)-th mass from equilibrium. A typical
quadratic Lagrangian is
\[
    L = \sum_{j=1}^{N}\frac{m}{2}\dot q_j^2
        - \sum_{j=1}^{N}\frac{\kappa}{2}(q_{j+1}-q_j)^2,
\]
with boundary conditions chosen so that the notation is meaningful. The
coordinates \(q_j(t)\) are coupled because the spring energy contains nearest
neighbour differences.

The usual normal-mode method introduces new variables \(Q_r(t)\) by expanding
\[
    q_j(t) = \sum_r Q_r(t) u_r(j),
\]
where \(u_r(j)\) are spatial mode functions adapted to the boundary conditions.
For a suitable choice of normalization, the Lagrangian becomes a sum of
independent oscillators,
\[
    L = \sum_r
    \left(
        \frac{1}{2}\dot Q_r^{\,2}
        - \frac{1}{2}\omega_r^2 Q_r^2
    \right).
\]
The coupled coordinates \(q_j\) were not the best variables. The normal
coordinates \(Q_r\) reveal that the system is a collection of independent
oscillations.

\begin{examplebox}
A finite chain is not yet a field theory, because it has finitely many degrees
of freedom. Nevertheless, it shows the essential mechanism. A spatially extended
system may look complicated in position variables, but simple in normal-mode
variables.
\end{examplebox}

\section{From a Chain to a Field}

A field is the continuum analogue of a system with one degree of freedom at
each spatial point. In a one-dimensional chain with spacing \(a\), the position
of the \(j\)-th mass is \(x_j=ja\), and one may view
\[
    q_j(t) \approx \phi(t,x_j).
\]
As \(a\) becomes small and the number of points becomes large, sums over sites
are replaced by integrals over space, and discrete differences are replaced by
spatial derivatives.

For a real scalar field in three spatial dimensions, the free classical
Lagrangian is
\[
    L = \int \dd^3x\,
    \left(
        \frac{1}{2}\dot\phi^2
        - \frac{1}{2}(\nabla\phi)^2
        - \frac{1}{2}m^2\phi^2
    \right).
\]
The corresponding field equation is the Klein--Gordon equation,
\[
    (\partial_t^2-\nabla^2+m^2)\phi(t,\vect{x})=0.
\]
This is the classical starting point for the simplest relativistic quantum
field.

\section{Normal Modes of the Free Scalar Field}

To avoid technical distractions from continuous spectra, place the field in a
box of volume \(V\) and impose periodic boundary conditions. The allowed momenta
are then discrete. The field can be expanded in spatial Fourier modes,
\[
    \phi(t,\vect{x})
    = \frac{1}{\sqrt{V}}\sum_{\vect{k}} q_{\vect{k}}(t)
        e^{i\vect{k}\cdot\vect{x}},
\]
with the appropriate reality condition for a real field. One may equivalently
use real sine and cosine modes, in which all mode amplitudes are real. The
choice of basis is a matter of convenience; the physical point is that the free
field decomposes into independent modes.

For each momentum \(\vect{k}\), the frequency is
\[
    \omega_{\vect{k}}=\sqrt{\vect{k}^{\,2}+m^2}
\]
in units with \(c=\hbar=1\). After diagonalizing the quadratic Lagrangian, one
obtains the schematic form
\[
    L = \sum_{\vect{k}}
    \left(
        \frac{1}{2}\dot q_{\vect{k}}^{\,2}
        - \frac{1}{2}\omega_{\vect{k}}^2 q_{\vect{k}}^{\,2}
    \right).
\]
Thus a free scalar field is a collection of harmonic oscillators, one for each
allowed momentum mode.

\begin{remarkbox}
The phrase ``schematic form'' hides normalization conventions and the reality
condition of a real field. These details are important in precise calculations,
but they do not change the main structural fact: the free field is diagonalized
by normal modes, and each mode behaves as an oscillator.
\end{remarkbox}

\section{The Hamiltonian of the Modes}

For one mode, the canonical momentum is
\[
    p_{\vect{k}} = \dot q_{\vect{k}}.
\]
The Hamiltonian of the free field becomes a sum of mode Hamiltonians,
\[
    H = \sum_{\vect{k}} H_{\vect{k}},
    \qquad
    H_{\vect{k}}
    = \frac{1}{2}p_{\vect{k}}^{\,2}
      + \frac{1}{2}\omega_{\vect{k}}^2 q_{\vect{k}}^{\,2}.
\]
This is the decisive simplification. The quantization of the free field reduces
to the quantization of the oscillator Hamiltonian \(H_{\vect{k}}\) for every
\(\vect{k}\).

In an interacting theory the Hamiltonian also contains terms such as
\(\phi^4\), \(\bar\psi\psi\phi\), or gauge interactions. Those terms mix the
free modes. Perturbative QFT begins by quantizing the free modes exactly and
then treating interactions as controlled corrections.

\section{Quantizing One Harmonic Oscillator}

Recall the quantum harmonic oscillator. The classical variables \(q\) and \(p\)
are replaced by operators \(\hat q\) and \(\hat p\) satisfying
\[
    [\hat q,\hat p]=i.
\]
The Hamiltonian is
\[
    \hat H = \frac{1}{2}\hat p^2+rac{1}{2}\omega^2\hat q^2.
\]
Introduce annihilation and creation operators by
\[
    \hat a
    = \sqrt{\frac{\omega}{2}}\,\hat q
      + \frac{i}{\sqrt{2\omega}}\,\hat p,
    \qquad
    \hat a^\dagger
    = \sqrt{\frac{\omega}{2}}\,\hat q
      - \frac{i}{\sqrt{2\omega}}\,\hat p.
\]
These definitions can be inverted:
\[
    \hat q = \frac{1}{\sqrt{2\omega}}(\hat a+\hat a^\dagger),
    \qquad
    \hat p = -i\sqrt{\frac{\omega}{2}}(\hat a-\hat a^\dagger).
\]
The commutation relation becomes
\[
    [\hat a,\hat a^\dagger]=1.
\]
The Hamiltonian is then
\[
    \hat H = \omega\left(\hat a^\dagger\hat a+\frac{1}{2}\right).
\]

\section{Number States}

The number operator is
\[
    \hat N=\hat a^\dagger\hat a.
\]
Its eigenvalues are non-negative integers. The vacuum of the oscillator is the
state \(|0\rangle\) satisfying
\[
    \hat a|0\rangle=0.
\]
The normalized excited states are
\[
    |n\rangle=\frac{(\hat a^\dagger)^n}{\sqrt{n!}}|0\rangle,
    \qquad n=0,1,2,\ldots.
\]
They satisfy
\[
    \hat N|n\rangle=n|n\rangle,
    \qquad
    \hat H|n\rangle=\omega\left(n+\frac{1}{2}\right)|n\rangle.
\]

The operators \(\hat a^\dagger\) and \(\hat a\) raise and lower the excitation
number. This is why they are called creation and annihilation operators. For a
single oscillator this language is only a convenient algebraic description. In
a field theory it becomes the language of particles.

\section{Quantizing All Modes}

For the field, each normal mode receives its own pair of operators,
\[
    \hat a_{\vect{k}},
    \qquad
    \hat a_{\vect{k}}^\dagger.
\]
The oscillator algebra becomes
\[
    [\hat a_{\vect{k}},\hat a_{\vect{k}'}^\dagger]
    = \delta_{\vect{k},\vect{k}'},
    \qquad
    [\hat a_{\vect{k}},\hat a_{\vect{k}'}]=0,
    \qquad
    [\hat a_{\vect{k}}^\dagger,\hat a_{\vect{k}'}^\dagger]=0.
\]
The free-field vacuum is the state annihilated by every annihilation operator,
\[
    \hat a_{\vect{k}}|0\rangle=0
    \qquad \text{for every } \vect{k}.
\]
A one-particle state with momentum \(\vect{k}\) is
\[
    |\vect{k}\rangle=\hat a_{\vect{k}}^\dagger|0\rangle.
\]
A state with two particles of momenta \(\vect{k}_1\) and \(\vect{k}_2\) is
\[
    |\vect{k}_1,\vect{k}_2\rangle
    = \hat a_{\vect{k}_1}^\dagger
      \hat a_{\vect{k}_2}^\dagger|0\rangle,
\]
up to normalization conventions.

\begin{definitionbox}
The \emph{Fock space} of a free quantum field is the Hilbert space generated
from the vacuum by applying creation operators for all allowed modes. It is the
natural state space for a theory in which the particle number can vary.
\end{definitionbox}

\section{The Free-Field Hamiltonian}

After quantization, the free scalar Hamiltonian takes the form
\[
    \hat H
    = \sum_{\vect{k}}\omega_{\vect{k}}
    \left(
        \hat a_{\vect{k}}^\dagger\hat a_{\vect{k}}
        + \frac{1}{2}
    \right).
\]
The first term counts excitations of each mode. The second term is the
zero-point energy of each oscillator. Since a field has infinitely many modes,
the formal vacuum energy
\[
    E_0=\frac{1}{2}\sum_{\vect{k}}\omega_{\vect{k}}
\]
is divergent in the continuum limit.

This divergence is not an accident of notation. It is the first warning that
field theory has infinitely many degrees of freedom and requires new tools.
Normal ordering, regularization, renormalization, and the physical meaning of
vacuum energy will appear later. For now, the important fact is that energy
above the free vacuum is counted by number operators.

\section{The Field Operator}

The quantum field itself is reconstructed from the mode operators. In the
continuum normalization commonly used in relativistic QFT, the free real scalar
field has the form
\[
    \hat\phi(t,\vect{x})
    = \int \frac{\dd^3k}{(2\pi)^3}\,
      \frac{1}{\sqrt{2\omega_{\vect{k}}}}
      \left(
          \hat a_{\vect{k}} e^{-i\omega_{\vect{k}}t+i\vect{k}\cdot\vect{x}}
          + \hat a_{\vect{k}}^\dagger
            e^{i\omega_{\vect{k}}t-i\vect{k}\cdot\vect{x}}
      \right).
\]
This formula should be read as the quantum version of the classical mode
expansion. The coefficients are now operators, not ordinary numbers.

The field operator satisfies the free Klein--Gordon equation,
\[
    (\Box+m^2)\hat\phi(x)=0.
\]
This does not mean that the field has a definite classical value. It means that
the operator-valued distribution evolves according to the free equation of
motion.

\begin{remarkbox}
A quantum field is not simply a wavefunction. A wavefunction describes a state.
A field operator acts on states. This distinction is essential: particles are
states in Fock space, while fields are operators that create, annihilate, and
probe those states.
\end{remarkbox}

\section{Particles as Mode Excitations}

In the free theory, the particle interpretation is sharp. The state
\(\hat a_{\vect{k}}^\dagger|0\rangle\) has momentum \(\vect{k}\) and energy
\(\omega_{\vect{k}}\). It is therefore interpreted as a one-particle state.
Repeated creation operators produce many-particle states.

This interpretation is powerful but limited. In interacting QFT, particles can
be dressed by interactions, unstable particles can appear as resonances rather
than exact states, and the vacuum may differ from the free vacuum. Nevertheless,
the free-particle picture is the starting approximation behind perturbative
scattering theory.

\section{What Has Changed from Classical Field Theory?}

The classical field \(\phi(x)\) assigned a number to each spacetime point. The
quantum field \(\hat\phi(x)\) assigns an operator to each spacetime point. The
classical mode amplitude \(q_{\vect{k}}\) became an oscillator operator, and the
classical normal mode became a possible particle excitation.

The conceptual dictionary is therefore:
\[
\begin{array}{ccl}
\text{classical mode amplitude} & \longrightarrow & \text{oscillator operator},\\[0.25em]
\text{classical normal mode} & \longrightarrow & \text{quantum mode},\\[0.25em]
\text{oscillator excitation} & \longrightarrow & \text{particle state},\\[0.25em]
\text{field configuration} & \longrightarrow & \text{operator acting on Fock space}.
\end{array}
\]

\section{Common Pitfalls}

\begin{itemize}
    \item A field quantum is not a small localized piece of classical field. It
    is an excitation of a mode of the quantum field.

    \item The vacuum is not ``nothing''. It is the state annihilated by all free
    annihilation operators, and it still carries formal zero-point energy.

    \item The field operator is not the same object as a particle state. The
    field operator acts on the vacuum to create a superposition of one-particle
    states.

    \item The free-particle picture is exact only for the free theory. In an
    interacting theory it becomes an asymptotic and perturbative organizing
    principle.
\end{itemize}

\section{Chapter Checklist}

After reading this chapter, one should be able to explain the following points:
\begin{itemize}
    \item why normal modes diagonalize a free classical field;
    \item why each normal mode behaves like a harmonic oscillator;
    \item how a harmonic oscillator is quantized using \(\hat a\) and
    \(\hat a^\dagger\);
    \item how the free-field vacuum is defined;
    \item how one-particle and many-particle states are built;
    \item why Fock space is the natural Hilbert space for free fields;
    \item why particles in free QFT are mode excitations rather than tiny
    classical objects.
\end{itemize}

\begin{summarybox}
The free scalar field is the prototype of quantum field theory. Its classical
normal modes are harmonic oscillators. Quantizing those oscillators produces
creation and annihilation operators. Acting with creation operators on the
vacuum builds Fock space, and the resulting mode excitations are interpreted as
particles. This construction is the foundation on which propagators,
perturbation theory, and scattering amplitudes will be built.
\end{summarybox}
