\chapter{The Complex Scalar Field and Antiparticles}

\begin{remarkbox}
The real scalar field has one neutral quantum for each momentum mode. The
complex scalar field is the next step: the field is no longer Hermitian, and its
quantization produces two independent kinds of quanta. They are naturally
interpreted as particles and antiparticles.
\end{remarkbox}

\section{From a Real Field to a Complex Field}

Chapter 2 quantized a real scalar field. The reality condition
\(\hat\phi^\dagger=\hat\phi\) forced the same field operator to contain both
annihilation and creation operators. A complex scalar field is different. The
field \(\phi\) and its Hermitian conjugate \(\phi^\dagger\) are independent
classical variables before quantization.

One may think of a complex scalar field as two real scalar fields packaged into
one object:
\[
    \phi
    = \frac{1}{\sqrt{2}}(\phi_1+i\phi_2),
    \qquad
    \phi^\dagger
    = \frac{1}{\sqrt{2}}(\phi_1-i\phi_2).
\]
The important new feature is that the theory has a global phase symmetry. The
field may be rotated by a constant phase,
\[
    \phi\longrightarrow e^{-i\alpha}\phi,
    \qquad
    \phi^\dagger\longrightarrow e^{i\alpha}\phi^\dagger.
\]
By Noether's theorem this symmetry gives a conserved charge. In the quantum
theory, that charge will distinguish particles from antiparticles.

\section{Classical Lagrangian and Canonical Momenta}

The free complex scalar field is described by
\[
    \mathcal{L}
    = \partial_\mu\phi^\dagger\partial^\mu\phi
      - m^2\phi^\dagger\phi.
\]
In terms of time and space derivatives,
\[
    \mathcal{L}
    = \dot\phi^\dagger\dot\phi
      - \nabla\phi^\dagger\cdot\nabla\phi
      - m^2\phi^\dagger\phi.
\]
The conjugate momenta are
\[
    \pi
    = \frac{\partial\mathcal{L}}{\partial\dot\phi}
    = \dot\phi^\dagger,
    \qquad
    \pi^\dagger
    = \frac{\partial\mathcal{L}}{\partial\dot\phi^\dagger}
    = \dot\phi.
\]
Thus \(\pi\) is conjugate to \(\phi\), while \(\pi^\dagger\) is conjugate to
\(\phi^\dagger\).

The equations of motion are
\[
    (\Box+m^2)\phi=0,
    \qquad
    (\Box+m^2)\phi^\dagger=0.
\]
Both components satisfy the Klein--Gordon equation, but the quantum
interpretation will be richer than in the real case.

\section{Canonical Commutation Relations}

Canonical quantization promotes
\(\phi,\phi^\dagger,\pi,\pi^\dagger\) to operators. The nonzero equal-time
commutators are
\[
    [\hat\phi(t,\vect{x}),\hat\pi(t,\vect{y})]
    = i\delta^{(3)}(\vect{x}-\vect{y}),
\]
and
\[
    [\hat\phi^\dagger(t,\vect{x}),\hat\pi^\dagger(t,\vect{y})]
    = i\delta^{(3)}(\vect{x}-\vect{y}).
\]
All other equal-time commutators vanish. In particular,
\[
    [\hat\phi(t,\vect{x}),\hat\pi^\dagger(t,\vect{y})]=0,
    \qquad
    [\hat\phi^\dagger(t,\vect{x}),\hat\pi(t,\vect{y})]=0.
\]
This is the canonical field algebra for a charged bosonic scalar field.

\section{Mode Expansion}

The complex field is not Hermitian, so its mode expansion cannot be the same as
the real scalar expansion. The standard continuum expansion is
\[
    \hat\phi(t,\vect{x})
    = \int\frac{\dd^3k}{(2\pi)^3}
      \frac{1}{\sqrt{2\omega_{\vect{k}}}}
      \left(
          \hat a_{\vect{k}}
          e^{-i\omega_{\vect{k}}t+i\vect{k}\cdot\vect{x}}
          +
          \hat b_{\vect{k}}^\dagger
          e^{i\omega_{\vect{k}}t-i\vect{k}\cdot\vect{x}}
      \right),
\]
where
\[
    \omega_{\vect{k}}=\sqrt{\vect{k}^{\,2}+m^2}.
\]
The Hermitian conjugate field is
\[
    \hat\phi^\dagger(t,\vect{x})
    = \int\frac{\dd^3k}{(2\pi)^3}
      \frac{1}{\sqrt{2\omega_{\vect{k}}}}
      \left(
          \hat a_{\vect{k}}^\dagger
          e^{i\omega_{\vect{k}}t-i\vect{k}\cdot\vect{x}}
          +
          \hat b_{\vect{k}}
          e^{-i\omega_{\vect{k}}t+i\vect{k}\cdot\vect{x}}
      \right).
\]
There are now two independent sets of ladder operators: the \(a\)-operators and
the \(b\)-operators. This is the algebraic origin of antiparticles in the
complex scalar theory.

\section{Two Independent Oscillator Families}

The nonzero ladder-operator commutators are
\[
    [\hat a_{\vect{k}},\hat a_{\vect{p}}^\dagger]
    = (2\pi)^3\delta^{(3)}(\vect{k}-\vect{p}),
\]
and
\[
    [\hat b_{\vect{k}},\hat b_{\vect{p}}^\dagger]
    = (2\pi)^3\delta^{(3)}(\vect{k}-\vect{p}).
\]
The two families commute with each other:
\[
    [\hat a_{\vect{k}},\hat b_{\vect{p}}]=0,
    \qquad
    [\hat a_{\vect{k}},\hat b_{\vect{p}}^\dagger]=0,
\]
with similar relations for their Hermitian conjugates.

The vacuum is defined by
\[
    \hat a_{\vect{k}}|0\rangle=0,
    \qquad
    \hat b_{\vect{k}}|0\rangle=0
    \qquad \text{for every }\vect{k}.
\]
There are two basic one-particle states:
\[
    |\vect{k};a\rangle
    = \hat a_{\vect{k}}^\dagger|0\rangle,
    \qquad
    |\vect{k};b\rangle
    = \hat b_{\vect{k}}^\dagger|0\rangle.
\]
They have the same mass and the same energy-momentum relation, but opposite
charges.

\section{Hamiltonian and Momentum}

After normal ordering, the Hamiltonian of the free complex scalar field is
\[
    :\hat H:
    = \int\frac{\dd^3k}{(2\pi)^3}\,
      \omega_{\vect{k}}
      \left(
          \hat a_{\vect{k}}^\dagger\hat a_{\vect{k}}
          +
          \hat b_{\vect{k}}^\dagger\hat b_{\vect{k}}
      \right).
\]
The momentum operator is
\[
    :\hat{\vect{P}}:
    = \int\frac{\dd^3k}{(2\pi)^3}\,
      \vect{k}
      \left(
          \hat a_{\vect{k}}^\dagger\hat a_{\vect{k}}
          +
          \hat b_{\vect{k}}^\dagger\hat b_{\vect{k}}
      \right).
\]
Both kinds of quanta carry positive energy. The difference between them is not
energy or mass, but charge.

\section{The Charge Operator}

The global phase symmetry gives a conserved current. With the convention
\(\phi\to e^{-i\alpha}\phi\), the corresponding quantum charge may be written,
after normal ordering, as
\[
    \hat Q
    = \int\frac{\dd^3k}{(2\pi)^3}
      \left(
          \hat a_{\vect{k}}^\dagger\hat a_{\vect{k}}
          -
          \hat b_{\vect{k}}^\dagger\hat b_{\vect{k}}
      \right).
\]
This convention gives
\[
    \hat Q\,\hat a_{\vect{k}}^\dagger|0\rangle
    = +\hat a_{\vect{k}}^\dagger|0\rangle,
    \qquad
    \hat Q\,\hat b_{\vect{k}}^\dagger|0\rangle
    = -\hat b_{\vect{k}}^\dagger|0\rangle.
\]
Thus the \(a\)-quanta have charge \(+1\), while the \(b\)-quanta have charge
\(-1\). They are particle and antiparticle excitations of the same complex
field.

The charge operator also generates the phase transformation of the field:
\[
    [\hat Q,\hat\phi(x)] = -\hat\phi(x),
    \qquad
    [\hat Q,\hat\phi^\dagger(x)] = +\hat\phi^\dagger(x).
\]
This is the operator statement that \(\hat\phi\) carries charge \(-1\) as an
operator, even though \(\hat a^\dagger\) creates a charge \(+1\) state in the
chosen convention.

\section{Positive and Negative Frequency Revisited}

In the real scalar field, the negative-frequency part was simply the Hermitian
conjugate of the positive-frequency part of the same field. In the complex
field, the negative-frequency part of \(\hat\phi\) contains
\(\hat b^\dagger\), not \(\hat a^\dagger\). This is the crucial difference.

The field \(\hat\phi\) can annihilate an \(a\)-particle or create a
\(b\)-antiparticle. The conjugate field \(\hat\phi^\dagger\) can create an
\(a\)-particle or annihilate a \(b\)-antiparticle. Symbolically,
\[
    \hat\phi \sim \hat a + \hat b^\dagger,
    \qquad
    \hat\phi^\dagger \sim \hat a^\dagger + \hat b.
\]
This is one of the cleanest ways to see why relativistic quantum fields contain
antiparticles.

\section{Why Antiparticles Appear Naturally}

The appearance of antiparticles is not an optional addition. It follows from
combining three requirements:
\begin{itemize}
    \item relativistic positive-energy modes;
    \item canonical quantization;
    \item a complex field with a conserved charge.
\end{itemize}
The field must contain both positive- and negative-frequency solutions in order
to be local and to satisfy the canonical commutation relations. For a charged
field, those two frequency sectors are associated with two independent oscillator
families. The quanta created by the second family have the opposite charge and
are interpreted as antiparticles.

\section{Complex Scalar Propagator}

The natural two-point function of the complex scalar field is
\[
    D_F(x-y)
    = \langle0|T\hat\phi(x)\hat\phi^\dagger(y)|0\rangle.
\]
It has the same momentum-space denominator as the real scalar propagator:
\[
    D_F(x-y)
    = \int\frac{\dd^4k}{(2\pi)^4}\,
      \frac{i}{k^2-m^2+i\epsilon}
      e^{-ik\cdot(x-y)}.
\]
The difference is in the charge flow. The propagator connects \(\phi\) with
\(\phi^\dagger\), because charge must be conserved. Correlators such as
\(\langle0|T\hat\phi(x)\hat\phi(y)|0\rangle\) vanish in the free theory when the
vacuum has zero charge.

\section{Physical Interpretation of the Fock Space}

The Fock space of the complex scalar field contains particles, antiparticles,
and arbitrary multiparticle states. Examples are
\[
    \hat a_{\vect{k}}^\dagger|0\rangle,
    \qquad
    \hat b_{\vect{p}}^\dagger|0\rangle,
    \qquad
    \hat a_{\vect{k}}^\dagger\hat b_{\vect{p}}^\dagger|0\rangle.
\]
The last state contains one particle and one antiparticle. Its total charge is
zero, but it is not the vacuum. Charge conservation restricts which transitions
are possible once interactions are introduced.

\begin{summarybox}
A complex scalar field is equivalent to two real scalar degrees of freedom, but
its global phase symmetry gives a conserved charge. Quantization produces two
independent oscillator families. One creates particles and the other creates
antiparticles, with equal mass and opposite charge. The field \(\hat\phi\)
contains an annihilation operator for particles and a creation operator for
antiparticles, while \(\hat\phi^\dagger\) does the reverse.
\end{summarybox}

\section{Chapter Checklist}

After reading this chapter, one should be able to:
\begin{itemize}
    \item explain why \(\phi\) and \(\phi^\dagger\) are independent field
    variables before quantization;
    \item write the canonical momenta of the complex scalar field;
    \item state the equal-time commutation relations;
    \item write the mode expansions of \(\hat\phi\) and \(\hat\phi^\dagger\);
    \item explain why two independent ladder-operator families appear;
    \item construct particle and antiparticle states;
    \item write the charge operator and interpret its eigenvalues;
    \item explain why antiparticles are required in a relativistic charged field;
    \item identify the charge structure of the complex scalar propagator.
\end{itemize}
