\chapter{\texorpdfstring{\(\phi^4\) Theory as the First Interacting QFT}{phi4 Theory as the First Interacting QFT}}

\begin{remarkbox}
The previous chapters built the general machinery: canonical quantization,
propagators, perturbation theory, scattering, and path integrals. The theory
with interaction \(\lambda\phi^4/4!\) is the simplest laboratory where all of
these ideas become concrete.
\end{remarkbox}

\section{Why Start with \texorpdfstring{\(\phi^4\)}{phi4} Theory?}

The free scalar field is exactly solvable because it is a collection of
independent oscillators. To study genuine quantum field theory, we need an
interaction. The simplest Lorentz-invariant self-interaction of a real scalar
field in four dimensions is \(\phi^4\).

The theory is simple enough to compute with, but rich enough to display the main
features of perturbative QFT:
\begin{itemize}
    \item interaction vertices;
    \item scattering amplitudes;
    \item loop corrections;
    \item ultraviolet divergences;
    \item counterterms;
    \item renormalized masses and couplings.
\end{itemize}
For this reason, \(\phi^4\) theory is the standard first interacting quantum
field theory.

\section{The Lagrangian}

The classical Lagrangian density is
\[
    \mathcal{L}
    = \frac{1}{2}\partial_\mu\phi\partial^\mu\phi
      -\frac{1}{2}m^2\phi^2
      -\frac{\lambda}{4!}\phi^4.
\]
The free part is
\[
    \mathcal{L}_0
    = \frac{1}{2}\partial_\mu\phi\partial^\mu\phi
      -\frac{1}{2}m^2\phi^2,
\]
and the interaction part is
\[
    \mathcal{L}_{\mathrm{int}}
    = -\frac{\lambda}{4!}\phi^4.
\]
The factor \(4!\) is conventional. It cancels the number of permutations of four
identical scalar fields at a vertex, giving a simple Feynman rule.

The classical equation of motion is
\[
    (\Box+m^2)\phi+\frac{\lambda}{3!}\phi^3=0.
\]
Unlike the free Klein--Gordon equation, this equation is nonlinear. The
nonlinearity is the classical shadow of particle interactions in the quantum
theory.

\section{Dimensional Analysis}

In four spacetime dimensions, the action is dimensionless in units
\(\hbar=c=1\). Since
\[
    S=\int\dd^4x\,\mathcal{L},
\]
the Lagrangian density has mass dimension four. From the kinetic term one finds
\[
    [\phi]=1.
\]
Therefore
\[
    [\lambda]=0.
\]
The coupling \(\lambda\) is dimensionless in four dimensions. This is one reason
\(\phi^4\) theory is important: it is power-counting renormalizable in four
dimensions.

In dimensions other than four, the mass dimension of \(\lambda\) changes. This
will become important in the discussion of renormalization and effective field
theory.

\section{Path-Integral Generating Functional}

The generating functional is
\[
    Z[J]
    = \int\mathcal{D}\phi\,
      \exp\left(i\int\dd^4x\,
      \left[\frac{1}{2}\partial_\mu\phi\partial^\mu\phi
      -\frac{1}{2}m^2\phi^2
      -\frac{\lambda}{4!}\phi^4
      +J\phi\right]\right).
\]
It can be written in terms of the free generating functional as
\[
    Z[J]
    = \exp\left[-i\frac{\lambda}{4!}\int\dd^4x\,
      \left(\frac{1}{i}\frac{\delta}{\delta J(x)}\right)^4\right]Z_0[J].
\]
Expanding the exponential in powers of \(\lambda\) generates perturbation theory.
The free functional \(Z_0[J]\) supplies propagators; the derivative operator
supplies interaction vertices.

This is the functional-integral version of the Dyson expansion and Wick's
theorem.

\section{Feynman Rules in Position Space}

The basic ingredients are simple. The free propagator is
\[
    D_F(x-y)
    = \int\frac{\dd^4k}{(2\pi)^4}\,
      \frac{i}{k^2-m^2+i\epsilon}e^{-ik\cdot(x-y)}.
\]
Each interaction vertex contributes
\[
    -i\lambda\int\dd^4x.
\]
Four scalar lines meet at each vertex. Correlation functions are obtained by
summing over all contractions, or equivalently all diagrams, with the appropriate
symmetry factors.

Position-space diagrams are useful conceptually because they show that each
vertex is integrated over spacetime. Momentum-space diagrams are usually more
convenient for scattering amplitudes because translation invariance turns those
integrals into momentum-conserving delta functions.

\section{Feynman Rules in Momentum Space}

In momentum space, the propagator is
\[
    \frac{i}{k^2-m^2+i\epsilon}.
\]
The four-point vertex contributes
\[
    -i\lambda.
\]
At each vertex, four-momentum is conserved. For each undetermined internal
momentum, one integrates
\[
    \int\frac{\dd^4\ell}{(2\pi)^4}.
\]
Finally, one includes the symmetry factor of the diagram.

These rules compute contributions to correlation functions and, after LSZ
reduction, to scattering amplitudes.

\section{Tree-Level \texorpdfstring{\(2\to2\)}{2 to 2} Scattering}

The simplest nontrivial process is
\[
    \phi\phi\longrightarrow\phi\phi.
\]
At lowest order there is one four-point vertex. The amplitude is
\[
    i\mathcal{M}_{\mathrm{tree}}=-i\lambda,
    \qquad
    \mathcal{M}_{\mathrm{tree}}=-\lambda.
\]
This is a contact interaction. There is no internal propagator at this order.

The full matrix element contains the momentum-conserving delta function:
\[
    \langle p_3,p_4|i\hat T|p_1,p_2\rangle
    = i(2\pi)^4\delta^{(4)}(p_1+p_2-p_3-p_4)\mathcal{M}_{\mathrm{tree}}.
\]
This is the same structure introduced in scattering theory, now obtained from a
specific interacting Lagrangian.

\section{First Loop Correction: The Tadpole}

The two-point function receives a one-loop correction from the tadpole diagram.
In momentum space, the loop integral has the form
\[
    \int\frac{\dd^4\ell}{(2\pi)^4}\,
    \frac{i}{\ell^2-m^2+i\epsilon}.
\]
This integral diverges at large \(|\ell|\). The divergence is ultraviolet: it
comes from arbitrarily short-distance fluctuations.

The tadpole correction shifts the mass parameter. This is the first appearance
of an important lesson: the parameters in the Lagrangian are not automatically
the physical measured parameters. Quantum fluctuations correct them.

\section{One-Loop Four-Point Correction}

The four-point function receives one-loop corrections from diagrams with two
vertices and two internal lines. A typical loop integral has the schematic form
\[
    \int\frac{\dd^4\ell}{(2\pi)^4}\,
    \frac{i}{\ell^2-m^2+i\epsilon}
    \frac{i}{(\ell+p)^2-m^2+i\epsilon}.
\]
There are contributions in the \(s\)-, \(t\)-, and \(u\)-channels, corresponding
to different ways of routing the external momenta through the loop.

These loop corrections modify the effective strength of the four-point
interaction. In other words, they correct the coupling \(\lambda\).

\section{Ultraviolet Divergences}

The loop integrals of \(\phi^4\) theory are typically divergent. This does not
mean the theory is useless. It means that the parameters appearing in the
Lagrangian must be related carefully to measured quantities.

A regulator is introduced to make integrals finite. Examples include a momentum
cutoff \(\Lambda\) or dimensional regularization. A regulated divergent integral
might depend on \(\Lambda\), or contain poles in \(4-d\) when computed in
\(d\) dimensions.

The goal is not to keep the regulator as a physical object. The goal is to
absorb regulator dependence into redefinitions of parameters and then express
predictions in terms of finite measured quantities.

\section{Counterterms}

To renormalize the theory, write the Lagrangian in terms of renormalized
parameters plus counterterms:
\[
    \mathcal{L}
    = \frac{1}{2}Z_\phi\partial_\mu\phi\partial^\mu\phi
      -\frac{1}{2}Z_m m^2\phi^2
      -\frac{Z_\lambda\lambda}{4!}\phi^4.
\]
Equivalently, one may write
\[
    \mathcal{L}
    = \mathcal{L}_{\mathrm{ren}}+\mathcal{L}_{\mathrm{ct}},
\]
where
\[
    \mathcal{L}_{\mathrm{ct}}
    = \frac{1}{2}\delta Z\,\partial_\mu\phi\partial^\mu\phi
      -\frac{1}{2}\delta m^2\phi^2
      -\frac{\delta\lambda}{4!}\phi^4.
\]
The counterterms are chosen to cancel the divergences from loop diagrams.

In \(\phi^4\) theory in four dimensions, the required counterterms have the same
form as terms already present in the Lagrangian. This is the practical meaning
of perturbative renormalizability.

\section{Renormalized Parameters}

The mass and coupling in the Lagrangian are not directly observable until a
renormalization prescription is chosen. A renormalization condition defines the
parameters in terms of physical or reference quantities.

For example, one may define the physical mass by the pole of the full propagator.
The renormalized coupling may be defined by the value of the four-point amplitude
at a chosen reference momentum scale.

Different schemes define parameters differently, but physical predictions agree
when expressed in terms of measured quantities. The scheme dependence of
intermediate parameters is the beginning of the renormalization group story.

\section{Connected and One-Particle-Irreducible Functions}

The generating functional \(Z[J]\) produces all diagrams. The functional
\(W[J]=-i\ln Z[J]\) produces connected diagrams. The effective action
\(\Gamma[\phi_c]\) produces one-particle-irreducible diagrams.

In \(\phi^4\) theory:
\begin{itemize}
    \item the connected two-point function gives the full propagator;
    \item the 1PI two-point function gives the self-energy;
    \item the 1PI four-point function gives the quantum-corrected vertex;
    \item counterterms are fixed by imposing conditions on these functions.
\end{itemize}
This is the language used systematically in renormalization.

\section{Perturbative Expansion and Its Meaning}

A typical observable is computed as a formal power series:
\[
    \mathcal{M}
    = \mathcal{M}^{(0)}
      +\lambda\mathcal{M}^{(1)}
      +\lambda^2\mathcal{M}^{(2)}+\cdots.
\]
The superscript does not always count loops directly, but higher powers of
\(\lambda\) generally correspond to more vertices and more complicated diagrams.

Perturbation theory is useful when \(\lambda\) is small enough that the first few
terms give a good approximation. It is not a complete nonperturbative definition
of the theory. Nevertheless, it is the main calculational tool for weakly
coupled QFT.

\section{Physical Lessons from \texorpdfstring{\(\phi^4\)}{phi4} Theory}

The theory teaches several lessons that recur throughout QFT:
\begin{itemize}
    \item interactions are represented by vertices;
    \item amplitudes are built from propagators and vertices;
    \item loop diagrams express virtual quantum fluctuations;
    \item loop integrals can diverge at high momentum;
    \item counterterms absorb divergences into parameter definitions;
    \item measured masses and couplings are renormalized quantities.
\end{itemize}
These ideas reappear in QED, non-Abelian gauge theory, the Standard Model, and
effective field theory.

\begin{summarybox}
The \(\phi^4\) model is the first complete interacting QFT laboratory. Its
single four-scalar vertex gives tree-level scattering. Its loop diagrams produce
ultraviolet divergences that require regularization and counterterms. The theory
introduces the distinction between bare and renormalized parameters and shows why
physical predictions must be expressed in terms of measured quantities rather
than unrenormalized Lagrangian parameters.
\end{summarybox}

\section{Chapter Checklist}

After reading this chapter, one should be able to:
\begin{itemize}
    \item write the \(\phi^4\) Lagrangian and identify free and interacting
    parts;
    \item explain why the factor \(4!\) is included;
    \item derive the basic propagator and vertex Feynman rules;
    \item compute the tree-level \(2\to2\) amplitude;
    \item identify the tadpole as a one-loop mass correction;
    \item describe the one-loop correction to the four-point function;
    \item explain what an ultraviolet divergence is;
    \item state why counterterms are introduced;
    \item distinguish bare and renormalized parameters;
    \item explain why \(\phi^4\) theory is the first useful laboratory for
    renormalization.
\end{itemize}
