\chapter{Propagators as Quantum Correlation Functions}

\begin{remarkbox}
The previous chapters constructed free quantum fields and their particle states.
This chapter introduces the next central object: the propagator. A propagator is
not merely a line in a Feynman diagram. It is a vacuum correlation function of
quantum fields, and it encodes both the equation of motion and the particle
content of the theory.
\end{remarkbox}

\section{From Classical Green Functions to Quantum Correlators}

In classical field theory, a Green function solves an inhomogeneous differential
equation. For the Klein--Gordon operator, one writes
\[
    (\Box_x+m^2)G(x-y)=\delta^{(4)}(x-y).
\]
Different boundary conditions give different Green functions: retarded,
advanced, or other choices.

In quantum field theory, a closely related object appears, but with a different
interpretation. The basic two-point function is a vacuum expectation value of a
product of field operators. For a real scalar field, the simplest one is
\[
    \langle0|\hat\phi(x)\hat\phi(y)|0\rangle.
\]
This is not the value of a classical field at two points. It is a quantum
correlation between two field operators in the vacuum.

\section{Wightman Functions}

The positive-frequency Wightman function of the real scalar field is
\[
    \Delta^+(x-y)
    = \langle0|\hat\phi(x)\hat\phi(y)|0\rangle.
\]
Using the mode expansion, one finds
\[
    \Delta^+(x-y)
    = \int\frac{\dd^3k}{(2\pi)^3}\,
      \frac{1}{2\omega_{\vect{k}}}
      e^{-i\omega_{\vect{k}}(x^0-y^0)+i\vect{k}\cdot(\vect{x}-\vect{y})}.
\]
The reversed ordering gives
\[
    \Delta^-(x-y)
    = \langle0|\hat\phi(y)\hat\phi(x)|0\rangle.
\]
These functions measure vacuum correlations. Even in the vacuum, field
operators at different spacetime points are correlated.

The commutator function is the difference
\[
    [\hat\phi(x),\hat\phi(y)]
    = \Delta^+(x-y)-\Delta^-(x-y)
    = i\Delta(x-y).
\]
Microcausality says that this commutator vanishes for spacelike separation.

\section{Time-Ordered Products}

Scattering theory is built from time-ordered products. For two bosonic field
operators, the time-ordered product is defined by
\[
    T\hat\phi(x)\hat\phi(y)
    =
    \begin{cases}
        \hat\phi(x)\hat\phi(y), & x^0>y^0,\\
        \hat\phi(y)\hat\phi(x), & y^0>x^0.
    \end{cases}
\]
Equivalently,
\[
    T\hat\phi(x)\hat\phi(y)
    = \theta(x^0-y^0)\hat\phi(x)\hat\phi(y)
      +\theta(y^0-x^0)\hat\phi(y)\hat\phi(x).
\]
For fermionic fields, time ordering includes a minus sign when two fermion
operators are exchanged. This sign is one of the first places where fermionic
statistics enters perturbation theory.

\section{The Feynman Propagator}

The scalar Feynman propagator is the vacuum expectation value of the time-ordered
two-point function:
\[
    D_F(x-y)
    = \langle0|T\hat\phi(x)\hat\phi(y)|0\rangle.
\]
Using the Wightman functions, this may be written as
\[
    D_F(x-y)
    = \theta(x^0-y^0)\Delta^+(x-y)
      +\theta(y^0-x^0)\Delta^+(y-x).
\]
It has the four-dimensional momentum-space representation
\[
    D_F(x-y)
    = \int\frac{\dd^4k}{(2\pi)^4}\,
      \frac{i}{k^2-m^2+i\epsilon}
      e^{-ik\cdot(x-y)}.
\]
The small \(i\epsilon\) prescription specifies how the poles are passed in the
complex \(k^0\)-plane. This prescription is what makes the propagator the
vacuum time-ordered correlation function rather than some other Green function.

\section{The Propagator as a Green Function}

The Feynman propagator satisfies
\[
    (\Box_x+m^2)D_F(x-y)
    = -i\delta^{(4)}(x-y),
\]
with the sign convention following from the definition above. Thus the Feynman
propagator is also a Green function of the Klein--Gordon operator, but with a
specific quantum boundary condition.

This dual nature is important:
\[
    \text{propagator}
    = \text{Green function}
    = \text{vacuum two-point correlation function}.
\]
The first viewpoint emphasizes equations of motion. The second emphasizes
quantum states and operator expectation values.

\section{Retarded and Advanced Propagators}

The retarded propagator is built from the commutator:
\[
    D_R(x-y)
    = i\theta(x^0-y^0)[\hat\phi(x),\hat\phi(y)].
\]
It vanishes when \(x^0<y^0\), so it describes causal response to a disturbance
in the past. The advanced propagator is
\[
    D_A(x-y)
    = -i\theta(y^0-x^0)[\hat\phi(x),\hat\phi(y)].
\]
It vanishes when \(x^0>y^0\).

Retarded and advanced propagators are especially natural in classical response
problems and nonequilibrium physics. The Feynman propagator is the object that
appears in ordinary perturbative scattering theory.

\section{The \texorpdfstring{\(i\epsilon\)}{i-epsilon} Prescription}

The expression
\[
    \frac{i}{k^2-m^2+i\epsilon}
\]
contains more information than the algebraic denominator. The poles are located
near
\[
    k^0=\pm\omega_{\vect{k}}.
\]
The \(i\epsilon\) prescription tells us how to deform the integration contour
around these poles. It is the momentum-space encoding of time ordering.

For positive time separation, the contour picks up the positive-energy pole. For
negative time separation, it picks up the negative-energy pole in the way
appropriate to a time-ordered vacuum expectation value. This is how the compact
four-dimensional expression reproduces the two time orderings in position
space.

\section{Momentum-Space Propagators for Free Fields}

For the real scalar field,
\[
    D_F(k)=\frac{i}{k^2-m^2+i\epsilon}.
\]
For the complex scalar field,
\[
    \langle0|T\hat\phi(x)\hat\phi^\dagger(y)|0\rangle
\]
has the same momentum-space denominator. The difference is that the complex
scalar propagator carries charge flow from \(\phi^\dagger\) to \(\phi\).

For the Dirac field,
\[
    S_F(k)=\frac{i(\gamma^\mu k_\mu+m)}{k^2-m^2+i\epsilon}.
\]
The denominator is the same mass-shell structure, while the numerator carries
spinor information.

For the photon in Feynman gauge,
\[
    D_{\mu\nu}(k)=\frac{-i\eta_{\mu\nu}}{k^2+i\epsilon}.
\]
In a general covariant gauge, the numerator changes by gauge-dependent terms.
Physical amplitudes do not depend on that gauge choice.

\section{Pole Structure and Particle Interpretation}

The location of poles in a propagator reveals the particle content of the free
theory. The scalar propagator has poles at
\[
    k^2=m^2,
\]
which is the relativistic mass shell. This is the analytic trace of a particle
of mass \(m\).

For a massless photon, the pole is at
\[
    k^2=0.
\]
For a free Dirac field, the denominator again has poles at \(k^2=m^2\), while
the numerator encodes spinor structure.

In an interacting theory, the pole structure becomes even more informative. A
shift in the pole position corresponds to a mass correction. A pole with an
imaginary part corresponds to an unstable resonance. Branch cuts correspond to
multiparticle thresholds. These ideas will be developed later, but the free
propagator already contains the basic principle: particles appear as singular
structures of correlation functions.

\section{Propagation Versus Correlation}

The word ``propagator'' can be misleading. It suggests that a particle literally
travels from \(y\) to \(x\). In QFT the primary object is a correlation function
of fields. It measures the amplitude associated with inserting one field at
\(y\) and another at \(x\), with the vacuum in between.

For spacelike separation, the Feynman propagator need not vanish. This does not
violate causality, because the commutator vanishes outside the light cone. The
causal statement is not that all vacuum correlations vanish at spacelike
separation, but that spacelike separated observables commute.

Thus one should distinguish
\[
    \text{vacuum correlation}
    \neq
    \text{signal propagation}.
\]
The propagator is a correlation function used in the computation of amplitudes;
it is not by itself a directly observable signal.

\section{What a Propagator Means in Perturbation Theory}

In perturbation theory, propagators connect field insertions. Wick's theorem
will show that time-ordered products of free fields can be reduced to sums of
contractions, and each contraction is a Feynman propagator.

This is why propagators become internal lines in Feynman diagrams. A diagram is
a compact bookkeeping device for a term in the perturbative expansion. The line
stands for a two-point function of the corresponding free field.

For example, a scalar internal line carries
\[
    \frac{i}{k^2-m^2+i\epsilon},
\]
a fermion internal line carries
\[
    \frac{i(\gamma^\mu k_\mu+m)}{k^2-m^2+i\epsilon},
\]
and a photon internal line in Feynman gauge carries
\[
    \frac{-i\eta_{\mu\nu}}{k^2+i\epsilon}.
\]
The diagrammatic rules are therefore not arbitrary. They are built from the
free-field correlation functions developed in this chapter.

\section{Summary of the Main Propagators}

The most important free propagators introduced so far are
\[
\begin{array}{ccl}
\text{real scalar} &:& \displaystyle \frac{i}{k^2-m^2+i\epsilon},\\[1.0em]
\text{complex scalar} &:& \displaystyle \frac{i}{k^2-m^2+i\epsilon}
\text{ with charge flow},\\[1.0em]
\text{Dirac field} &:& \displaystyle
\frac{i(\gamma^\mu k_\mu+m)}{k^2-m^2+i\epsilon},\\[1.0em]
\text{photon, Feynman gauge} &:& \displaystyle
\frac{-i\eta_{\mu\nu}}{k^2+i\epsilon}.
\end{array}
\]
The denominators encode the mass shell. The numerators encode spin, vector, or
gauge structure.

\begin{summarybox}
A propagator is a time-ordered vacuum two-point function. It is also a Green
function with a specific quantum boundary condition. Its poles encode particle
content, while its numerator encodes spin and gauge structure. Propagators are
the building blocks of perturbative QFT because Wick's theorem turns
contractions of free fields into propagators.
\end{summarybox}

\section{Chapter Checklist}

After reading this chapter, one should be able to:
\begin{itemize}
    \item distinguish classical Green functions from quantum two-point functions;
    \item define Wightman functions;
    \item define the time-ordered product;
    \item write the scalar Feynman propagator in position and momentum space;
    \item explain the role of the \(i\epsilon\) prescription;
    \item distinguish Feynman, retarded, and advanced propagators;
    \item write the free scalar, Dirac, and photon propagators;
    \item explain how pole structure encodes particle content;
    \item explain why vacuum correlation is not the same as signal propagation;
    \item explain why propagators become internal lines in Feynman diagrams.
\end{itemize}
