\chapter{Gauge Fields in the Path Integral}

\begin{remarkbox}
Gauge fields require special care in the path integral because the functional
integral over \(A_\mu\) overcounts physically equivalent configurations. Gauge
fixing removes this overcounting. In non-Abelian theories, the gauge-fixing
procedure introduces ghost fields and leads naturally to BRST symmetry.
\end{remarkbox}

\section{Overcounting Gauge-Equivalent Configurations}

For electromagnetism, the gauge potential transforms as
\[
    A_\mu\longrightarrow A_\mu+\partial_\mu\Lambda.
\]
The two potentials describe the same physical electromagnetic field. Therefore a
naive path integral
\[
    \int\mathcal{D}A_\mu\,e^{iS[A]}
\]
integrates over infinitely many copies of the same physical configuration.

This is not merely an aesthetic problem. The redundancy makes the quadratic
operator non-invertible, so the propagator of \(A_\mu\) is not defined until one
chooses a gauge. In operator language this was the statement that not all
components of \(A_\mu\) are physical. In the path-integral language it becomes a
problem of dividing by the volume of the gauge group.

\section{Gauge Fixing in the Functional Integral}

A gauge condition is a function \(G[A]\) imposed on gauge fields. A common
covariant choice is
\[
    G[A]=\partial_\mu A^\mu.
\]
Instead of integrating over all representatives of each gauge orbit, one wants
to integrate once over each physical configuration. Formally,
\[
    \int\mathcal{D}A_\mu
    \quad\longrightarrow\quad
    \int\mathcal{D}A_\mu\,\delta(G[A])\,\Delta[A],
\]
where \(\Delta[A]\) is a determinant that compensates for the change of
variables from gauge orbits to gauge-fixed representatives.

In practice, one often replaces the sharp delta-function condition by a Gaussian
average over gauges. This leads to the covariant gauge-fixing term
\[
    \mathcal{L}_{\mathrm{gf}}
    = -\frac{1}{2\xi}(\partial_\mu A^\mu)^2.
\]
The parameter \(\xi\) labels the gauge. The choice \(\xi=1\) is Feynman gauge.

\section{The Faddeev--Popov Method}

The Faddeev--Popov method inserts a carefully chosen form of unity into the path
integral:
\[
    1=\Delta_{\mathrm{FP}}[A]
      \int\mathcal{D}\alpha\,\delta(G[A^\alpha]).
\]
Here \(A^\alpha\) denotes the gauge transform of \(A\), and
\(\Delta_{\mathrm{FP}}[A]\) is the Faddeev--Popov determinant.

The determinant is given formally by
\[
    \Delta_{\mathrm{FP}}[A]
    = \det\left(\frac{\delta G[A^\alpha]}{\delta\alpha}\right).
\]
This factor tells us how the gauge condition changes as one moves along a gauge
orbit. It is the Jacobian associated with gauge fixing.

For Abelian gauge theory in a linear covariant gauge, this determinant is
field-independent. It contributes only an overall constant and is usually
ignored. For non-Abelian gauge theory, the determinant depends on the gauge
field and has dynamical consequences.

\section{Ghost Fields}

A determinant can be represented as a Grassmann integral. For an operator \(M\),
\[
    \det M
    = \int\mathcal{D}\bar c\mathcal{D}c\,
      \exp\left(i\int\dd^4x\,\bar c M c\right),
\]
with appropriate conventions. The Grassmann fields \(c\) and \(\bar c\) are
called ghost fields.

Ghosts are anticommuting scalar fields. This phrase sounds paradoxical: they
have scalar Lorentz transformation properties but fermionic statistics. They are
not introduced because the theory has new physical particles. They are
introduced because the gauge-fixing determinant must be represented inside the
functional integral.

For non-Abelian gauge theory, ghosts interact with gauge fields. Their diagrams
are essential for unitarity and gauge invariance in covariant gauges.

\section{Why Ghosts Are Not Physical Particles}

Ghost fields appear as internal lines in Feynman diagrams, but they do not
appear as physical external states. They are artifacts of gauge fixing, not
quanta in the physical Hilbert space.

This is analogous to the unphysical photon polarizations discussed in Chapter 5.
Covariant quantization introduces redundant components for calculational
convenience. The formalism must then ensure that unphysical states do not appear
in physical observables.

Ghosts are part of that mechanism. Their contributions cancel gauge-dependent
and unphysical pieces of gauge-field loops. Removing them would spoil the
consistency of non-Abelian perturbation theory.

\section{Gauge-Fixed Photon Path Integral}

For QED, the gauge-fixed action is
\[
    S[A]
    = \int\dd^4x\,
      \left[-\frac{1}{4}F_{\mu\nu}F^{\mu\nu}
      -\frac{1}{2\xi}(\partial_\mu A^\mu)^2\right].
\]
The corresponding generating functional is
\[
    Z[J]
    = \int\mathcal{D}A\,
      \exp\left(iS[A]+i\int\dd^4x\,J_\mu A^\mu\right).
\]
Because the Abelian Faddeev--Popov determinant is field-independent in this
gauge, ghosts decouple from ordinary QED in covariant gauges.

The photon propagator in momentum space is
\[
    D_{\mu\nu}(k)
    = \frac{-i}{k^2+i\epsilon}
      \left(\eta_{\mu\nu}-(1-\xi)\frac{k_\mu k_\nu}{k^2+i\epsilon}\right).
\]
This is the same propagator introduced in Chapter 5, now derived as the inverse
of the gauge-fixed quadratic operator.

\section{Gauge-Fixed Yang--Mills Path Integral}

For a non-Abelian gauge field \(A_\mu^a\), the field strength is
\[
    F_{\mu\nu}^a
    = \partial_\mu A_\nu^a-\partial_\nu A_\mu^a
      +g f^{abc}A_\mu^b A_\nu^c.
\]
The Yang--Mills action is
\[
    S_{\mathrm{YM}}
    = -\frac{1}{4}\int\dd^4x\,F_{\mu\nu}^aF^{a\mu\nu}.
\]
A covariant gauge-fixing term is
\[
    \mathcal{L}_{\mathrm{gf}}
    = -\frac{1}{2\xi}(\partial_\mu A^{a\mu})^2.
\]
The Faddeev--Popov ghost Lagrangian is
\[
    \mathcal{L}_{\mathrm{gh}}
    = \bar c^a\,\partial^\mu D_\mu^{ab}c^b,
\]
where
\[
    D_\mu^{ab}=\delta^{ab}\partial_\mu+g f^{acb}A_\mu^c
\]
is the covariant derivative in the adjoint representation.

The full gauge-fixed path integral is schematically
\[
    Z[J]
    = \int\mathcal{D}A\mathcal{D}\bar c\mathcal{D}c\,
      e^{i\int\dd^4x(\mathcal{L}_{\mathrm{YM}}+\mathcal{L}_{\mathrm{gf}}+\mathcal{L}_{\mathrm{gh}}+J\cdot A)}.
\]
Unlike QED, the ghosts do not decouple. They interact with the gauge field.

\section{First Introduction to BRST Symmetry}

Gauge fixing seems to break gauge invariance because a particular gauge
condition has been chosen. However, the gauge-fixed theory has a remnant symmetry
called BRST symmetry. It mixes gauge fields, ghosts, and auxiliary fields.

At this stage, the details are less important than the role of the symmetry.
BRST symmetry keeps track of the original gauge redundancy after gauge fixing. It
ensures that physical observables are gauge independent and that unphysical
states decouple from the physical spectrum.

The physical state space can be characterized in terms of BRST cohomology:
physical states are BRST-closed, and states differing by BRST-exact terms are
identified. This language will become more important in the later chapter on
BRST symmetry.

\section{Gauge Independence of Physical Observables}

Intermediate quantities can depend on the gauge parameter \(\xi\). Propagators,
individual diagrams, and off-shell Green functions may change when \(\xi\)
changes. Physical observables should not.

Examples of physical observables include \(S\)-matrix elements between physical
states, cross sections, decay rates, and expectation values of gauge-invariant
operators. Gauge independence of these quantities is a consistency requirement
of the theory.

In perturbation theory, gauge independence may involve cancellations among many
diagrams, including ghost diagrams. Therefore it is generally wrong to assign
physical meaning to a single gauge-dependent diagram in isolation.

\section{Meaning of Gauge Fixing in Quantum Theory}

Gauge fixing is not the selection of a physically preferred gauge. It is a
method for coordinatizing the space of physically distinct configurations. Just
as one may choose coordinates on an ordinary manifold without changing the
geometry, one may choose a gauge without changing physical predictions.

The path integral makes this especially clear. The original integral over
\(A_\mu\) includes redundant gauge copies. Gauge fixing removes the redundancy
and supplies the correct determinant. Different gauges are different ways of
performing the same physical integral.

This point is crucial for non-Abelian theories. Without gauge fixing and ghosts,
the perturbative expansion would not correctly represent the quantum gauge
system.

\begin{summarybox}
Gauge fields cannot be integrated naively because gauge-equivalent
configurations are overcounted. Gauge fixing inserts a condition selecting one
representative per gauge orbit, together with the Faddeev--Popov determinant. In
Abelian covariant gauges the determinant decouples, while in non-Abelian gauge
theory it becomes a ghost-field action. Ghosts are not physical particles; they
are the functional-integral mechanism that preserves consistency of covariant
quantization. BRST symmetry records the original gauge redundancy after gauge
fixing.
\end{summarybox}

\section{Chapter Checklist}

After reading this chapter, one should be able to:
\begin{itemize}
    \item explain why the naive gauge-field path integral overcounts
    configurations;
    \item describe the role of a gauge-fixing condition;
    \item state the basic Faddeev--Popov identity;
    \item explain why determinants can be represented by ghost fields;
    \item distinguish ghosts from physical particles;
    \item write the gauge-fixed photon path integral;
    \item write the schematic gauge-fixed Yang--Mills path integral;
    \item explain why ghosts decouple in Abelian theory but not in non-Abelian
    theory;
    \item state the conceptual role of BRST symmetry;
    \item explain why physical observables must be gauge independent.
\end{itemize}
