\section{What combinatorics studies}

Combinatorics is the mathematics of counting possibilities. It studies how many objects can be formed when a set of rules is imposed. These objects may be arrangements, lists, selections, strings, words, groups, schedules, routes, or distributions of identical items among different boxes.

The central question is simple:
\[
\text{How many objects of this type can be formed?}
\]
The difficulty is not usually arithmetic. The difficulty is recognizing the structure of the construction. Before using a formula, we must understand what is being built.

In this chapter we will not use the language of probability. We will also not start from abstract set theory. Instead, we will build a practical and precise language for counting. Later, this language will be useful in many places, but here it is studied on its own.

\subsection*{Counting as a construction process}

A counting problem should be read as a recipe for constructing objects. For example, suppose we want to form a three-letter code from the letters
\[
A,B,C,D.
\]
Before counting, we must ask:
\begin{itemize}
    \item Does the order of the letters matter?
    \item May a letter be used more than once?
    \item Do we use all available letters, or only some of them?
    \item Are some objects indistinguishable from each other?
\end{itemize}
Different answers lead to different counting schemes.

\begin{theorembox}
\textbf{Multiplication principle.} If a construction is performed in several successive steps, and if there are \(m_1\) choices for the first step, \(m_2\) choices for the second step, and so on up to \(m_k\) choices for the \(k\)-th step, then the total number of possible constructions is
\[
m_1m_2\cdots m_k.
\]
\end{theorembox}

The multiplication principle is the engine behind most formulas in this chapter. Factorials, powers, binomial coefficients, and stars-and-bars formulas all arise by applying this principle carefully and then correcting for possible overcounting.

\subsection*{The six basic counting schemes}

We will study six basic schemes. They are organized by three questions: whether order matters, whether repetition is allowed, and whether all objects are used or only a fixed number of positions is filled.

\[
\begin{array}{c|c|c}
\text{scheme} & \text{order matters?} & \text{repetition allowed?} \\
\hline
\text{permutations without repetition} & \text{yes} & \text{no} \\
\text{permutations with repetition} & \text{yes} & \text{yes, through identical objects} \\
\text{variations without repetition} & \text{yes} & \text{no} \\
\text{variations with repetition} & \text{yes} & \text{yes} \\
\text{combinations without repetition} & \text{no} & \text{no} \\
\text{combinations with repetition} & \text{no} & \text{yes}
\end{array}
\]

The formulas are important, but the main goal is recognition. A formula used in the wrong situation gives a wrong answer even if the arithmetic is perfect.

\begin{remarkbox}
When solving a counting problem, do not start by asking which formula you remember. Start by asking what kind of object is being constructed.
\end{remarkbox}

\input{chapters/combinatorics/permutations_without_repetition}
\input{chapters/combinatorics/permutations_with_repetition}
\input{chapters/combinatorics/variations_without_repetition}
\input{chapters/combinatorics/variations_with_repetition}
\input{chapters/combinatorics/combinations_without_repetition}
\input{chapters/combinatorics/combinations_with_repetition}
\input{chapters/combinatorics/recognizing_counting_schemes}
