\section{Why Classical Field Theory Matters}

Classical field theory is the point at which many separate parts of physics begin
to look like one subject. Waves on a string, electromagnetic radiation, fluids,
elastic media, scalar fields, gauge fields, and even the metric of spacetime are
all described by quantities that vary from point to point and evolve in time. At
first these examples look unrelated. Field theory teaches us to see the common
structure behind them.

The first idea is simple. A mechanical system has a finite list of generalized
coordinates. A field theory replaces that finite list by a function, or by
several functions, defined over space or spacetime. This replacement changes the
scale of the problem. Instead of asking how one coordinate changes with time, we
ask how a whole spatially distributed object changes, propagates, stores energy,
interacts with sources, and responds to boundary conditions.

This shift is not only technical. It changes how one thinks about physics. A
field equation is local: it says how neighboring values of a field constrain one
another. A conservation law is local: it says how density can flow through the
boundary of a region. A symmetry is local or global: it says which changes of
description or physical transformations leave the model intact. The language of
field theory is therefore the language of locality, propagation, symmetry, and
conservation.

These notes are written as a bridge. On one side is classical mechanics: action,
energy, momentum, Hamiltonian evolution, and variational principles. On the other
side are quantum field theory, gauge theory, and modern geometry. The aim here is
not to quantize fields. The aim is to understand the classical layer so well that
later quantum and geometric structures no longer appear as magic.

\begin{remarkbox}
Classical field theory is not a preliminary obstacle before quantum field theory.
It is the structural skeleton on which quantum field theory, gauge theory, and
relativistic physics are built.
\end{remarkbox}

A good way to read these notes is to keep asking three questions. What is the
field? What equation does it satisfy? What physical mechanism does that equation
express? The formalism is introduced to answer these questions, not to hide them.
