\section{Models, Assumptions, Equations, and Interpretation}

A field equation is not a theory by itself. It becomes physics only after we say
what the field represents, what assumptions were made, what domain is being
studied, what boundary or initial data are allowed, and which quantities are
observable. Without that surrounding interpretation, an equation is only a
formal object.

The same mathematical equation may occur in different physical settings. The
wave equation may describe a vibrating string, sound in an idealized medium, a
small perturbation of a field around equilibrium, or a massless relativistic
scalar field. The equation alone does not tell us which one is meant. The model
does.

This is why the notes repeatedly move between four layers:
\[
    \text{assumptions}
    \longrightarrow
    \text{fields}
    \longrightarrow
    \text{equations}
    \longrightarrow
    \text{interpretation}.
\]
The order is important. Assumptions decide what is ignored. Fields decide what
kind of data the model assigns to spacetime. Equations decide how that data is
constrained and evolves. Interpretation tells us what the solutions mean.

A useful example is electromagnetism. One can write Maxwell's equations as four
vector equations, or more compactly as equations for the tensor \(F_{\mu\nu}\),
or geometrically as equations for a two-form. These are not three different
physical theories. They are three ways of organizing the same model. However,
once the potential \(A_\mu\) is introduced, gauge freedom appears. Then one must
understand the difference between a change of physical field and a change of
description.

\begin{summarybox}
The goal of these notes is not only to manipulate equations. The goal is to read
a field theory as a model: what it assumes, what its variables mean, how its
equations work, what it conserves, and what questions it can answer.
\end{summarybox}

A technically correct calculation can still be physically empty if the symbols
are not interpreted. Conversely, a clear physical idea needs formalism when it is
to become precise. The course is about learning to move in both directions.
