\section{Notation as a Language}

Field theory needs notation, but notation is not the subject. Symbols are useful
only when they make structure visible. Index notation, differential forms,
Fourier modes, covariant derivatives, and gamma matrices are introduced in these
notes because they help us say something clearly about fields.

The reader should not try to memorize every convention at once. At the beginning,
it is enough to know what kind of object is being discussed. Is it a scalar, a
vector, a tensor, a spinor, or a gauge potential? Does an index label spacetime,
space, an internal symmetry, or a component of a multiplet? Most mistakes in
field theory come from mixing these roles.

Spatial vectors are written in boldface, for example \(\vect{x}\), \(\vect{E}\),
and \(\vect{B}\). Spacetime indices are usually Greek, such as
\(\mu,\nu,\rho,\sigma\), while spatial indices are usually Latin, such as
\(i,j,k\). Internal indices for Lie algebras are often written as \(a,b,c\). The
same letter should not be read mechanically; it should be read as part of the
object's transformation meaning.

Repeated indices are summed when the Einstein convention is used. Thus an
expression such as
\[
    \partial_\mu F^{\mu\nu}=J^\nu
\]
contains one summed index, \(\mu\), and one free index, \(\nu\). The free index
tells us that this is a family of equations, one for each value of \(\nu\).

Units are also part of the language. In more physical examples, constants such
as \(c\), \(\hbar\), or electromagnetic constants may be displayed to clarify
dimensions. In more theoretical sections, natural units may be used to reduce
clutter. The point is not to hide dimensions, but to avoid letting constants
obscure the structure being studied.

\begin{remarkbox}
Notation should be read semantically. An index is not decoration; it tells us how
an object transforms, what it can be contracted with, and what kind of equation
is being written.
\end{remarkbox}

Detailed conventions are collected in Appendix A. The main text introduces only
what is needed at the moment it becomes useful.
