\section{Recommended Mathematical Background}

The notes assume some familiarity with classical mechanics, multivariable
calculus, linear algebra, and ordinary differential equations. The most important
skills are not exotic. The reader should be comfortable differentiating functions
of several variables, integrating by parts, manipulating matrices, reading vector
calculus identities, and following a variational argument.

Partial differential equations appear throughout the course. Full mathematical
rigour about existence and uniqueness is not the main goal, but one should
remember that a field equation is not solved by writing down the equation. A
solution requires a domain, initial data, boundary data, and regularity or falloff
conditions. These choices are part of the physical problem.

Special relativity enters because many important field theories are
relativistic. Tensor notation is introduced as needed. Differential forms, Lie
groups, and spinors appear later, but only to the extent needed for classical
field theory. The appendices provide reference material for these tools.

The best preparation is conceptual discipline. When a new symbol appears, ask
what object it denotes. When a derivative appears, ask with respect to which
variable. When a conservation law appears, ask what can flow through the
boundary. When a symmetry appears, ask whether it is physical or gauge.

\begin{summarybox}
The required mathematical attitude is more important than a long prerequisite
list: follow definitions carefully, track assumptions, check dimensions, control
boundary terms, and ask what each symbol represents physically.
\end{summarybox}
