\section{Structure of the Course}

The course is organized as a gradual widening of perspective. It begins with the
question of what a field is, then builds the variational and Hamiltonian
machinery, and only then studies the standard classical fields. This order is
intentional. The examples become easier to understand when the reader already
knows what an action is, what a symmetry does, and how a conserved current is
found.

Part I develops the foundations. It explains how field theory grows out of
mechanics, why fields have infinitely many degrees of freedom, how initial and
boundary data enter, and why relativistic notation is useful. This part is meant
to slow the reader down before the formalism accelerates.

Part II develops the general formalism. The central objects are the action
functional, the Lagrangian density, Euler--Lagrange equations, boundary terms,
Noether currents, conserved charges, Hamiltonian phase space, Poisson brackets,
and constraints. This is the technical core of the notes, but it should always
be read with the physical interpretation in mind.

Part III studies fundamental classical fields. The real scalar field is the
simplest model. The complex scalar field introduces internal symmetry and charge.
Electromagnetism introduces gauge freedom. Yang--Mills theory generalizes this to
non-Abelian gauge fields. Spinor fields introduce the classical Dirac equation and
prepare the reader for fermions in quantum field theory.

The appendices collect tools that should not interrupt the main line of the
course: conventions, variational calculus, Lie groups, differential forms,
Green's functions, and suggested problems. They are meant to be used repeatedly,
not necessarily read straight through at the beginning.

\begin{remarkbox}
The notes should be read actively. A field equation should be varied,
differentiated, dimensionally checked, solved in simple cases, and interpreted.
The point is not to admire the formalism, but to learn what it makes visible.
\end{remarkbox}

The intended outcome is not that the reader remembers many formulas. The intended
outcome is that the reader can look at a field theory and ask the right
questions: what is the field, what is the action, what are the symmetries, what
are the conserved quantities, what is gauge, what is physical, and what boundary
conditions make the problem meaningful?
