\section{What Is a Field?}

A field is a physical quantity spread over a continuous domain. The value of the
field may be a number, a vector, a tensor, a spinor, or a more general geometric
object. What matters is not the name of the object, but the fact that it assigns
some data to every point of space, spacetime, or another continuous background.

A temperature distribution in a solid is a scalar field: each point has a
number. The velocity of a fluid is a vector field: each point has a vector. The
electromagnetic field is richer: electric and magnetic fields are observed in a
chosen frame, while the relativistic theory packages them into the field strength
of a gauge potential. In gravity, the metric itself is a field; it tells us how
spacetime intervals are measured.

This is the conceptual move from particles to fields. A point particle has a
position. A field has a value everywhere. A particle trajectory is a curve. A
field history is an extended pattern in spacetime. A particle may move through a
region; a field may oscillate, radiate, form waves, carry energy, satisfy
constraints, or approach different boundary values at infinity.

The word ``field'' should therefore not be reduced to a formula. A formula such
as \(\phi(t,\vect{x})\) is only notation. The physical question is: what does
\(\phi\) measure? Is it a displacement, density, potential, phase, component of a
connection, or something else? The same mathematical type can represent very
different physics.

\begin{definitionbox}
A \emph{classical field} is a physical quantity represented by data assigned to
each point of a continuous domain. The field is classical when it is treated as a
definite object rather than as an operator or quantum state.
\end{definitionbox}

The simplest fields are scalar fields, but they are not the end of the story.
Vector fields lead naturally to electromagnetism. Lie-algebra-valued fields lead
to Yang--Mills theory. Spinor fields are needed for spin-one-half matter. The
course begins with the simplest examples because they teach the grammar; later
examples show why the grammar is powerful.
