\chapter{Variational Calculus for Fields}

\begin{remarkbox}
The variational calculus of fields is the technical engine behind the action
principle, Euler--Lagrange equations, Noether currents, stress-energy tensors,
and boundary terms. This appendix collects the core manipulations in one place.
\end{remarkbox}

\section{Variation of a Functional}

A functional assigns a number to an entire field configuration. The basic action
functional has the form
\[
    S[\phi]=\int_\Omega d^{d+1}x\,\mathcal{L}(\phi^A,\partial_\mu\phi^A,x).
\]
A variation is defined by considering a one-parameter family of fields
\[
    \phi^A_\varepsilon=\phi^A+\varepsilon\eta^A,
\]
where \(\eta^A\) is an arbitrary test variation satisfying whatever boundary
conditions are appropriate. The first variation is
\[
    \delta S[\phi;\eta]
    =\left.\frac{d}{d\varepsilon}S[\phi+\varepsilon\eta]\right|_{\varepsilon=0}.
\]

For a first-order Lagrangian density,
\[
    \delta S
    =\int_\Omega d^{d+1}x\left(
        \frac{\partial\mathcal{L}}{\partial\phi^A}\delta\phi^A
        +\frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}
        \delta(\partial_\mu\phi^A)
    \right).
\]
Since variation and partial differentiation commute,
\[
    \delta(\partial_\mu\phi^A)=\partial_\mu\delta\phi^A.
\]
Thus
\[
    \delta S
    =\int_\Omega d^{d+1}x\left(
        \frac{\partial\mathcal{L}}{\partial\phi^A}\delta\phi^A
        +\frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}
        \partial_\mu\delta\phi^A
    \right).
\]

\section{Integration by Parts}

The fundamental step is integration by parts:
\[
    \frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}
    \partial_\mu\delta\phi^A
    =\partial_\mu\left(
        \frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}
        \delta\phi^A
    \right)
    -\partial_\mu\left(
        \frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}
    \right)\delta\phi^A.
\]
Substituting into \(\delta S\) gives
\[
\begin{aligned}
    \delta S
    ={}&\int_\Omega d^{d+1}x\left[
        \frac{\partial\mathcal{L}}{\partial\phi^A}
        -\partial_\mu\left(
            \frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}
        \right)
    \right]\delta\phi^A \\
    &+\int_\Omega d^{d+1}x\,\partial_\mu\left(
        \frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}
        \delta\phi^A
    \right).
\end{aligned}
\]
Using Gauss's theorem, the second term becomes a boundary integral:
\[
    \int_{\partial\Omega} d\Sigma_\mu\,
    \frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}
    \delta\phi^A.
\]

The bulk coefficient of \(\delta\phi^A\) defines the Euler--Lagrange derivative:
\[
    E_A(\mathcal{L})
    =\frac{\partial\mathcal{L}}{\partial\phi^A}
     -\partial_\mu\left(
        \frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}
     \right).
\]
The Euler--Lagrange equations are
\[
    E_A(\mathcal{L})=0.
\]

\begin{theorembox}
For a first-order local action, stationarity under variations with vanishing
boundary term gives
\[
    \frac{\partial\mathcal{L}}{\partial\phi^A}
    -\partial_\mu\left(
        \frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}
    \right)=0.
\]
\end{theorembox}

\section{Boundary Terms}

Boundary terms are not mistakes. They decide which variational problem is being
posed. The general first variation can be written schematically as
\[
    \delta S=\int_\Omega d^{d+1}x\,E_A(\mathcal{L})\delta\phi^A
    +\int_{\partial\Omega}d\Sigma_\mu\,\Theta^\mu(\phi,\delta\phi).
\]
For first-order theories,
\[
    \Theta^\mu(\phi,\delta\phi)
    =\frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}\delta\phi^A.
\]

Common ways to control boundary terms include:
\begin{enumerate}
    \item requiring \(\delta\phi^A=0\) on \(\partial\Omega\), called Dirichlet
    boundary conditions;
    \item imposing boundary conditions on normal derivatives or momenta;
    \item adding a boundary term to the action so that the desired boundary data
    define a well-posed variational principle;
    \item requiring sufficiently fast falloff at infinity.
\end{enumerate}

A total derivative added to the Lagrangian density,
\[
    \mathcal{L}\longrightarrow \mathcal{L}+\partial_\mu K^\mu,
\]
does not change the bulk Euler--Lagrange equations, but it changes the boundary
term. Therefore it can change canonical momenta, Noether charges, and the correct
boundary conditions.

\begin{remarkbox}
A statement such as ``total derivatives do not matter'' is incomplete. They do
not change bulk equations under appropriate assumptions, but they can matter for
boundaries, charges, and Hamiltonian generators.
\end{remarkbox}

\section{Functional Derivatives}

The functional derivative is defined by
\[
    \delta F[\phi]
    =\int d^{d+1}x\,\frac{\delta F}{\delta\phi^A(x)}\delta\phi^A(x).
\]
For the action,
\[
    \frac{\delta S}{\delta\phi^A(x)}=E_A(\mathcal{L})(x)
\]
when boundary terms are discarded or controlled.

A basic identity is
\[
    \frac{\delta\phi^A(y)}{\delta\phi^B(x)}
    =\delta^A{}_B\delta^{(d+1)}(y-x).
\]
For example, if
\[
    F[\phi]=\int d^d x\,\frac{1}{2}\phi(x)^2,
\]
then
\[
    \delta F=\int d^d x\,\phi(x)\delta\phi(x),
\]
so
\[
    \frac{\delta F}{\delta\phi(x)}=\phi(x).
\]
If
\[
    F[\phi]=\int d^d x\,\frac{1}{2}|\nabla\phi|^2,
\]
then
\[
    \delta F=\int d^d x\,\nabla\phi\cdot\nabla\delta\phi
    =-\int d^d x\,\nabla^2\phi\,\delta\phi
\]
up to a boundary term. Hence
\[
    \frac{\delta F}{\delta\phi(x)}=-\nabla^2\phi(x).
\]

\section{Variation with Constraints}

Sometimes variations are constrained. A simple finite-dimensional analogy is the
method of Lagrange multipliers. In field theory, suppose one wants to extremize
\(S[\phi]\) subject to
\[
    C[\phi]=0.
\]
One introduces a multiplier \(\lambda\) and varies
\[
    S_\lambda[\phi]=S[\phi]+\lambda C[\phi].
\]
For local constraints
\[
    C(x,\phi,\partial\phi)=0,
\]
one uses a multiplier field \(\lambda(x)\):
\[
    S_\lambda[\phi,\lambda]
    =S[\phi]+\int d^{d+1}x\,\lambda(x)C(x,\phi,\partial\phi).
\]
Variation with respect to \(\lambda\) imposes the constraint, while variation
with respect to \(\phi\) gives modified field equations.

Constraints also arise in Hamiltonian field theory when the Legendre transform
is degenerate. In that setting, constraints are equations on phase space, such
as
\[
    \chi(\phi,\pi)\approx0.
\]
They must be handled with Dirac's constraint algorithm.

\section{Variation with Respect to the Metric}

The stress-energy tensor can be obtained by varying the matter action with
respect to the metric. A common convention is
\[
    T_{\mu\nu}=-\frac{2}{\sqrt{|g|}}\frac{\delta S_{\mathrm{matter}}}{\delta g^{\mu\nu}}.
\]
Equivalently,
\[
    \delta S_{\mathrm{matter}}
    =-\frac{1}{2}\int d^{d+1}x\,\sqrt{|g|}\,T_{\mu\nu}\delta g^{\mu\nu}.
\]
The variation of the determinant is
\[
    \delta\sqrt{|g|}=-\frac{1}{2}\sqrt{|g|}\,g_{\mu\nu}\delta g^{\mu\nu}.
\]
Also,
\[
    \delta g_{\mu\nu}=-g_{\mu\rho}g_{\nu\sigma}\delta g^{\rho\sigma}.
\]

For a scalar field in curved spacetime,
\[
    S=-\int d^{d+1}x\sqrt{|g|}\left(
        \frac{1}{2}g^{\mu\nu}\partial_\mu\phi\partial_\nu\phi+V(\phi)
    \right)
\]
with sign conventions adjusted to the chosen metric convention. Varying with
respect to \(g^{\mu\nu}\) gives the Hilbert stress-energy tensor.

\section{Common Variational Identities}

Useful identities include
\[
    \delta(AB)=(\delta A)B+A(\delta B),
\]
\[
    \delta(A^{-1})=-A^{-1}(\delta A)A^{-1},
\]
for an invertible matrix \(A\), and
\[
    \delta\det A=(\det A)\operatorname{tr}(A^{-1}\delta A).
\]
For gauge theory,
\[
    \delta F_{\mu\nu}=D_\mu\delta A_\nu-D_\nu\delta A_\mu
\]
in the non-Abelian case, while in the Abelian case
\[
    \delta F_{\mu\nu}=\partial_\mu\delta A_\nu-\partial_\nu\delta A_\mu.
\]
For the Dirac field,
\(\psi\) and \(\bar\psi\) are varied independently:
\[
    \delta\bar\psi\neq (\delta\psi)^\dagger\gamma^0
\]
inside the variational calculation.

\begin{summarybox}
The core variational pattern is always the same: vary the action, integrate by
parts, separate bulk terms from boundary terms, impose boundary conditions or add
boundary counterterms, and read off the Euler--Lagrange equations from the bulk
coefficient of arbitrary variations.
\end{summarybox}
