\chapter{Lie Groups and Lie Algebras}

\begin{remarkbox}
Gauge theory is written in the language of Lie groups and Lie algebras. This
appendix gives only the minimal toolkit needed for the main text: generators,
structure constants, representations, adjoint action, the exponential map, and
basic examples.
\end{remarkbox}

\section{Groups, Algebras, and Generators}

A group is a set with multiplication, an identity element, and inverses. A Lie
group is a group that is also a smooth manifold, with smooth multiplication and
inversion. Examples include
\[
    U(1),\qquad SU(2),\qquad SU(3),\qquad SO(3),\qquad SO(1,3).
\]

The Lie algebra \(\mathfrak{g}\) of a Lie group \(G\) is the tangent space at the
identity element, equipped with a bracket. For matrix groups, the bracket is the
commutator:
\[
    [X,Y]=XY-YX.
\]
A basis of the Lie algebra is denoted
\[
    T_a,\qquad a=1,\ldots,\dim\mathfrak{g}.
\]
A general Lie-algebra element is
\[
    X=X^aT_a.
\]

In gauge theory, the gauge potential is Lie-algebra-valued:
\[
    A_\mu=A_\mu^aT_a.
\]
This means that for each spacetime index \(\mu\), the object \(A_\mu\) is an
internal-space generator-valued field.

\begin{definitionbox}
The Lie algebra of a matrix Lie group is the vector space of infinitesimal group
generators, equipped with the commutator bracket.
\end{definitionbox}

The normalization of generators is conventional. Some texts use Hermitian
generators, others use anti-Hermitian generators. In these notes, the Yang--Mills
chapter uses anti-Hermitian generators unless stated otherwise.

\section{Structure Constants}

The Lie bracket of basis generators is written
\[
    [T_a,T_b]=f_{ab}{}^cT_c.
\]
The coefficients \(f_{ab}{}^c\) are the structure constants of the algebra. They
are antisymmetric in the lower indices:
\[
    f_{ab}{}^c=-f_{ba}{}^c.
\]
The Jacobi identity is
\[
    [T_a,[T_b,T_c]]+[T_b,[T_c,T_a]]+[T_c,[T_a,T_b]]=0.
\]
In terms of structure constants,
\[
    f_{ab}{}^d f_{dc}{}^e
    +f_{bc}{}^d f_{da}{}^e
    +f_{ca}{}^d f_{db}{}^e=0.
\]

For an Abelian Lie algebra, all commutators vanish:
\[
    [T_a,T_b]=0.
\]
Therefore
\[
    f_{ab}{}^c=0.
\]
The Lie algebra of \(U(1)\) is Abelian. The Lie algebras of \(SU(2)\) and
\(SU(3)\) are non-Abelian.

\begin{remarkbox}
The nonzero structure constants are the algebraic source of the commutator term
in the Yang--Mills field strength:
\[
    F_{\mu\nu}^a=\partial_\mu A_\nu^a-\partial_\nu A_\mu^a
    +g f_{bc}{}^a A_\mu^bA_\nu^c.
\]
\end{remarkbox}

\section{Representations}

A representation of a Lie group \(G\) is a way of realizing its elements as
linear transformations on a vector space. If a matter field \(\psi\) takes values
in such a vector space, a group element \(U\in G\) acts by
\[
    \psi\longrightarrow U\psi.
\]
At the Lie-algebra level, generators act as matrices:
\[
    T_a\longrightarrow R(T_a).
\]
The representation must preserve the bracket:
\[
    [R(T_a),R(T_b)]=f_{ab}{}^cR(T_c).
\]

The covariant derivative acting on matter in a representation \(R\) is
\[
    D_\mu\psi=\partial_\mu\psi+gA_\mu^aR(T_a)\psi.
\]
Different matter fields may transform in different representations of the same
gauge group.

\begin{definitionbox}
A representation tells us how abstract group or algebra elements act on actual
field components. Gauge fields transform in the adjoint representation, while
matter fields may transform in fundamental, adjoint, or other representations.
\end{definitionbox}

\section{Adjoint Representation}

The adjoint representation is the representation of a Lie algebra on itself. A
Lie-algebra element \(X\) acts on another element \(Y\) by commutation:
\[
    \operatorname{ad}_X(Y)=[X,Y].
\]
For a basis generator,
\[
    \operatorname{ad}_{T_a}(T_b)=f_{ab}{}^cT_c.
\]
Thus the matrices of the adjoint representation are
\[
    (\operatorname{ad}_{T_a})_b{}^c=f_{ab}{}^c.
\]

Lie-algebra-valued fields such as \(F_{\mu\nu}\) transform in the adjoint
representation:
\[
    F_{\mu\nu}\longrightarrow UF_{\mu\nu}U^{-1}.
\]
Infinitesimally,
\[
    \delta_\alpha F_{\mu\nu}=[\alpha,F_{\mu\nu}].
\]
The adjoint covariant derivative is
\[
    D_\mu X=\partial_\mu X+g[A_\mu,X].
\]

\section{Exponential Map}

The exponential map connects Lie-algebra elements to Lie-group elements. For a
matrix Lie group,
\[
    U=\exp(\alpha^aT_a).
\]
For infinitesimal \(\alpha^a\),
\[
    U=1+\alpha^aT_a+O(\alpha^2).
\]
The exponential map allows us to move between finite and infinitesimal
transformations.

For \(U(1)\), the group element is
\[
    e^{i\alpha}.
\]
For non-Abelian groups, the exponential is matrix-valued, and the order of
multiplication matters. In general,
\[
    e^X e^Y\neq e^{X+Y}
\]
when \([X,Y]\neq0\). The Baker--Campbell--Hausdorff formula starts as
\[
    e^X e^Y=\exp\left(X+Y+\frac{1}{2}[X,Y]+\cdots\right).
\]

\section{Gauge Transformations}

A local gauge transformation is a spacetime-dependent group element
\[
    U=U(x).
\]
Matter fields transform as
\[
    \psi\longrightarrow \psi'=U\psi.
\]
The connection transforms as
\[
    A'_\mu=U A_\mu U^{-1}-\frac{1}{g}(\partial_\mu U)U^{-1}.
\]
The field strength transforms covariantly:
\[
    F'_{\mu\nu}=UF_{\mu\nu}U^{-1}.
\]
Therefore \(F_{\mu\nu}\) itself is not gauge invariant in a non-Abelian theory,
but traces such as
\[
    \operatorname{tr}(F_{\mu\nu}F^{\mu\nu})
\]
are gauge invariant.

For an infinitesimal gauge parameter
\[
    \alpha=\alpha^aT_a,
\]
one obtains
\[
    \delta_\alpha A_\mu=-\frac{1}{g}\partial_\mu\alpha+[\alpha,A_\mu]
    =-\frac{1}{g}D_\mu\alpha.
\]
The exact sign depends on the convention chosen for the covariant derivative and
finite transformation law.

\section{Examples: \texorpdfstring{\(U(1)\), \(SU(2)\), \(SU(3)\), \(SO(3)\), \(SO(1,3)\)}{U(1), SU(2), SU(3), SO(3), SO(1,3)}}

The group \(U(1)\) consists of phases
\[
    e^{i\alpha}.
\]
Its Lie algebra is one-dimensional and Abelian. This is the gauge group of
electromagnetism.

The group \(SU(2)\) consists of unitary \(2\times2\) matrices with determinant
one. Its Lie algebra is three-dimensional. In a Hermitian-generator convention,
one often uses Pauli matrices; in an anti-Hermitian convention, one uses
multiples of \(i\sigma_a\). The structure constants are proportional to
\(\epsilon_{ab}{}^c\).

The group \(SU(3)\) consists of unitary \(3\times3\) matrices with determinant
one. Its Lie algebra is eight-dimensional. It is the gauge group of the strong
interaction in quantum chromodynamics.

The group \(SO(3)\) describes rotations of Euclidean three-dimensional space.
Its Lie algebra is closely related to \(\mathfrak{su}(2)\), but the global group
structure differs: \(SU(2)\) is the double cover of \(SO(3)\).

The proper orthochronous Lorentz group \(SO^+(1,3)\) preserves the Minkowski
metric and the orientation of time. Its double cover is \(SL(2,\mathbb{C})\),
which is why spinors naturally transform under \(SL(2,\mathbb{C})\) rather than
directly under \(SO^+(1,3)\).

\begin{summarybox}
Lie groups describe continuous symmetries; Lie algebras describe their
infinitesimal generators. In gauge theory, potentials are Lie-algebra-valued,
curvatures contain structure constants, matter fields transform in
representations, and gauge-invariant quantities are built by contracting internal
indices using invariant structures such as traces.
\end{summarybox}
