\chapter{Gauge Fields and Yang--Mills Theory}

\begin{remarkbox}
Yang--Mills theory is the non-Abelian generalization of electromagnetism. The
main new feature is that the gauge field itself carries charge and therefore
interacts with itself. Mathematically, the electromagnetic potential is replaced
by a Lie-algebra-valued connection, and the field strength becomes its curvature.
\end{remarkbox}

\section{From Electromagnetism to Non-Abelian Gauge Fields}

Electromagnetism is a gauge theory with gauge group \(U(1)\). Its gauge
potential is a one-form
\[
    A=A_\mu dx^\mu,
\]
and its field strength is
\[
    F=dA.
\]
In components,
\[
    F_{\mu\nu}=\partial_\mu A_\nu-\partial_\nu A_\mu.
\]
Because \(U(1)\) is Abelian, there is no commutator term.

Yang--Mills theory starts from the same idea but replaces \(U(1)\) by a general
Lie group \(G\), such as \(SU(2)\) or \(SU(3)\). The gauge potential is no longer
an ordinary real-valued one-form. It is a Lie-algebra-valued one-form,
\[
    A=A_\mu dx^\mu,
    \qquad
    A_\mu=A_\mu^a T_a,
\]
where \(T_a\) are generators of the Lie algebra \(\mathfrak{g}\) of \(G\).

In this chapter we use an anti-Hermitian generator convention,
\[
    [T_a,T_b]=f_{ab}{}^c T_c,
\]
with real structure constants \(f_{ab}{}^c\). With this convention, the gauge
coupling appears in the covariant derivative as
\[
    D_\mu=\partial_\mu+gA_\mu.
\]
Other texts often use Hermitian generators and write factors of \(i\). The
physics is the same, but signs and factors of \(i\) move between definitions.

\begin{definitionbox}
A \emph{Yang--Mills field} is a gauge field whose potential takes values in a
non-Abelian Lie algebra. Its field strength contains both derivative terms and a
commutator term.
\end{definitionbox}

The essential jump from electromagnetism to Yang--Mills theory is therefore
\[
    F=dA
    \qquad\longrightarrow\qquad
    F=dA+gA\wedge A.
\]
The second term is absent in Abelian electromagnetism and is responsible for the
self-interaction of non-Abelian gauge fields.

\section{Lie Groups and Lie Algebras: Minimal Toolkit}

A Lie group is a group that is also a smooth manifold. Its elements can be
multiplied and inverted smoothly. A Lie algebra is the tangent space to the Lie
group at the identity, equipped with a bracket. For a matrix group, the bracket
is the commutator.

Let \(G\) be a Lie group and \(\mathfrak{g}\) its Lie algebra. A basis of
\(\mathfrak{g}\) will be denoted by \(T_a\). The Lie bracket is
\[
    [T_a,T_b]=f_{ab}{}^cT_c.
\]
The constants \(f_{ab}{}^c\) are the structure constants. They are antisymmetric
in the lower indices:
\[
    f_{ab}{}^c=-f_{ba}{}^c.
\]
They also satisfy the Jacobi identity,
\[
    [T_a,[T_b,T_c]]+[T_b,[T_c,T_a]]+[T_c,[T_a,T_b]]=0.
\]
In components this gives
\[
    f_{ab}{}^d f_{dc}{}^e
    +f_{bc}{}^d f_{da}{}^e
    +f_{ca}{}^d f_{db}{}^e=0.
\]

\begin{examplebox}
For \(SU(2)\), the Lie algebra is three-dimensional and the structure constants
are proportional to the Levi-Civita symbol:
\[
    f_{ab}{}^c=\epsilon_{ab}{}^c
\]
in a standard normalization. This is why \(SU(2)\) gauge theory has three gauge
potentials \(A_\mu^1,A_\mu^2,A_\mu^3\).
\end{examplebox}

Yang--Mills theory also requires an invariant inner product on the Lie algebra.
For matrix groups this is often built from a trace. We write schematically
\[
    \langle X,Y\rangle=-\operatorname{tr}(XY)
\]
for anti-Hermitian matrices, up to a conventional normalization. The invariance
condition is
\[
    \langle [Z,X],Y\rangle+\langle X,[Z,Y]\rangle=0.
\]
This property allows us to build gauge-invariant actions from expressions such
as \(\langle F_{\mu\nu},F^{\mu\nu}\rangle\).

\section{Gauge Connection}

The gauge potential of Yang--Mills theory is best understood as a connection.
In coordinates it is written
\[
    A_\mu=A_\mu^aT_a.
\]
The role of a connection is to define a covariant derivative for fields that
transform under the gauge group.

Suppose a matter field \(\psi\) takes values in a representation space of
\(G\). The gauge covariant derivative is
\[
    D_\mu\psi=\partial_\mu\psi+gA_\mu\psi,
\]
where \(A_\mu\) acts on \(\psi\) through the chosen representation. The ordinary
derivative \(\partial_\mu\psi\) does not transform covariantly under local gauge
transformations. The connection term corrects this failure.

Under a local gauge transformation
\[
    U(x)\in G,
\]
the matter field transforms as
\[
    \psi\longrightarrow \psi'=U\psi.
\]
We demand that the covariant derivative transform in the same way:
\[
    D'_\mu\psi'=U D_\mu\psi.
\]
This requirement determines the transformation law of the connection:
\[
    A'_\mu=U A_\mu U^{-1}-\frac{1}{g}(\partial_\mu U)U^{-1}.
\]

\begin{definitionbox}
A \emph{gauge connection} is a Lie-algebra-valued one-form \(A=A_\mu dx^\mu\)
whose transformation law makes the covariant derivative transform covariantly
under local gauge transformations.
\end{definitionbox}

For an infinitesimal gauge transformation
\[
    U=1+\alpha+O(\alpha^2),
    \qquad
    \alpha=\alpha^aT_a,
\]
the connection transforms as
\[
    \delta_\alpha A_\mu=-\frac{1}{g}\partial_\mu\alpha+[\alpha,A_\mu].
\]
Equivalently,
\[
    \delta_\alpha A_\mu=-\frac{1}{g}D_\mu\alpha,
\]
where the covariant derivative in the adjoint representation is
\[
    D_\mu\alpha=\partial_\mu\alpha+g[A_\mu,\alpha].
\]

\section{Curvature / Field Strength}

The Yang--Mills field strength is the curvature of the connection. In
differential-form notation,
\[
    F=dA+gA\wedge A.
\]
In components,
\[
    F_{\mu\nu}=\partial_\mu A_\nu-\partial_\nu A_\mu+g[A_\mu,A_\nu].
\]
Writing
\[
    F_{\mu\nu}=F_{\mu\nu}^aT_a,
\]
we get
\[
    F_{\mu\nu}^a
    =\partial_\mu A_\nu^a-\partial_\nu A_\mu^a
     +g f_{bc}{}^a A_\mu^b A_\nu^c.
\]
The final term is the non-Abelian term. It is zero for an Abelian Lie algebra.

The field strength also appears as the commutator of covariant derivatives. If
\[
    D_\mu=\partial_\mu+gA_\mu,
\]
then
\[
    [D_\mu,D_\nu]\psi=gF_{\mu\nu}\psi.
\]
Thus curvature measures the failure of covariant derivatives to commute.

\begin{theorembox}
The field strength of a Yang--Mills connection is
\[
    F_{\mu\nu}=\partial_\mu A_\nu-\partial_\nu A_\mu+g[A_\mu,A_\nu].
\]
The commutator term is responsible for the self-interaction of non-Abelian gauge
fields.
\end{theorembox}

This equation should be compared with electromagnetism:
\[
    F_{\mu\nu}^{U(1)}=\partial_\mu A_\nu-\partial_\nu A_\mu.
\]
The difference is not cosmetic. In Yang--Mills theory, even the free-looking
field strength is nonlinear in the potential.

\section{Gauge Transformations}

The finite gauge transformation of the connection is
\[
    A'_\mu=U A_\mu U^{-1}-\frac{1}{g}(\partial_\mu U)U^{-1}.
\]
The field strength transforms covariantly:
\[
    F'_{\mu\nu}=U F_{\mu\nu} U^{-1}.
\]
This is the non-Abelian analogue of gauge invariance. Notice the difference:
in electromagnetism the field strength itself is gauge invariant, while in
Yang--Mills theory it is gauge covariant.

A trace or invariant inner product of two field strengths is gauge invariant.
Indeed,
\[
    \operatorname{tr}(F'_{\mu\nu}F'^{\mu\nu})
    =\operatorname{tr}(UF_{\mu\nu}U^{-1}UF^{\mu\nu}U^{-1})
    =\operatorname{tr}(F_{\mu\nu}F^{\mu\nu}).
\]
This is why the Yang--Mills action is built from a trace or invariant inner
product.

\begin{remarkbox}
In a non-Abelian theory, the potential \(A_\mu\) is gauge-dependent, and the
field strength \(F_{\mu\nu}\) is gauge-covariant rather than gauge-invariant.
Gauge-invariant observables must be built by contracting group indices in an
invariant way or by using nonlocal objects such as Wilson loops.
\end{remarkbox}

An infinitesimal gauge transformation gives
\[
    \delta_\alpha F_{\mu\nu}=[\alpha,F_{\mu\nu}].
\]
Thus \(F_{\mu\nu}\) rotates in the adjoint representation of the gauge group.

\section{Covariant Derivative}

The covariant derivative depends on the representation of the field it acts on.
For a matter field \(\psi\) in a representation \(R\),
\[
    D_\mu\psi=\partial_\mu\psi+gA_\mu^a R(T_a)\psi.
\]
For a Lie-algebra-valued field \(X=X^aT_a\), such as \(F_{\mu\nu}\), the
covariant derivative is taken in the adjoint representation:
\[
    D_\mu X=\partial_\mu X+g[A_\mu,X].
\]
In components,
\[
    (D_\mu X)^a=\partial_\mu X^a+g f_{bc}{}^a A_\mu^b X^c.
\]

The purpose of the covariant derivative is not only to preserve notation. It
ensures that equations transform homogeneously under local gauge transformations.
For example, the Yang--Mills equation will have the covariant form
\[
    D_\mu F^{\mu\nu}=J^\nu.
\]
Both sides transform in the adjoint representation.

\begin{definitionbox}
The adjoint covariant derivative of a Lie-algebra-valued field \(X\) is
\[
    D_\mu X=\partial_\mu X+g[A_\mu,X].
\]
It replaces the ordinary derivative whenever the object being differentiated
carries non-Abelian gauge charge.
\end{definitionbox}

In Abelian theory, the commutator vanishes and the adjoint covariant derivative
reduces to the ordinary derivative. This is another way to see why Maxwell
theory is structurally simpler than Yang--Mills theory.

\section{Yang--Mills Action}

The Yang--Mills Lagrangian density is
\[
    \mathcal{L}_{\mathrm{YM}}
    =-\frac{1}{4}\langle F_{\mu\nu},F^{\mu\nu}\rangle.
\]
In a component normalization where
\[
    \langle T_a,T_b\rangle=\delta_{ab},
\]
this becomes
\[
    \mathcal{L}_{\mathrm{YM}}
    =-\frac{1}{4}F_{\mu\nu}^aF^{a\mu\nu}.
\]
The action is
\[
    S_{\mathrm{YM}}[A]
    =-\frac{1}{4}\int d^{d+1}x\,F_{\mu\nu}^aF^{a\mu\nu}.
\]

The action is gauge invariant because the field strength transforms
covariantly and the inner product on the Lie algebra is invariant. It is also
Lorentz invariant because all spacetime indices are contracted.

\begin{definitionbox}
The \emph{Yang--Mills action} is
\[
    S_{\mathrm{YM}}[A]
    =-\frac{1}{4}\int d^{d+1}x\,F_{\mu\nu}^aF^{a\mu\nu}.
\]
It is the natural non-Abelian generalization of the Maxwell action.
\end{definitionbox}

Expanding the field strength shows that the Yang--Mills action contains terms
quadratic, cubic, and quartic in \(A_\mu\). Schematically,
\[
    F^2\sim (\partial A)^2+g(\partial A)A^2+g^2A^4.
\]
Thus non-Abelian gauge fields self-interact even without matter.

\section{Yang--Mills Equations}

Varying the Yang--Mills action gives the Yang--Mills field equations. The
variation of the curvature is
\[
    \delta F_{\mu\nu}=D_\mu\delta A_\nu-D_\nu\delta A_\mu.
\]
Using antisymmetry of \(F^{\mu\nu}\), the variation of the action is
\[
    \delta S_{\mathrm{YM}}
    =-\frac{1}{2}\int d^{d+1}x\,\langle F^{\mu\nu},\delta F_{\mu\nu}\rangle
    =-\int d^{d+1}x\,\langle F^{\mu\nu},D_\mu\delta A_\nu\rangle.
\]
After integrating by parts covariantly and discarding the boundary term, one
obtains
\[
    \delta S_{\mathrm{YM}}
    =\int d^{d+1}x\,\langle D_\mu F^{\mu\nu},\delta A_\nu\rangle.
\]
Therefore the vacuum Yang--Mills equation is
\[
    D_\mu F^{\mu\nu}=0.
\]
In components,
\[
    (D_\mu F^{\mu\nu})^a
    =\partial_\mu F^{a\mu\nu}
     +g f_{bc}{}^a A_\mu^b F^{c\mu\nu}=0.
\]

With matter, the equation becomes
\[
    D_\mu F^{\mu\nu}=J^\nu,
\]
where \(J^\nu\) is a Lie-algebra-valued matter current. Consistency requires a
covariant conservation law,
\[
    D_\nu J^\nu=0,
\]
rather than the ordinary conservation law \(\partial_\nu J^\nu=0\).

\begin{theorembox}
The source-free Yang--Mills equation is
\[
    D_\mu F^{\mu\nu}=0.
\]
It is nonlinear because both \(D_\mu\) and \(F_{\mu\nu}\) depend on the gauge
potential \(A_\mu\).
\end{theorembox}

This equation is the non-Abelian analogue of the source-free Maxwell equation
\[
    \partial_\mu F^{\mu\nu}=0.
\]
The ordinary derivative has been replaced by the gauge covariant derivative.

\section{Bianchi Identity}

The Yang--Mills field strength satisfies the covariant Bianchi identity
\[
    D_{[\lambda}F_{\mu\nu]}=0.
\]
Equivalently,
\[
    D_\lambda F_{\mu\nu}
    +D_\mu F_{\nu\lambda}
    +D_\nu F_{\lambda\mu}=0.
\]
In differential-form notation, this is
\[
    DF=0,
\]
where \(D\) is the exterior covariant derivative.

This identity follows from the definition of curvature and the Jacobi identity
for covariant derivatives. Since
\[
    [D_\mu,D_\nu]=gF_{\mu\nu},
\]
the Jacobi identity
\[
    [D_\lambda,[D_\mu,D_\nu]]
    +[D_\mu,[D_\nu,D_\lambda]]
    +[D_\nu,[D_\lambda,D_\mu]]=0
\]
leads directly to the Bianchi identity.

\begin{remarkbox}
The Bianchi identity is not an equation of motion. It holds identically for any
connection because the field strength is a curvature. The Yang--Mills equation
comes from varying the action; the Bianchi identity comes from geometry.
\end{remarkbox}

For an Abelian gauge field, \(D\) reduces to \(d\), and the Bianchi identity
becomes
\[
    dF=0,
\]
which is the homogeneous Maxwell equation.

\section{Self-Interaction of Gauge Fields}

The most important physical difference between Abelian and non-Abelian gauge
fields is self-interaction. In Maxwell theory, the field strength is linear in
\(A_\mu\). Therefore the free Maxwell action is quadratic in \(A_\mu\), and the
vacuum equations are linear.

In Yang--Mills theory,
\[
    F_{\mu\nu}=\partial_\mu A_\nu-\partial_\nu A_\mu+g[A_\mu,A_\nu].
\]
The commutator term makes the field strength nonlinear in the gauge potential.
Consequently, the action contains interaction terms even without matter.
Schematically,
\[
    \mathcal{L}_{\mathrm{YM}}
    \sim (\partial A)^2+g(\partial A)A^2+g^2A^4.
\]
The cubic and quartic terms describe gauge-field self-interactions.

\begin{examplebox}
Gluons in quantum chromodynamics are quanta of an \(SU(3)\) Yang--Mills field.
Unlike photons in Abelian electromagnetism, gluons carry non-Abelian color
charge. This is the quantum version of the classical fact that Yang--Mills gauge
fields interact with themselves.
\end{examplebox}

Classically, self-interaction means that Yang--Mills waves do not simply pass
through one another by linear superposition. The field equations are nonlinear,
and the gauge field acts as a source for itself.

\section{Energy-Momentum Tensor of Yang--Mills Fields}

The Yang--Mills stress-energy tensor is the natural non-Abelian generalization
of the electromagnetic one:
\[
    T^{\mu\nu}
    =-\langle F^{\mu\rho},F^\nu{}_{\rho}\rangle
     +\frac{1}{4}\eta^{\mu\nu}\langle F_{\rho\sigma},F^{\rho\sigma}\rangle.
\]
In component notation with \(\langle T_a,T_b\rangle=\delta_{ab}\),
\[
    T^{\mu\nu}
    =-F^{a\mu\rho}F^{a\nu}{}_{\rho}
     +\frac{1}{4}\eta^{\mu\nu}F_{\rho\sigma}^aF^{a\rho\sigma}.
\]
In vacuum, on shell,
\[
    \partial_\mu T^{\mu\nu}=0.
\]
This is ordinary spacetime energy-momentum conservation, not a gauge-covariant
conservation law, because \(T^{\mu\nu}\) is gauge invariant and has no free Lie
algebra index.

The energy density is positive for compact gauge groups with positive invariant
inner product. In analogy with electromagnetism, one may decompose the field
strength into non-Abelian electric and magnetic fields,
\[
    E_i^a=F_{0i}^a,
\]
and
\[
    B_i^a=\frac{1}{2}\epsilon_{ijk}F_{jk}^a
\]
up to sign conventions. Then the energy density has the schematic form
\[
    \mathcal{E}=\frac{1}{2}\left(E_i^aE_i^a+B_i^aB_i^a\right).
\]

\begin{definitionbox}
The Yang--Mills stress-energy tensor is gauge invariant and measures the energy,
momentum, and stresses carried by the non-Abelian gauge field. It is quadratic in
\(F_{\mu\nu}\), but because \(F_{\mu\nu}\) itself is nonlinear in \(A_\mu\), it
contains nonlinear gauge-field interactions.
\end{definitionbox}

This tensor is the source that a Yang--Mills field would contribute to the
Einstein equations if coupled to gravity.

\section{Classical Meaning of Non-Abelian Charge}

In electromagnetism, charge is a single conserved number associated with a
global \(U(1)\) symmetry. In a non-Abelian gauge theory, charge takes values in a
Lie algebra or in a representation of the gauge group. It has components,
\[
    Q=Q^aT_a.
\]
These components do not represent independent Abelian charges. They rotate into
one another under gauge transformations.

Matter currents in Yang--Mills theory are Lie-algebra-valued:
\[
    J^\mu=J^{a\mu}T_a.
\]
Their conservation law is covariant:
\[
    D_\mu J^\mu=0.
\]
This means that non-Abelian charge is transported with parallel transport in the
internal gauge space. Ordinary derivatives are not sufficient because the
internal frame itself can vary from point to point.

\begin{remarkbox}
A non-Abelian charge is not just a list of separately conserved numbers. It is a
geometric object in internal space whose components depend on the choice of gauge
frame. Gauge-covariant conservation replaces ordinary componentwise conservation.
\end{remarkbox}

This is the classical seed of a major quantum distinction. In non-Abelian gauge
theories, the gauge bosons themselves carry the charge to which they couple.
This is why the gauge field participates directly in the dynamics of its own
charge distribution.

\section{Comparison: Abelian vs Non-Abelian Theory}

The difference between Maxwell theory and Yang--Mills theory can be summarized
as follows.

\begin{center}
\begin{tabular}{lll}
\toprule
Feature & Abelian \(U(1)\) & Non-Abelian Yang--Mills \\
\midrule
Gauge group & commutative & noncommutative \\
Potential & real-valued \(A_\mu\) & Lie-algebra-valued \(A_\mu^aT_a\) \\
Field strength & \(\partial_\mu A_\nu-\partial_\nu A_\mu\) & \(\partial_\mu A_\nu-\partial_\nu A_\mu+g[A_\mu,A_\nu]\) \\
Gauge transformation of \(F\) & invariant & covariant \\
Equation of motion & \(\partial_\mu F^{\mu\nu}=J^\nu\) & \(D_\mu F^{\mu\nu}=J^\nu\) \\
Gauge-field self-interaction & absent & present \\
Charge conservation & \(\partial_\mu J^\mu=0\) & \(D_\mu J^\mu=0\) \\
\bottomrule
\end{tabular}
\end{center}

The Abelian theory is linear in the absence of sources. The non-Abelian theory
is nonlinear even in vacuum. This single difference has enormous consequences:
self-interactions, nontrivial classical solutions, confinement in quantum
chromodynamics, and a much richer gauge structure.

\begin{summarybox}
Yang--Mills theory generalizes electromagnetism by replacing the Abelian
\(U(1)\) gauge group with a non-Abelian Lie group \(G\). The gauge potential is a
Lie-algebra-valued connection, the field strength is
\(F=dA+gA\wedge A\), and the action is
\(-\frac{1}{4}\int F_{\mu\nu}^aF^{a\mu\nu}\,d^{d+1}x\). The field equations are
\(D_\mu F^{\mu\nu}=J^\nu\), and the non-Abelian commutator term makes the gauge
field self-interacting.
\end{summarybox}
