\chapter{Symmetries and Noether's Theorems}

\begin{remarkbox}
Symmetry is not only a visual property of a solution. In field theory, symmetry
is a structural property of the action. Noether's theorems explain why this
matters: continuous global symmetries lead to conserved currents, while local
symmetries reveal redundancies and identities among the field equations.
\end{remarkbox}

\section{Continuous Symmetries in Field Theory}

A symmetry is a transformation that preserves the physical content of a theory.
In the action formulation, the most useful definition is that a transformation
is a symmetry if it leaves the action invariant, possibly up to a boundary term.
For fields \(\phi^A(x)\), a continuous transformation can be written as
\[
    \phi^A(x) \longrightarrow \phi'^A(x)
    = \phi^A(x)+\varepsilon\Delta\phi^A(x)+O(\varepsilon^2),
\]
where \(\varepsilon\) is a small parameter and \(\Delta\phi^A\) is the
infinitesimal change of the field.

The transformation is continuous because it is connected to the identity: when
\(\varepsilon=0\), nothing changes. This is the situation covered by Noether's
first theorem. Discrete symmetries, such as parity or charge conjugation, are
important, but they do not directly produce conserved currents in the same way.

For a local Lagrangian density
\[
    \mathcal{L}=\mathcal{L}(\phi^A,\partial_\mu\phi^A,x),
\]
the action is
\[
    S[\phi]=\int_\Omega \mathcal{L}\,d^{d+1}x.
\]
A continuous symmetry is present if the transformation changes the Lagrangian
density by at most a total derivative:
\[
    \delta\mathcal{L}=\partial_\mu K^\mu.
\]
Then the action changes only by a boundary contribution,
\[
    \delta S=\int_{\partial\Omega}K^\mu n_\mu\,d\Sigma.
\]
Under suitable boundary conditions, this is zero.

\begin{definitionbox}
A \emph{continuous variational symmetry} is an infinitesimal transformation of
fields and possibly coordinates for which the Lagrangian density changes by a
total derivative:
\[
    \delta\mathcal{L}=\partial_\mu K^\mu.
\]
\end{definitionbox}

The phrase ``up to a total derivative'' is essential. Many physically important
symmetries leave the action invariant without leaving the Lagrangian density
strictly unchanged. The boundary term may be irrelevant for local equations of
motion, but it can be important for charges and boundary physics.

\section{Internal and Spacetime Symmetries}

Symmetries in field theory come in two broad types. Internal symmetries act on
the field values without moving spacetime points. Spacetime symmetries act on
the coordinates and therefore also on the arguments of the fields.

An internal symmetry has the schematic form
\[
    \phi^A(x)\longrightarrow \phi'^A(x)=R^A{}_B\phi^B(x),
\]
where \(R^A{}_B\) acts on field components. For example, a complex scalar field
may have a global \(U(1)\) symmetry
\[
    \phi(x)\longrightarrow e^{i\alpha}\phi(x).
\]
The spacetime point \(x\) is not changed. Only the internal phase of the field
is changed.

A spacetime symmetry changes the coordinates:
\[
    x^\mu\longrightarrow x'^\mu=x^\mu+\varepsilon\xi^\mu(x).
\]
Translations, rotations, Lorentz boosts, and scale transformations are examples.
For a scalar field, the transformed field satisfies
\[
    \phi'(x')=\phi(x).
\]
If one compares fields at the same coordinate value \(x\), the infinitesimal
change contains the derivative of the field:
\[
    \delta\phi(x)=-\varepsilon\xi^\mu\partial_\mu\phi(x).
\]

\begin{examplebox}
A spacetime translation by a constant vector \(a^\mu\) sends
\[
    x^\mu\longrightarrow x'^\mu=x^\mu+a^\mu.
\]
For a scalar field, this corresponds to the infinitesimal field variation
\[
    \delta\phi=-a^\mu\partial_\mu\phi.
\]
The associated conserved quantities are energy and momentum.
\end{examplebox}

Internal and spacetime symmetries may coexist. The Standard Model, for example,
uses spacetime Poincare symmetry together with internal gauge symmetries. Even
in purely classical field theory, the distinction is crucial: internal
symmetries usually lead to charges such as electric charge, while spacetime
symmetries lead to energy, momentum, and angular momentum.

\section{Noether's First Theorem}

Noether's first theorem states that every continuous global variational symmetry
of the action gives a conserved current. We now derive the basic formula for
fields \(\phi^A\) with a first-order Lagrangian density
\[
    \mathcal{L}=\mathcal{L}(\phi^A,\partial_\mu\phi^A,x).
\]

The variation of the Lagrangian density is
\[
    \delta\mathcal{L}
    =\frac{\partial\mathcal{L}}{\partial\phi^A}\delta\phi^A
    +\frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}
      \partial_\mu\delta\phi^A.
\]
Integrating the second term by parts at the density level gives
\[
\begin{aligned}
    \delta\mathcal{L}
    ={}&\left[
    \frac{\partial\mathcal{L}}{\partial\phi^A}
    -\partial_\mu\left(
        \frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}
    \right)\right]\delta\phi^A \\
    &+\partial_\mu\left(
        \frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}
        \delta\phi^A
    \right).
\end{aligned}
\]
The expression in square brackets is the Euler--Lagrange derivative. Denote it
by
\[
    E_A(\mathcal{L})
    =\frac{\partial\mathcal{L}}{\partial\phi^A}
    -\partial_\mu\left(
        \frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}
    \right).
\]
Then
\[
    \delta\mathcal{L}
    =E_A(\mathcal{L})\delta\phi^A
    +\partial_\mu\Theta^\mu(\phi,\delta\phi),
\]
where
\[
    \Theta^\mu(\phi,\delta\phi)
    =\frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}\delta\phi^A
\]
is the symplectic-potential current density for this simple first-order theory.

If the transformation is a symmetry, then
\[
    \delta\mathcal{L}=\partial_\mu K^\mu.
\]
Therefore
\[
    E_A(\mathcal{L})\delta\phi^A
    +\partial_\mu\Theta^\mu
    =\partial_\mu K^\mu.
\]
Rearranging gives
\[
    \partial_\mu J^\mu=-E_A(\mathcal{L})\delta\phi^A,
\]
where
\[
    J^\mu=\Theta^\mu-K^\mu.
\]
On shell, \(E_A(\mathcal{L})=0\), so
\[
    \partial_\mu J^\mu=0.
\]

\begin{theorembox}
\textbf{Noether's first theorem.} For every continuous global variational
symmetry of the action, there exists a current \(J^\mu\) whose divergence
vanishes on shell:
\[
    \partial_\mu J^\mu=0.
\]
The associated charge is conserved under suitable boundary conditions.
\end{theorembox}

This theorem is one of the central structural results of theoretical physics.
It turns symmetry into conservation. Time translation symmetry gives energy
conservation. Space translation symmetry gives momentum conservation. Rotational
symmetry gives angular momentum conservation. Internal phase symmetry gives
charge conservation.

\section{Conserved Currents and Conserved Charges}

A conserved current is a vector field \(J^\mu\) satisfying
\[
    \partial_\mu J^\mu=0.
\]
Writing \(J^\mu=(\rho,\vect{J})\), this equation becomes
\[
    \partial_t\rho+\nabla\cdot\vect{J}=0.
\]
This is a local conservation law. It says that the density \(\rho\) changes in a
region only because the current \(\vect{J}\) flows through the boundary of that
region.

The associated charge on a constant-time slice \(\Sigma_t\) is
\[
    Q(t)=\int_{\Sigma_t}J^0(t,\vect{x})\,d^d x.
\]
Its time derivative is
\[
    \frac{dQ}{dt}=\int_{\Sigma_t}\partial_tJ^0\,d^d x
    =-\int_{\Sigma_t}\partial_iJ^i\,d^d x.
\]
Using the divergence theorem,
\[
    \frac{dQ}{dt}=-\int_{\partial\Sigma_t}J^i n_i\,d\Sigma.
\]
Thus the charge is conserved if the flux through the boundary vanishes:
\[
    \frac{dQ}{dt}=0.
\]

\begin{definitionbox}
A \emph{conserved charge} is the spatial integral of the time component of a
conserved current:
\[
    Q=\int_{\Sigma_t}J^0\,d^d x.
\]
It is time-independent if the current has no net flux through the boundary of
space.
\end{definitionbox}

The distinction between local and global conservation is important. The
equation \(\partial_\mu J^\mu=0\) is local. It holds point by point. The
conservation of \(Q\) is global and depends on boundary conditions. A locally
conserved current may still produce a time-dependent charge in a finite region
if current flows across the boundary.

\section{Energy, Momentum, and Angular Momentum}

Spacetime translations lead to energy and momentum. If the Lagrangian density
has no explicit dependence on spacetime coordinates, then it is invariant under
translations
\[
    x^\mu\longrightarrow x^\mu+a^\mu.
\]
The corresponding conserved object is the stress-energy tensor
\(T^\mu{}_{\nu}\), satisfying
\[
    \partial_\mu T^\mu{}_{\nu}=0
\]
on shell.

The conserved four-momentum is
\[
    P_\nu=\int_{\Sigma_t}T^0{}_{\nu}\,d^d x.
\]
The component associated with time translations is the energy. The components
associated with spatial translations are the momenta.

Rotations and Lorentz boosts lead to angular momentum and boost charges. The
corresponding current has the schematic form
\[
    M^{\mu\rho\sigma}
    =x^\rho T^{\mu\sigma}-x^\sigma T^{\mu\rho}+S^{\mu\rho\sigma},
\]
where \(S^{\mu\rho\sigma}\) is a possible spin contribution. For scalar fields,
the spin term is absent. For vector and spinor fields, it is generally present.

The conservation law is
\[
    \partial_\mu M^{\mu\rho\sigma}=0.
\]
The associated charges are
\[
    J^{\rho\sigma}=\int_{\Sigma_t}M^{0\rho\sigma}\,d^d x.
\]
Spatial components describe ordinary angular momentum; mixed time-space
components describe Lorentz boost generators.

\begin{remarkbox}
Energy, momentum, and angular momentum are not separate miracles. They are the
Noether charges of spacetime symmetries: time translations, space translations,
rotations, and Lorentz boosts.
\end{remarkbox}

This perspective is more powerful than memorizing conserved quantities one by
one. Once the action and the symmetry are known, Noether's theorem tells us what
kind of conservation law to expect.

\section{Stress-Energy Tensor: Canonical Form}

For a collection of fields \(\phi^A\) with Lagrangian density
\(\mathcal{L}(\phi^A,\partial_\mu\phi^A)\) and no explicit coordinate dependence,
the canonical stress-energy tensor is
\[
    T^\mu{}_{\nu}
    =\frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}
      \partial_\nu\phi^A
    -\delta^\mu{}_{\nu}\mathcal{L}.
\]
With the sign conventions used here, this tensor satisfies
\[
    \partial_\mu T^\mu{}_{\nu}=0
\]
on shell.

Let us check the structure. The first term is built from the momentum conjugate
to \(\partial_\mu\phi^A\) and the derivative \(\partial_\nu\phi^A\). The second
term subtracts the Lagrangian density when \(\mu=\nu\). This is the direct field
theory analogue of the mechanical relation between energy, momenta, velocities,
and the Lagrangian.

For a real scalar field
\[
    \mathcal{L}=\frac{1}{2}\partial_\rho\phi\partial^\rho\phi-V(\phi),
\]
one has
\[
    \frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi)}=\partial^\mu\phi.
\]
Therefore
\[
    T^\mu{}_{\nu}=\partial^\mu\phi\,\partial_\nu\phi
    -\delta^\mu{}_{\nu}\mathcal{L}.
\]
With both indices raised, this is commonly written as
\[
    T^{\mu\nu}=\partial^\mu\phi\,\partial^\nu\phi-\eta^{\mu\nu}\mathcal{L}.
\]

\begin{examplebox}
For the real scalar field, the energy density is
\[
    T^{00}=\frac{1}{2}(\partial_t\phi)^2
    +\frac{1}{2}|\nabla\phi|^2+V(\phi).
\]
This is positive for potentials bounded below, as expected for a stable scalar
field theory.
\end{examplebox}

The canonical stress-energy tensor is not always symmetric, and it is not always
gauge invariant. This is not a contradiction. It means that the canonical tensor
is not always the final or most physical stress-energy tensor.

\section{Improvement Terms}

A conserved current is not unique. If \(J^\mu\) is conserved, then
\[
    J'^\mu=J^\mu+\partial_\nu B^{[\mu\nu]}
\]
is also conserved, because
\[
    \partial_\mu\partial_\nu B^{[\mu\nu]}=0
\]
for antisymmetric \(B^{[\mu\nu]}=-B^{[\nu\mu]}\). The added term is called an
improvement term.

The same phenomenon occurs for the stress-energy tensor. One may add a total
divergence that does not change the local conservation law but changes the
local expression for the tensor:
\[
    T'^{\mu\nu}=T^{\mu\nu}+\partial_\rho B^{\rho\mu\nu},
\]
with suitable antisymmetry in the indices of \(B\). Under appropriate boundary
conditions, the integrated charges remain the same.

Improvement terms are not merely cosmetic. They can make the stress-energy
tensor symmetric, gauge invariant, or better adapted to coupling with gravity.
For example, the Belinfante--Rosenfeld procedure combines the canonical
stress-energy tensor with spin-current terms to produce a symmetric tensor.

\begin{remarkbox}
Noether currents are defined up to improvement terms. Therefore one should not
attach too much physical meaning to a particular local representative of a
current without specifying boundary conditions and the class of allowed
improvements.
\end{remarkbox}

This ambiguity is a preview of a deeper issue: conserved charges are often more
robust than local current representatives. In gauge theory and gravity, the
boundary part of a current can become the most important part of the charge.

\section{Noether's Second Theorem and Gauge Symmetries}

Noether's first theorem concerns global symmetries depending on finitely many
constant parameters. Noether's second theorem concerns local symmetries depending
on arbitrary functions. Gauge symmetries are the main example.

A simple gauge transformation in electromagnetism is
\[
    A_\mu\longrightarrow A_\mu+\partial_\mu\lambda,
\]
where \(\lambda(x)\) is an arbitrary function. The field strength
\[
    F_{\mu\nu}=\partial_\mu A_\nu-\partial_\nu A_\mu
\]
is unchanged:
\[
    \delta F_{\mu\nu}=0.
\]
Therefore the Maxwell action is gauge invariant.

The key point is that \(\lambda(x)\) is arbitrary. A gauge symmetry is not an
ordinary physical transformation from one distinct state to another. It is a
redundancy in the description. Two gauge-related configurations represent the
same physical field configuration.

Noether's second theorem says that local gauge symmetries imply identities among
the Euler--Lagrange equations. Schematically, if
\[
    \delta_\lambda\phi^A=R^A(\lambda,\partial\lambda,\ldots)
\]
is a gauge symmetry with arbitrary \(\lambda(x)\), then the field equations
cannot be independent. There must be a differential identity among them.

In electromagnetism, the field equation is
\[
    \partial_\mu F^{\mu\nu}=J^\nu.
\]
Taking \(\partial_\nu\) of both sides gives
\[
    \partial_\nu\partial_\mu F^{\mu\nu}=\partial_\nu J^\nu.
\]
The left-hand side vanishes identically because \(F^{\mu\nu}\) is antisymmetric
and partial derivatives commute. Thus
\[
    \partial_\nu J^\nu=0.
\]
This expresses charge conservation as a consistency condition of the gauge field
equation.

\begin{theorembox}
\textbf{Noether's second theorem.} Local gauge symmetries imply differential
identities among the Euler--Lagrange equations. These identities reflect the
redundancy of the description and are not ordinary conservation laws of the
same type as those produced by global symmetries.
\end{theorembox}

This distinction is essential. A global symmetry usually produces a physical
charge. A gauge symmetry usually signals that some variables are redundant and
that constraints are present.

\section{Global vs Local Symmetries}

A global symmetry depends on constant parameters. For example,
\[
    \phi\longrightarrow e^{i\alpha}\phi
\]
with constant \(\alpha\) is a global \(U(1)\) symmetry. Noether's first theorem
then gives a conserved current and a conserved charge.

A local symmetry depends on arbitrary functions. For example,
\[
    \phi(x)\longrightarrow e^{i\alpha(x)}\phi(x)
\]
with arbitrary \(\alpha(x)\) is a local phase transformation. To make such a
symmetry compatible with derivatives, one introduces a gauge field and replaces
ordinary derivatives by covariant derivatives:
\[
    \partial_\mu\phi\longrightarrow D_\mu\phi.
\]

The relation between global and local symmetries is subtle. A global symmetry is
often associated with a conserved charge. A local symmetry is usually a
redundancy. However, boundary conditions can convert part of a gauge symmetry
into a genuine physical transformation at the boundary. This is why gauge
charges and asymptotic symmetries play such an important role in modern field
theory and gravity.

\begin{definitionbox}
A \emph{global symmetry} has parameters that are constant over spacetime. A
\emph{local symmetry} has parameters that are arbitrary functions of spacetime.
Global symmetries usually generate conserved charges; local symmetries usually
express redundancies and constraints.
\end{definitionbox}

The safest rule is this: global symmetries act on physical states, while gauge
symmetries identify different descriptions of the same physical state. Boundary
conditions and charges can complicate this rule, but it is the correct starting
point.

\section{Physical Meaning of Conservation Laws}

A conservation law is not merely the statement that a number does not change.
The local equation
\[
    \partial_\mu J^\mu=0
\]
expresses a balance between density and flux. In nonrelativistic notation,
\[
    \partial_t\rho+\nabla\cdot\vect{J}=0.
\]
This says that charge cannot disappear at a point; it can only flow from one
place to another.

Energy conservation means that energy can move, transform between forms, or flow
through boundaries, but it is not locally created or destroyed if the relevant
symmetry assumptions hold. Momentum conservation expresses spatial homogeneity.
Angular momentum conservation expresses rotational symmetry. Electric charge
conservation expresses global phase symmetry.

The assumptions matter. If a Lagrangian has explicit time dependence, energy
need not be conserved. If space translation symmetry is broken by an external
background, momentum need not be conserved. If charge flows through the boundary
of the region under study, the charge inside that region need not be constant.

\begin{remarkbox}
Noether's theorem does not say that every quantity is conserved. It says that a
specific conservation law follows from a specific continuous symmetry of the
action, under specified boundary conditions.
\end{remarkbox}

This is why the action formulation is so economical. Instead of deriving every
conservation law separately from the equations of motion, one studies the
symmetries of the action and lets Noether's theorem organize the results.

\section{Examples and Common Misinterpretations}

\subsection*{Complex scalar field and phase symmetry}

For a complex scalar field with Lagrangian density
\[
    \mathcal{L}=\partial_\mu\phi^*\partial^\mu\phi-m^2\phi^*\phi,
\]
there is a global \(U(1)\) symmetry
\[
    \phi\longrightarrow e^{i\alpha}\phi,
    \qquad
    \phi^*\longrightarrow e^{-i\alpha}\phi^*.
\]
The infinitesimal variations are
\[
    \delta\phi=i\alpha\phi,
    \qquad
    \delta\phi^*=-i\alpha\phi^*.
\]
The associated Noether current is, up to the conventional normalization,
\[
    J^\mu=i(\phi\partial^\mu\phi^*-\phi^*\partial^\mu\phi).
\]
On shell,
\[
    \partial_\mu J^\mu=0.
\]
This is the classical origin of charge conservation for a charged scalar field.

\subsection*{Misinterpretation: every symmetry gives a new particle}

Noether's theorem says that a continuous global symmetry gives a conserved
current. It does not say that every symmetry directly gives a new particle. The
relationship between symmetries and particles belongs to quantum field theory
and depends on representation theory, spontaneous symmetry breaking, and the
spectrum of the theory.

\subsection*{Misinterpretation: gauge symmetry is an ordinary physical motion}

Gauge symmetry is primarily a redundancy. Gauge-related configurations describe
the same physical state. This is why gauge fixing is allowed and why gauge
symmetries lead to constraints. Nevertheless, transformations that do not vanish
at a boundary can generate physical charges. This is a boundary-sensitive
refinement, not a contradiction.

\subsection*{Misinterpretation: total derivatives never matter}

Total derivatives do not change the bulk Euler--Lagrange equations under
suitable assumptions. But they can change boundary terms, canonical structures,
and charges. Therefore total derivatives are harmless only after the boundary
problem has been specified.

\begin{summarybox}
Noether's first theorem turns continuous global symmetries into conserved
currents and charges. Spacetime translations give energy and momentum; rotations
and Lorentz transformations give angular momentum and boost charges; internal
symmetries give internal charges. Noether's second theorem explains why local
gauge symmetries imply identities and constraints rather than ordinary physical
motions. Conservation laws are local balance equations plus boundary conditions.
\end{summarybox}
