\chapter{Mathematical Preliminaries}

\begin{remarkbox}
This chapter collects the mathematical tools needed before the action principle
for fields can be used cleanly. The goal is not to give a complete course in
geometry or analysis, but to fix the language that will be used throughout the
notes: fields, indices, derivatives, integrals, distributions, variations,
functional derivatives, boundary terms, and Lorentzian notation.
\end{remarkbox}

\section{Scalar, Vector, Tensor, and Spinor Fields}

A classical field is a quantity assigned to points of space or spacetime. The
simplest example is a real scalar field
\[
    \phi : M \to \mathbb{R},
\]
where \(M\) is the space or spacetime on which the theory is defined. A scalar
field has one value at each point and does not carry a direction or an index.
In ordinary language, a temperature distribution \(T(t,\vect{x})\) is a scalar
field. In relativistic field theory, the Klein--Gordon field \(\phi(x)\) is the
basic example.

A vector field assigns a vector to each point. In three-dimensional space one
may write
\[
    \vect{v}(t,\vect{x}) = v^1(t,\vect{x})\vect{e}_1
    + v^2(t,\vect{x})\vect{e}_2
    + v^3(t,\vect{x})\vect{e}_3.
\]
The velocity field of a fluid and the electric field \(\vect{E}(t,\vect{x})\)
are familiar examples. In relativistic notation, vector fields are written with
spacetime indices, for example \(A_\mu(x)\) or \(V^\mu(x)\).

Tensor fields assign tensors to each point. A rank-two tensor field may be
written as \(T^{\mu\nu}(x)\), \(T_{\mu\nu}(x)\), or \(T^\mu{}_{\nu}(x)\),
depending on the position of the indices. The electromagnetic field strength
\(F_{\mu\nu}\), the metric tensor \(g_{\mu\nu}\), and the stress-energy tensor
\(T^{\mu\nu}\) are all tensor fields.

Spinor fields are more subtle. They are not ordinary vectors, although they
also have components. Spinors are objects whose transformation law is tied to
the spin representation of the Lorentz group. The Dirac field \(\psi(x)\) is the
main example. In this course, spinors will be introduced only after the scalar
and gauge-field examples are understood.

\begin{definitionbox}
A \emph{field type} specifies what kind of object is assigned to each point:
a scalar, vector, tensor, spinor, or a collection of such objects. The field
equations must be compatible with the transformation properties of that object.
\end{definitionbox}

The notation \(\phi^A(x)\) is often useful. The label \(A\) may represent one
component, many tensor indices, an internal symmetry index, or a spinor index.
With this convention, many formulas can be written once and then applied to a
large class of fields.

\section{Coordinates, Indices, and Summation Convention}

Coordinates label points. In nonrelativistic field theory one often separates
time and space,
\[
    (t,\vect{x}) = (t,x^1,x^2,x^3).
\]
In relativistic field theory, it is convenient to combine them into spacetime
coordinates
\[
    x^\mu = (x^0,x^1,x^2,x^3),
\]
where \(x^0\) denotes the time coordinate, often \(ct\) or simply \(t\) in units
where \(c=1\). Greek indices such as \(\mu,\nu,\rho\) usually run over spacetime
values \(0,1,2,3\). Latin spatial indices such as \(i,j,k\) usually run over
\(1,2,3\).

The Einstein summation convention will be used throughout the notes: whenever
an index appears once upstairs and once downstairs in the same term, summation
over that index is understood. Thus
\[
    A_\mu B^\mu
\]
means
\[
    \sum_{\mu=0}^{3} A_\mu B^\mu.
\]
Similarly,
\[
    T^\mu{}_{\nu}V^\nu
\]
means that the index \(\nu\) is summed over, while \(\mu\) remains a free index.

\begin{definitionbox}
An index is \emph{free} if it appears in the final expression and is not summed
over. An index is \emph{dummy} if it is summed over. Dummy indices may be renamed;
free indices may not.
\end{definitionbox}

For example,
\[
    A_\mu B^\mu = A_\nu B^\nu
\]
because \(\mu\) and \(\nu\) are dummy indices. But
\[
    T^\mu{}_{\nu}V^\nu
\]
is not the same kind of object as
\[
    T^\rho{}_{\nu}V^\nu,
\]
unless one also changes the name of the free index consistently. The free index
labels the component of the resulting vector.

Indices are not decoration. They encode the tensorial type of a formula. A
valid tensor equation must have the same free indices on both sides. For
example,
\[
    \partial_\mu F^{\mu\nu} = J^\nu
\]
is consistent because both sides carry the free index \(\nu\). The expression
\(\partial_\mu F^{\mu\nu}=J_\mu\), by contrast, would not be a valid tensor
equation as written, because the free indices do not match.

\section{Partial Derivatives and Functional Dependence}

Fields depend on spacetime coordinates. For a scalar field \(\phi(x)\), the
partial derivative is written as
\[
    \partial_\mu \phi = \frac{\partial \phi}{\partial x^\mu}.
\]
The collection \(\partial_\mu\phi\) is the gradient of the scalar field in
spacetime notation. Higher derivatives are written as
\[
    \partial_\mu\partial_\nu \phi.
\]
When the coordinates are ordinary flat coordinates and the fields are smooth,
partial derivatives commute:
\[
    \partial_\mu\partial_\nu\phi = \partial_\nu\partial_\mu\phi.
\]

In a Lagrangian field theory, the Lagrangian density usually depends on the
fields and their first derivatives:
\[
    \mathcal{L}=\mathcal{L}(\phi^A,\partial_\mu\phi^A,x).
\]
The symbol \(x\) at the end allows explicit coordinate dependence. Many
important theories have no explicit coordinate dependence, in which case
spacetime translation symmetry leads to energy-momentum conservation.

It is important to distinguish ordinary dependence from functional dependence.
The expression \(\mathcal{L}(\phi,\partial_\mu\phi,x)\) is a function of the
local variables \(\phi(x)\), \(\partial_\mu\phi(x)\), and \(x\). The action
\[
    S[\phi] = \int \mathcal{L}(\phi,\partial_\mu\phi,x)\,d^{d+1}x
\]
is a functional: it takes an entire field configuration \(\phi\) and returns a
number.

\begin{examplebox}
For a real scalar field in flat spacetime, a standard Lagrangian density is
\[
    \mathcal{L}
    = \frac{1}{2}\partial_\mu\phi\,\partial^\mu\phi
      - \frac{1}{2}m^2\phi^2.
\]
This expression is local: at each spacetime point it depends only on \(\phi\) and
its first derivatives at that same point.
\end{examplebox}

\section{Differential Forms: A Minimal Introduction}

Differential forms provide a compact language for integration, Stokes' theorem,
and gauge fields. A zero-form is just a function. A one-form is an object that
can be written locally as
\[
    \alpha = \alpha_\mu dx^\mu.
\]
A two-form has the form
\[
    \omega = \frac{1}{2}\omega_{\mu\nu}dx^\mu\wedge dx^\nu,
\]
where \(\wedge\) is the wedge product. The wedge product is antisymmetric:
\[
    dx^\mu\wedge dx^\nu = -dx^\nu\wedge dx^\mu.
\]
This antisymmetry is why the components of a two-form satisfy
\(\omega_{\mu\nu}=-\omega_{\nu\mu}\).

The exterior derivative maps \(p\)-forms to \((p+1)\)-forms. For a function
\(f\),
\[
    df = \partial_\mu f\,dx^\mu.
\]
For a one-form \(A=A_\mu dx^\mu\),
\[
    dA = \frac{1}{2}(\partial_\mu A_\nu-\partial_\nu A_\mu)
    dx^\mu\wedge dx^\nu.
\]
This is the natural form-language version of an antisymmetric field strength.
In electromagnetism one writes
\[
    F=dA,
\]
where \(A\) is the electromagnetic potential one-form and \(F\) is the field
strength two-form.

A crucial identity is
\[
    d^2=0.
\]
Applied to electromagnetism, this gives
\[
    dF = d(dA)=0,
\]
which is the geometric form of the homogeneous Maxwell equations.

\begin{remarkbox}
Differential forms are not introduced here for elegance alone. They make the
boundary structure of field theory visible. The formula \(d^2=0\), Stokes'
theorem, and the compact expression \(F=dA\) are the natural language of gauge
fields.
\end{remarkbox}

\section{Integration on Space and Spacetime}

A field theory action is usually an integral over spacetime. In \(d\) spatial
dimensions, the spacetime integration measure is written as
\[
    d^{d+1}x = dx^0\,dx^1\cdots dx^d.
\]
For a relativistic theory in four spacetime dimensions, one often writes
\[
    S[\phi] = \int \mathcal{L}\,d^4x.
\]
If one separates time and space, the same integral is written as
\[
    S[\phi] = \int dt\int d^d x\,\mathcal{L}.
\]

The Lagrangian density \(\mathcal{L}\) should not be confused with the
Lagrangian \(L\). The relation is
\[
    L(t) = \int d^d x\,\mathcal{L}(t,\vect{x}).
\]
Thus the action may be written either as
\[
    S = \int L(t)\,dt
\]
or as
\[
    S = \int \mathcal{L}\,dt\,d^d x.
\]

When the spacetime region has a boundary, one must keep track of boundary
integrals. In ordinary vector calculus, the divergence theorem says that
\[
    \int_\Omega \nabla\cdot\vect{J}\,d^d x
    = \int_{\partial\Omega} \vect{J}\cdot\vect{n}\,d\Sigma.
\]
In spacetime notation, an analogous statement is
\[
    \int_\Omega \partial_\mu J^\mu\,d^{d+1}x
    = \int_{\partial\Omega} J^\mu n_\mu\,d\Sigma.
\]
This is the reason total derivative terms are often said to become boundary
terms.

\section{Delta Functions and Distributions}

The Dirac delta is not an ordinary function. It is a distribution defined by
its action under an integral:
\[
    \int_{-\infty}^{\infty}\delta(x-a)f(x)\,dx = f(a).
\]
In \(d\) spatial dimensions,
\[
    \int \delta^{(d)}(\vect{x}-\vect{y})f(\vect{x})\,d^d x
    = f(\vect{y}).
\]
The delta distribution is used to represent point sources, Green's functions,
and functional derivatives.

Point charge density, for example, can be written as
\[
    \rho(\vect{x}) = q\,\delta^{(3)}(\vect{x}-\vect{x}_0),
\]
meaning that the total charge is localized at \(\vect{x}_0\). Indeed,
\[
    \int \rho(\vect{x})\,d^3x = q.
\]

Derivatives of delta distributions are defined by integration by parts. In one
dimension,
\[
    \int \delta'(x-a)f(x)\,dx = -f'(a),
\]
assuming boundary terms vanish. This rule is essential when manipulating
Green's functions and variational derivatives.

\begin{definitionbox}
A \emph{distribution} is an object defined by how it acts under an integral
against smooth test functions. The Dirac delta is the distribution that extracts
the value of a test function at a point.
\end{definitionbox}

\section{Variations of Fields}

A variation of a field is an infinitesimal change of the field configuration.
For a scalar field one writes
\[
    \phi(x) \longrightarrow \phi(x)+\varepsilon\eta(x),
\]
where \(\varepsilon\) is a small parameter and \(\eta(x)\) is an arbitrary test
variation. The variation of the field is
\[
    \delta\phi(x)=\eta(x).
\]
The variation of a derivative is the derivative of the variation:
\[
    \delta(\partial_\mu\phi)=\partial_\mu(\delta\phi).
\]
This simple identity is the mechanism behind the Euler--Lagrange equations for
fields.

For an action
\[
    S[\phi]=\int \mathcal{L}(\phi,\partial_\mu\phi,x)\,d^{d+1}x,
\]
the first variation is
\[
    \delta S
    = \int \left(
        \frac{\partial\mathcal{L}}{\partial\phi}\delta\phi
        + \frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi)}
          \delta(\partial_\mu\phi)
    \right)d^{d+1}x.
\]
Using \(\delta(\partial_\mu\phi)=\partial_\mu\delta\phi\), one obtains
\[
    \delta S
    = \int \left(
        \frac{\partial\mathcal{L}}{\partial\phi}\delta\phi
        + \frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi)}
          \partial_\mu\delta\phi
    \right)d^{d+1}x.
\]
The second term is then integrated by parts. This produces a bulk term and a
boundary term. The bulk term gives the field equation; the boundary term tells
us what must be assumed or added for the variational principle to be well
defined.

\section{Functional Derivatives}

A functional derivative measures how a functional changes when the value of a
field changes at a point. For a functional \(F[\phi]\), the functional derivative
is defined by
\[
    \delta F
    = \int \frac{\delta F}{\delta\phi(x)}\,\delta\phi(x)\,d^{d+1}x,
\]
up to boundary terms and under the chosen class of variations.

The simplest example is
\[
    F[\phi] = \int f(x)\phi(x)\,d^{d+1}x.
\]
Then
\[
    \delta F = \int f(x)\delta\phi(x)\,d^{d+1}x,
\]
so
\[
    \frac{\delta F}{\delta\phi(x)} = f(x).
\]
A second important identity is
\[
    \frac{\delta\phi(y)}{\delta\phi(x)} = \delta^{(d+1)}(y-x).
\]
This is the field-theoretic analogue of
\(\partial q^j/\partial q^i=\delta^j_i\).

For a local action of the form
\[
    S[\phi]=\int \mathcal{L}(\phi,\partial_\mu\phi,x)\,d^{d+1}x,
\]
the functional derivative is
\[
    \frac{\delta S}{\delta\phi}
    = \frac{\partial\mathcal{L}}{\partial\phi}
      - \partial_\mu\left(
        \frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi)}
      \right),
\]
provided the boundary term is handled appropriately. The Euler--Lagrange
equation is then simply
\[
    \frac{\delta S}{\delta\phi}=0.
\]
For several fields \(\phi^A\), one has one equation for each field component:
\[
    \frac{\delta S}{\delta\phi^A}=0.
\]

\section{Boundary Terms and the Meaning of ``Up to a Total Derivative''}

Boundary terms are not a technical nuisance. They are one of the main structural
features of field theory. Consider a total derivative in spacetime,
\[
    \partial_\mu K^\mu.
\]
Its integral over a spacetime region \(\Omega\) is
\[
    \int_\Omega \partial_\mu K^\mu\,d^{d+1}x
    = \int_{\partial\Omega} K^\mu n_\mu\,d\Sigma.
\]
Thus adding a total derivative to the Lagrangian density changes the action by a
boundary term:
\[
    \mathcal{L}' = \mathcal{L}+\partial_\mu K^\mu.
\]
The bulk Euler--Lagrange equations are unchanged, but the variational principle,
canonical momenta, conserved charges, and boundary conditions may change.

This is the precise meaning of the phrase ``up to a total derivative.'' It does
not mean that the term is zero. It means that the term does not affect the bulk
field equations under the assumed boundary conditions. If the boundary is
physically important, the term may become essential.

\begin{remarkbox}
A total derivative is harmless only after the boundary problem has been stated.
In theories with boundaries, asymptotic regions, gauge symmetry, gravity, or
conserved charges, boundary terms often contain physical information.
\end{remarkbox}

In the variational derivation, the characteristic integration by parts is
\[
    \int_\Omega P^\mu\partial_\mu\delta\phi\,d^{d+1}x
    = -\int_\Omega (\partial_\mu P^\mu)\delta\phi\,d^{d+1}x
      + \int_{\partial\Omega} P^\mu n_\mu\delta\phi\,d\Sigma.
\]
The first term contributes to the Euler--Lagrange equation. The second is the
boundary term. One may remove it by fixing \(\delta\phi=0\) on the boundary, by
choosing falloff conditions, or by adding compensating boundary terms to the
action.

\section{Lorentzian Geometry Needed for Field Theory}

Relativistic field theory is formulated on spacetime. The simplest spacetime is
Minkowski spacetime, equipped with the metric \(\eta_{\mu\nu}\). In these notes
we will mostly use the signature
\[
    \eta_{\mu\nu}=\operatorname{diag}(+1,-1,-1,-1),
\]
unless stated otherwise. With this convention,
\[
    ds^2 = \eta_{\mu\nu}dx^\mu dx^\nu
    = (dx^0)^2 - (dx^1)^2 - (dx^2)^2 - (dx^3)^2.
\]
The metric is used to raise and lower indices:
\[
    V_\mu = \eta_{\mu\nu}V^\nu,
    \qquad
    V^\mu = \eta^{\mu\nu}V_\nu.
\]
It is also used to form Lorentz-invariant contractions, such as
\[
    V_\mu V^\mu.
\]

The derivative with a raised index is defined by
\[
    \partial^\mu = \eta^{\mu\nu}\partial_\nu.
\]
The d'Alembert operator is
\[
    \Box = \partial_\mu\partial^\mu.
\]
With the above signature this becomes
\[
    \Box = \partial_t^2 - \nabla^2
\]
in units where \(c=1\). The Klein--Gordon equation may then be written compactly
as
\[
    (\Box + m^2)\phi=0.
\]

In curved spacetime, the metric becomes a field \(g_{\mu\nu}(x)\), partial
derivatives are replaced by covariant derivatives, and the volume element is
modified. For much of Classical Field Theory I, however, the main stage will be
flat Minkowski spacetime. Curved spacetime will appear later as a natural
extension and as a bridge toward gravity.

\begin{summarybox}
The mathematical transition from mechanics to field theory is the transition
from coordinates \(q^i(t)\) to fields \(\phi^A(x)\), from ordinary derivatives to
spacetime derivatives, from a Lagrangian \(L\) to a Lagrangian density
\(\mathcal{L}\), and from finite-dimensional derivatives to functional
derivatives. Boundary terms are part of the structure, not an afterthought.
\end{summarybox}
