\chapter{The Real Scalar Field}

\begin{remarkbox}
The real scalar field is the simplest relativistic field theory with local
degrees of freedom. It is simple enough to solve explicitly in the free case,
but rich enough to introduce the essential structures of classical field theory:
a Lorentz-invariant action, a wave equation, a dispersion relation, conserved
energy-momentum, Green's functions, sources, interactions, stability, and small
fluctuations around backgrounds.
\end{remarkbox}

\section{Klein--Gordon Field}

A real scalar field is a real-valued function on spacetime,
\[
    \phi : M \to \mathbb{R}.
\]
In flat Minkowski spacetime, with coordinates \(x^\mu=(t,\vect{x})\), the
standard free massive scalar-field action is
\[
    S[\phi]=\int d^{d+1}x\,\mathcal{L},
\]
where
\[
    \mathcal{L}
    =\frac{1}{2}\partial_\mu\phi\partial^\mu\phi
     -\frac{1}{2}m^2\phi^2.
\]
With the mostly-minus convention,
\[
    \partial_\mu\phi\partial^\mu\phi
    =(\partial_t\phi)^2-|\nabla\phi|^2.
\]
Thus the Lagrangian density can also be written as
\[
    \mathcal{L}
    =\frac{1}{2}(\partial_t\phi)^2
     -\frac{1}{2}|\nabla\phi|^2
     -\frac{1}{2}m^2\phi^2.
\]
The first term is the kinetic term, the second is the spatial-gradient term,
and the third is the mass term.

Varying the action gives
\[
    \delta S
    =\int d^{d+1}x\left(
        \partial_\mu\phi\partial^\mu\delta\phi
        -m^2\phi\delta\phi
    \right).
\]
Integrating the first term by parts and assuming that the boundary term
vanishes, we obtain
\[
    \delta S
    =-\int d^{d+1}x\left(\partial_\mu\partial^\mu\phi+m^2\phi\right)\delta\phi.
\]
The stationarity condition \(\delta S=0\) for arbitrary \(\delta\phi\) gives
\[
    (\Box+m^2)\phi=0,
\]
where
\[
    \Box=\partial_\mu\partial^\mu=\partial_t^2-\nabla^2.
\]
This is the Klein--Gordon equation.

\begin{definitionbox}
The \emph{Klein--Gordon field} is a real scalar field satisfying
\[
    (\Box+m^2)\phi=0.
\]
It is the relativistic field equation associated with the invariant mass
parameter \(m\).
\end{definitionbox}

The equation is linear. Therefore, if \(\phi_1\) and \(\phi_2\) are solutions,
then any linear combination
\[
    a\phi_1+b\phi_2
\]
is also a solution. This superposition property will fail once nonlinear
interactions are added.

\section{Massless and Massive Scalar Fields}

The parameter \(m\) controls the relation between frequency and wavelength, the
range of static response, and the interpretation of excitations. When
\(m=0\), the equation becomes
\[
    \Box\phi=0,
\]
the relativistic wave equation. Disturbances propagate at the speed of light in
flat spacetime.

When \(m\neq0\), the equation is
\[
    \partial_t^2\phi-\nabla^2\phi+m^2\phi=0.
\]
The mass term acts as a local restoring term: even a spatially homogeneous field
oscillates,
\[
    \ddot\phi+m^2\phi=0.
\]
Thus a massive scalar field combines wave propagation with local oscillation.

The mass parameter also defines a length scale,
\[
    \ell_m=\frac{1}{m}
\]
in units \(\hbar=c=1\). In classical static problems, this length scale controls
the decay of the response to localized sources. The larger the mass, the shorter
the range of the field.

\begin{remarkbox}
The word ``mass'' has a relativistic meaning: the plane-wave excitations of the
field satisfy the invariant relation \(p_\mu p^\mu=m^2\). Classically, the same
parameter also controls oscillation frequency and static screening length.
\end{remarkbox}

The massless case has additional infrared subtleties, especially in low spatial
dimensions or in the presence of long-range sources. The massive case is often
technically cleaner because the mass term suppresses long-distance response.

\section{Plane-Wave Solutions}

Because the free Klein--Gordon equation has constant coefficients, it is natural
to look for plane-wave solutions. Let
\[
    \phi(x)=A e^{-ik_\mu x^\mu},
\]
where
\[
    k_\mu x^\mu=\omega t-\vect{k}\cdot\vect{x}
\]
with our conventions. Acting with the d'Alembert operator gives
\[
    \Box e^{-ik_\mu x^\mu}=-k_\mu k^\mu e^{-ik_\mu x^\mu}.
\]
The Klein--Gordon equation therefore requires
\[
    -k_\mu k^\mu+m^2=0,
\]
or
\[
    k_\mu k^\mu=m^2.
\]
Since
\[
    k_\mu k^\mu=\omega^2-|\vect{k}|^2,
\]
we get
\[
    \omega^2=|\vect{k}|^2+m^2.
\]

A real scalar field cannot be a single complex exponential unless one takes the
real part. A general real solution can be written as a superposition of modes,
for example
\[
    \phi(t,\vect{x})
    =\int \frac{d^d k}{(2\pi)^d}\left[
        a(\vect{k})e^{-i\omega_{\vect{k}}t+i\vect{k}\cdot\vect{x}}
        +a^*(\vect{k})e^{i\omega_{\vect{k}}t-i\vect{k}\cdot\vect{x}}
    \right],
\]
where
\[
    \omega_{\vect{k}}=\sqrt{|\vect{k}|^2+m^2}.
\]
The complex conjugate term ensures that \(\phi\) is real.

\begin{examplebox}
In one spatial dimension, a real standing-wave solution of the massive equation
is
\[
    \phi(t,x)=A\cos(kx)\cos(\omega t),
    \qquad
    \omega^2=k^2+m^2.
\]
For \(m=0\), this reduces to the familiar wave relation \(\omega=|k|\).
\end{examplebox}

Plane waves are not localized. They are idealized basis solutions. Localized
wave packets are constructed by superposing many plane waves with different
wave vectors.

\section{Dispersion Relation}

The dispersion relation of the free massive scalar field is
\[
    \omega^2=|\vect{k}|^2+m^2.
\]
This relation determines how different Fourier modes oscillate. It is the field
theory version of the relativistic energy-momentum relation
\[
    E^2=|\vect{p}|^2+m^2.
\]
The analogy becomes exact in quantum field theory, where field modes are
quantized and interpreted in terms of particles. In the present classical
context, it is already useful: it tells us how frequency depends on spatial
wave number.

The phase velocity is
\[
    v_{\mathrm{ph}}=\frac{\omega}{|\vect{k}|}
    =\sqrt{1+\frac{m^2}{|\vect{k}|^2}}.
\]
For \(m\neq0\), this is larger than one. This does not violate relativity,
because phase velocity is not the signal velocity. The group velocity is
\[
    v_{\mathrm{g}}=\frac{d\omega}{d|\vect{k}|}
    =\frac{|\vect{k}|}{\sqrt{|\vect{k}|^2+m^2}}.
\]
For massive modes,
\[
    v_{\mathrm{g}}<1.
\]
In the high-momentum limit \(|\vect{k}|\gg m\), the group velocity approaches
one. In the low-momentum limit \(|\vect{k}|\ll m\), the group velocity is small.

\begin{remarkbox}
The massive Klein--Gordon field is dispersive: different wavelengths propagate
with different group velocities. The massless field in flat spacetime has
\(\omega=|\vect{k}|\), so wave packets propagate without this mass-induced
dispersion.
\end{remarkbox}

Dispersion is a crucial qualitative feature. A localized massive wave packet
spreads because its Fourier components travel with different velocities. This
is already visible at the classical level, before quantization.

\section{Energy and Momentum of the Scalar Field}

The canonical stress-energy tensor for the scalar field is
\[
    T^{\mu\nu}=\partial^\mu\phi\partial^\nu\phi-\eta^{\mu\nu}\mathcal{L}.
\]
For the Lagrangian density
\[
    \mathcal{L}=\frac{1}{2}\partial_\rho\phi\partial^\rho\phi-V(\phi),
\]
the energy density is
\[
    T^{00}
    =\frac{1}{2}(\partial_t\phi)^2
     +\frac{1}{2}|\nabla\phi|^2
     +V(\phi).
\]
For the free massive field,
\[
    V(\phi)=\frac{1}{2}m^2\phi^2,
\]
so
\[
    T^{00}
    =\frac{1}{2}\dot\phi^2
     +\frac{1}{2}|\nabla\phi|^2
     +\frac{1}{2}m^2\phi^2.
\]
This is positive definite for real \(\phi\) and \(m^2>0\).

The total energy is
\[
    E=\int d^d x\,T^{00}.
\]
The momentum density is
\[
    T^{0i}=\partial^0\phi\partial^i\phi.
\]
Since \(\partial^0=\partial_t\) and \(\partial^i=-\partial_i\), signs depend on
index placement. A commonly used spatial momentum is
\[
    P^i=\int d^d x\,\dot\phi\,\partial_i\phi
\]
up to the convention for raising or lowering the spatial index. The important
structural statement is that spacetime translation invariance gives a conserved
stress-energy tensor:
\[
    \partial_\mu T^{\mu\nu}=0.
\]

\begin{definitionbox}
For a scalar field with potential \(V(\phi)\), the energy density is
\[
    \mathcal{E}=\frac{1}{2}\dot\phi^2+
    \frac{1}{2}|\nabla\phi|^2+V(\phi).
\]
Finite-energy configurations must usually approach minima of \(V\) sufficiently
rapidly at spatial infinity.
\end{definitionbox}

The condition of finite energy is physically important. If space is noncompact,
one cannot allow arbitrary field behavior at infinity. For finite total energy,
\(\nabla\phi\) and \(V(\phi)\) must be integrable over space. This often forces
the field to approach a vacuum value at infinity.

\section{Green's Functions for the Klein--Gordon Operator}

The Klein--Gordon operator is
\[
    \Box+m^2.
\]
A Green's function for this operator is a distribution \(G(x-y)\) satisfying
\[
    (\Box_x+m^2)G(x-y)=\delta^{(d+1)}(x-y).
\]
Green's functions allow one to solve inhomogeneous equations of the form
\[
    (\Box+m^2)\phi(x)=J(x),
\]
where \(J(x)\) is a source.

Formally, a solution is
\[
    \phi(x)=\int d^{d+1}y\,G(x-y)J(y),
\]
provided the Green's function and boundary conditions are chosen appropriately.
There is not just one Green's function. The operator admits different inverses
corresponding to different causal or boundary prescriptions.

The retarded Green's function \(G_{\mathrm{ret}}\) satisfies
\[
    G_{\mathrm{ret}}(x-y)=0
    \quad\text{unless }x\text{ lies in or on the future light cone of }y.
\]
It describes causal response: the field at \(x\) depends only on sources in its
past. The advanced Green's function has support in the opposite direction and is
usually not used for causal initial-value problems.

\begin{definitionbox}
A \emph{retarded Green's function} is a Green's function whose support enforces
causal propagation from source to response. It gives the solution determined by
past sources, not future ones.
\end{definitionbox}

In Fourier representation, one formally writes
\[
    G(x)=\int \frac{d^{d+1}k}{(2\pi)^{d+1}}\,
    \frac{e^{-ik_\mu x^\mu}}{-k_\mu k^\mu+m^2}
\]
with a prescription for how to treat the poles. Different pole prescriptions
produce different Green's functions. This issue becomes central in quantum field
theory, where the Feynman propagator appears.

\section{Sources and Response}

A scalar field can be coupled linearly to an external source \(J(x)\) by adding
a source term to the Lagrangian density:
\[
    \mathcal{L}
    =\frac{1}{2}\partial_\mu\phi\partial^\mu\phi
     -\frac{1}{2}m^2\phi^2
     +J\phi.
\]
Varying the action gives
\[
    (\Box+m^2)\phi=J.
\]
The sign of the source term is a convention; what matters is consistency between
the Lagrangian and the resulting field equation.

For a static source, the equation becomes
\[
    (-\nabla^2+m^2)\phi(\vect{x})=J(\vect{x}).
\]
For a point source in three spatial dimensions,
\[
    J(\vect{x})=q\delta^{(3)}(\vect{x}),
\]
the solution is proportional to the Yukawa potential,
\[
    \phi(r)\propto \frac{e^{-mr}}{r}.
\]
This shows explicitly that the massive scalar field mediates a short-range
response with characteristic length \(1/m\). In the massless limit, the response
becomes Coulomb-like:
\[
    \phi(r)\propto \frac{1}{r}.
\]

\begin{examplebox}
The massive scalar Green's function in the static three-dimensional problem
shows the physical meaning of mass: mass suppresses long-distance response.
The field sourced at one point decays exponentially rather than purely as
\(1/r\).
\end{examplebox}

The source formalism is useful even when the source is not physical. It lets us
probe the response of a field theory and define Green's functions operationally:
the Green's function is the response to a point source.

\section{Interacting Scalar Fields}

The free scalar field is linear. Interactions are introduced by replacing the
quadratic potential with a more general potential:
\[
    \mathcal{L}=\frac{1}{2}\partial_\mu\phi\partial^\mu\phi - V(\phi).
\]
The field equation is
\[
    \Box\phi+\frac{dV}{d\phi}=0.
\]
If \(V\) is not purely quadratic, the equation is nonlinear.

Nonlinearity changes the structure of the theory dramatically. Superposition no
longer holds: if \(\phi_1\) and \(\phi_2\) are solutions, their sum is generally
not a solution. Interactions allow energy transfer between modes, nonlinear wave
phenomena, solitons in some models, spontaneous symmetry breaking in suitable
potentials, and nontrivial perturbation theory.

A common polynomial potential is
\[
    V(\phi)=\frac{1}{2}m^2\phi^2+\frac{\lambda}{4}\phi^4.
\]
The equation of motion is then
\[
    \Box\phi+m^2\phi+\lambda\phi^3=0.
\]
This is the classical \(\phi^4\) equation with a single-well potential when
\(m^2>0\) and \(\lambda>0\).

\begin{remarkbox}
The free theory is exactly solvable because Fourier modes evolve independently.
Interactions couple modes to each other. This mode coupling is the classical
ancestor of interaction vertices in quantum field theory.
\end{remarkbox}

Interactions must be chosen with care. A formally Lorentz-invariant equation may
still define an unstable theory if its potential is unbounded below.

\section{\texorpdfstring{\(\phi^4\)}{phi4} Theory as a Classical Model}

The name \(\phi^4\) theory refers to scalar field models with a quartic
self-interaction. A standard Lagrangian density is
\[
    \mathcal{L}
    =\frac{1}{2}\partial_\mu\phi\partial^\mu\phi
     -\left(\frac{1}{2}m^2\phi^2+\frac{\lambda}{4}\phi^4\right).
\]
The field equation is
\[
    \Box\phi+m^2\phi+\lambda\phi^3=0.
\]
For \(m^2>0\) and \(\lambda>0\), the potential has a single minimum at
\(\phi=0\). Small oscillations around \(\phi=0\) behave approximately like a free
massive scalar field.

Another important form is the double-well potential
\[
    V(\phi)=\frac{\lambda}{4}(\phi^2-v^2)^2,
    \qquad \lambda>0.
\]
This potential has two minima,
\[
    \phi=\pm v.
\]
The equation of motion is
\[
    \Box\phi+\lambda\phi(\phi^2-v^2)=0.
\]
This model has degenerate vacua and supports nontrivial classical structures in
one spatial dimension, such as kink solutions.

\begin{examplebox}
For the double-well potential, finite-energy configurations in one spatial
dimension may approach different vacua at opposite infinities:
\[
    \phi(x\to-\infty)=-v,
    \qquad
    \phi(x\to+\infty)=v.
\]
Such configurations cannot be continuously deformed to the trivial vacuum while
preserving finite energy and boundary behavior.
\end{examplebox}

Even when this course does not solve kink equations in detail, the lesson is
important: nonlinear scalar fields can have qualitatively different sectors of
solutions distinguished by boundary conditions.

\section{Stability of the Potential}

The energy density of a scalar field with potential \(V\) is
\[
    \mathcal{E}=\frac{1}{2}\dot\phi^2+\frac{1}{2}|\nabla\phi|^2+V(\phi).
\]
For the theory to have a stable ground state, the potential should be bounded
below. If \(V(\phi)\to-\infty\) as \(|\phi|\to\infty\), the energy can decrease
without bound, and the classical theory has no stable vacuum.

For example,
\[
    V(\phi)=\frac{1}{2}m^2\phi^2+\frac{\lambda}{4}\phi^4
\]
is bounded below when \(\lambda>0\). If \(\lambda<0\), it is unbounded below at
large \(|\phi|\), even if \(m^2>0\). Thus the sign of the highest-power term is
crucial.

A vacuum is a constant field configuration \(\phi_0\) minimizing the potential:
\[
    V'(\phi_0)=0,
    \qquad
    V''(\phi_0)>0.
\]
The value of \(V(\phi_0)\) is the vacuum energy density in this classical model.
Often one shifts the potential by a constant so that the minimum has zero
energy:
\[
    V(\phi_0)=0.
\]
This shift does not affect the scalar equation of motion in flat spacetime,
because the equation depends on \(V'\), not on the absolute value of \(V\).

\begin{definitionbox}
A \emph{classical vacuum} of a scalar field model is a constant field value
\(\phi_0\) at which the potential is minimized. Small disturbances around a
stable vacuum have positive squared mass \(V''(\phi_0)>0\).
\end{definitionbox}

In gravity, absolute vacuum energy matters because it contributes to the
stress-energy tensor. In nongravitational flat-spacetime field theory, only
differences in potential energy usually affect the classical scalar equation.

\section{Small Fluctuations Around a Background Solution}

Let \(\bar\phi(x)\) be a background solution of the interacting scalar equation
\[
    \Box\bar\phi+V'(\bar\phi)=0.
\]
Now consider a small fluctuation \(\eta(x)\) around this background:
\[
    \phi(x)=\bar\phi(x)+\eta(x),
\]
where \(\eta\) is assumed small. Substituting into the field equation gives
\[
    \Box(\bar\phi+\eta)+V'(\bar\phi+\eta)=0.
\]
Expanding the potential derivative to first order,
\[
    V'(\bar\phi+\eta)=V'(\bar\phi)+V''(\bar\phi)\eta+O(\eta^2).
\]
Using the background equation, the linearized equation becomes
\[
    \left[\Box+V''(\bar\phi(x))\right]\eta=0.
\]
Thus small fluctuations see an effective spacetime-dependent mass squared
\[
    m_{\mathrm{eff}}^2(x)=V''(\bar\phi(x)).
\]

For a constant vacuum \(\bar\phi=\phi_0\), this reduces to
\[
    (\Box+m_{\mathrm{eff}}^2)\eta=0,
    \qquad
    m_{\mathrm{eff}}^2=V''(\phi_0).
\]
This formula is central: the mass of small excitations around a vacuum is the
curvature of the potential at that vacuum.

\begin{examplebox}
For
\[
    V(\phi)=\frac{\lambda}{4}(\phi^2-v^2)^2,
\]
one has
\[
    V''(\phi)=\lambda(3\phi^2-v^2).
\]
At the vacua \(\phi_0=\pm v\),
\[
    m_{\mathrm{eff}}^2=V''(\pm v)=2\lambda v^2.
\]
Thus fluctuations around either vacuum are massive.
\end{examplebox}

Linearization is not just an approximation technique. It is how one reads the
local particle-like content of a classical field model, identifies stability,
and studies propagation on nontrivial backgrounds.

\begin{summarybox}
The real scalar field is the simplest relativistic field model. The free theory
is governed by the Klein--Gordon equation, has plane-wave solutions with
\(\omega^2=|\vect{k}|^2+m^2\), and carries positive energy when the potential is
bounded below. Sources are handled using Green's functions, while interactions
come from nonlinear potentials. Around a vacuum \(\phi_0\), the mass of small
fluctuations is determined by the curvature of the potential:
\(m_{\mathrm{eff}}^2=V''(\phi_0)\).
\end{summarybox}
