\chapter{The Action Principle for Fields}

\begin{remarkbox}
The action principle is the central organizing idea of classical field theory.
Instead of postulating a differential equation directly, one postulates an
action functional. The field equations are then obtained by demanding that the
action be stationary under arbitrary allowed variations of the fields.
\end{remarkbox}

\section{The Action Functional}

In classical mechanics, the action is a functional of a path \(q(t)\):
\[
    S[q]=\int_{t_1}^{t_2}L(q,\dot q,t)\,dt.
\]
A physical trajectory is selected by the condition
\[
    \delta S=0
\]
for variations of the path that obey the chosen endpoint conditions.

In field theory, the dynamical variable is not a finite list of coordinates but
a field, or a collection of fields,
\[
    \phi^A(x),
\]
where \(A\) labels the components or species of the field. The action is a
functional of the entire field configuration:
\[
    S[\phi]=\int_\Omega \mathcal{L}(\phi^A,\partial_\mu\phi^A,x)\,d^{d+1}x.
\]
Here \(\Omega\) is the spacetime region under consideration, and
\(\mathcal{L}\) is the Lagrangian density.

\begin{definitionbox}
An \emph{action functional} assigns a number to each allowed field
configuration. In a local first-order field theory it usually has the form
\[
    S[\phi]=\int_\Omega \mathcal{L}(\phi^A,\partial_\mu\phi^A,x)\,d^{d+1}x.
\]
\end{definitionbox}

The word functional is important. A function takes numbers as input and returns
a number. A functional takes a function as input and returns a number. The action
therefore depends on the entire history of the field over a spacetime region,
not merely on the value of the field at one point.

The principle of stationary action states that the physical fields are those
for which the first-order change of the action vanishes:
\[
    \delta S=0.
\]
This condition is not a minimum principle in general. The action may have a
minimum, a maximum, or a saddle point. The correct statement is stationarity.

\section{Lagrangian Density}

The Lagrangian density is the local quantity from which the action is built. If
space and time are separated, the ordinary Lagrangian is obtained by integrating
the Lagrangian density over space:
\[
    L(t)=\int d^d x\,\mathcal{L}(t,\vect{x}).
\]
Then
\[
    S=\int dt\,L(t)=\int dt\int d^d x\,\mathcal{L}.
\]

For a real scalar field in flat spacetime, a standard Lagrangian density is
\[
    \mathcal{L}=\frac{1}{2}\partial_\mu\phi\partial^\mu\phi
    -\frac{1}{2}m^2\phi^2.
\]
The first term is the kinetic-gradient term, and the second term is a mass term.
Using the mostly-minus convention, the first term expands as
\[
    \frac{1}{2}\partial_\mu\phi\partial^\mu\phi
    =\frac{1}{2}(\partial_t\phi)^2-\frac{1}{2}|\nabla\phi|^2.
\]
Thus the Lagrangian density contains a time-derivative part and a spatial-gradient
part with opposite signs.

A more general scalar-field Lagrangian often has the form
\[
    \mathcal{L}=\frac{1}{2}\partial_\mu\phi\partial^\mu\phi - V(\phi),
\]
where \(V(\phi)\) is the potential energy density. For example,
\[
    V(\phi)=\frac{1}{2}m^2\phi^2+\frac{\lambda}{4}\phi^4
\]
leads to a nonlinear scalar field equation.

\begin{remarkbox}
The Lagrangian density is not unique. Different Lagrangian densities can give
the same bulk field equations if they differ by a total derivative. However,
they may still differ in their boundary terms, canonical variables, and charges.
\end{remarkbox}

A Lagrangian density should be built from quantities compatible with the
symmetries one wants the theory to have. In relativistic field theory, this
usually means building Lorentz scalars by contracting all spacetime indices. In
gauge theory, it also means building expressions compatible with gauge symmetry.

\section{Variational Principle for Fields}

To vary a field, one considers a one-parameter family of field configurations
\[
    \phi^A_\varepsilon(x)=\phi^A(x)+\varepsilon\eta^A(x),
\]
where \(\varepsilon\) is a small parameter and \(\eta^A(x)\) is an arbitrary
allowed variation. The variation of the field is
\[
    \delta\phi^A(x)=\eta^A(x).
\]
The variation of a derivative is
\[
    \delta(\partial_\mu\phi^A)=\partial_\mu(\delta\phi^A).
\]

The first variation of the action is defined by
\[
    \delta S
    =\left.\frac{d}{d\varepsilon}S[\phi_\varepsilon]\right|_{\varepsilon=0}.
\]
For a first-order local Lagrangian density,
\[
    \mathcal{L}=\mathcal{L}(\phi^A,\partial_\mu\phi^A,x),
\]
we obtain
\[
    \delta S
    =\int_\Omega\left(
        \frac{\partial\mathcal{L}}{\partial\phi^A}\delta\phi^A
        +\frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}
        \delta(\partial_\mu\phi^A)
    \right)d^{d+1}x.
\]
Using \(\delta(\partial_\mu\phi^A)=\partial_\mu\delta\phi^A\), this becomes
\[
    \delta S
    =\int_\Omega\left(
        \frac{\partial\mathcal{L}}{\partial\phi^A}\delta\phi^A
        +\frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}
        \partial_\mu\delta\phi^A
    \right)d^{d+1}x.
\]
The repeated field label \(A\) and spacetime index \(\mu\) are summed over.

The second term contains a derivative of the variation. To isolate
\(\delta\phi^A\), we integrate by parts:
\[
\begin{aligned}
    \delta S
    ={}&\int_\Omega\left[
        \frac{\partial\mathcal{L}}{\partial\phi^A}
        -\partial_\mu\left(
            \frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}
        \right)
    \right]\delta\phi^A\,d^{d+1}x \\
    &+\int_{\partial\Omega}
        \frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}
        \delta\phi^A n_\mu\,d\Sigma.
\end{aligned}
\]
The first integral is the bulk variation. The second is the boundary variation.

\section{Euler--Lagrange Equations for Fields}

If the boundary term vanishes under the chosen class of variations, then the
stationarity condition \(\delta S=0\) implies that the bulk coefficient of each
independent \(\delta\phi^A\) must vanish. This gives the Euler--Lagrange
equations for fields:
\[
    \frac{\partial\mathcal{L}}{\partial\phi^A}
    -\partial_\mu\left(
        \frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}
    \right)=0.
\]
Equivalently,
\[
    \partial_\mu\left(
        \frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}
    \right)
    -\frac{\partial\mathcal{L}}{\partial\phi^A}=0.
\]
Both conventions are common; they differ only by multiplying the equation by
\(-1\).

\begin{definitionbox}
The \emph{Euler--Lagrange equation for a field} is the differential equation
obtained by setting the first variation of the action to zero for arbitrary
allowed variations of the field.
\end{definitionbox}

In functional-derivative notation, the same statement is
\[
    \frac{\delta S}{\delta\phi^A}=0.
\]
For a first-order local Lagrangian density,
\[
    \frac{\delta S}{\delta\phi^A}
    =\frac{\partial\mathcal{L}}{\partial\phi^A}
    -\partial_\mu\left(
        \frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}
    \right).
\]
This compact notation is useful, but it should not hide the boundary term that
appears during the derivation.

\section{Boundary Conditions in the Variational Principle}

The variational principle is incomplete until its boundary conditions are
specified. The integration by parts produces the boundary contribution
\[
    \int_{\partial\Omega}
        \frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi^A)}
        \delta\phi^A n_\mu\,d\Sigma.
\]
A common choice is to fix the fields on the boundary:
\[
    \delta\phi^A|_{\partial\Omega}=0.
\]
Then the boundary term vanishes, and the bulk Euler--Lagrange equations follow.

This choice is not the only possible one. One may instead impose falloff
conditions at infinity, periodic boundary conditions, Neumann-type boundary
conditions, or add boundary terms to the action so that the total variation is
well-defined. The correct choice depends on the physical problem.

\begin{examplebox}
For a vibrating string on an interval \([0,L]\), fixed endpoints correspond to
\[
    \delta\phi(t,0)=\delta\phi(t,L)=0.
\]
Free endpoints lead to different boundary conditions. The same bulk equation
can therefore describe different physical systems depending on the boundary
conditions.
\end{examplebox}

Boundary terms become especially important in gauge theory and gravity. In such
systems, surface terms may carry conserved charges, fluxes, or edge degrees of
freedom. For now, the essential point is that the action principle always
contains both a bulk statement and a boundary statement.

\section{Total Derivatives and Equivalent Lagrangians}

Suppose two Lagrangian densities differ by a total derivative:
\[
    \mathcal{L}'=\mathcal{L}+\partial_\mu K^\mu.
\]
Then the corresponding actions differ by a boundary term:
\[
    S'[\phi]-S[\phi]
    =\int_\Omega\partial_\mu K^\mu\,d^{d+1}x
    =\int_{\partial\Omega}K^\mu n_\mu\,d\Sigma.
\]
Therefore \(\mathcal{L}\) and \(\mathcal{L}'\) give the same bulk
Euler--Lagrange equations when the boundary term does not contribute to the
variation.

This is the precise sense in which two Lagrangians may be called equivalent.
They are equivalent for the bulk equations under a specified class of boundary
conditions. They are not necessarily equivalent as complete variational
principles if the boundary is physically relevant.

\begin{remarkbox}
The phrase ``up to a total derivative'' should always be read with care. It is
safe for deriving local bulk equations under suitable boundary assumptions. It
is not automatically safe for conserved charges, Hamiltonian generators,
asymptotic symmetries, or gravitational actions.
\end{remarkbox}

A simple mechanical analogue is adding a total time derivative to a Lagrangian:
\[
    L'(q,\dot q,t)=L(q,\dot q,t)+\frac{dF(q,t)}{dt}.
\]
The action changes by
\[
    F(q(t_2),t_2)-F(q(t_1),t_1),
\]
which does not affect the Euler--Lagrange equations if the endpoints are fixed.
Field theory is the spacetime version of the same idea.

\section{On-Shell and Off-Shell Statements}

A field configuration is said to be on shell if it satisfies the equations of
motion. It is off shell if it is an arbitrary field configuration not assumed to
satisfy those equations.

For example, for a scalar field with equation
\[
    (\Box+m^2)\phi=0,
\]
an on-shell field satisfies this differential equation. An off-shell field is
any sufficiently regular function \(\phi(x)\) in the configuration space of the
theory.

\begin{definitionbox}
An identity or formula is \emph{on shell} if it holds after imposing the field
equations. It is \emph{off shell} if it holds without using the field equations.
\end{definitionbox}

This distinction is fundamental. The variational formula for \(\delta S\) is an
off-shell statement. The Euler--Lagrange equations are obtained by demanding
stationarity and then setting the bulk functional derivative to zero. A
conservation law may hold only on shell, while a symmetry of the action is often
an off-shell statement.

The notation
\[
    \frac{\delta S}{\delta\phi^A}=0
\]
therefore defines the shell: it specifies the subspace of field configurations
that solve the equations of motion.

\section{Worked Example: Real Scalar Field}

Consider the real scalar field action in flat spacetime:
\[
    S[\phi]=\int \left(
        \frac{1}{2}\partial_\mu\phi\partial^\mu\phi
        -\frac{1}{2}m^2\phi^2
    \right)d^{d+1}x.
\]
The Lagrangian density is
\[
    \mathcal{L}
    =\frac{1}{2}\partial_\mu\phi\partial^\mu\phi
    -\frac{1}{2}m^2\phi^2.
\]
We compute
\[
    \frac{\partial\mathcal{L}}{\partial\phi}=-m^2\phi.
\]
The derivative with respect to \(\partial_\mu\phi\) is
\[
    \frac{\partial\mathcal{L}}{\partial(\partial_\mu\phi)}=\partial^\mu\phi.
\]
The Euler--Lagrange equation is therefore
\[
    -m^2\phi-\partial_\mu\partial^\mu\phi=0.
\]
Multiplying by \(-1\), we obtain
\[
    (\Box+m^2)\phi=0.
\]
This is the Klein--Gordon equation.

\begin{examplebox}
For \(m=0\), the equation becomes
\[
    \Box\phi=0,
\]
which is the relativistic wave equation. For \(m\neq 0\), the field has a mass
scale and the dispersion relation becomes \(\omega^2=|\vect{k}|^2+m^2\).
\end{examplebox}

If the Lagrangian density is instead
\[
    \mathcal{L}=\frac{1}{2}\partial_\mu\phi\partial^\mu\phi - V(\phi),
\]
then the field equation is
\[
    \Box\phi+\frac{dV}{d\phi}=0.
\]
This form will be useful for interacting scalar field models.

\section{Worked Example: Wave Equation from an Action}

The ordinary wave equation in one spatial dimension can be obtained from the
action
\[
    S[\phi]=\int dt\,dx\,
    \frac{1}{2}\left[(\partial_t\phi)^2-c^2(\partial_x\phi)^2\right].
\]
Here the Lagrangian density is
\[
    \mathcal{L}=\frac{1}{2}(\partial_t\phi)^2
    -\frac{c^2}{2}(\partial_x\phi)^2.
\]
There is no explicit dependence on \(\phi\), so
\[
    \frac{\partial\mathcal{L}}{\partial\phi}=0.
\]
The derivatives with respect to the field derivatives are
\[
    \frac{\partial\mathcal{L}}{\partial(\partial_t\phi)}=\partial_t\phi,
    \qquad
    \frac{\partial\mathcal{L}}{\partial(\partial_x\phi)}=-c^2\partial_x\phi.
\]
The Euler--Lagrange equation gives
\[
    \partial_t(\partial_t\phi)+\partial_x(-c^2\partial_x\phi)=0.
\]
Thus
\[
    \partial_t^2\phi-c^2\partial_x^2\phi=0.
\]
This is the wave equation.

This example shows the physical meaning of the two terms in the Lagrangian
density. The term \((\partial_t\phi)^2/2\) is associated with kinetic energy,
while \(c^2(\partial_x\phi)^2/2\) is associated with the elastic cost of spatial
variation. The sign difference between them is what gives a hyperbolic evolution
equation rather than an elliptic equation.

\section{Physical Interpretation of Field Equations}

The Euler--Lagrange equations are not just formal consequences of calculus of
variations. They express local balance laws for distributed degrees of freedom.
In a mechanical system, the equation of motion balances inertia against forces.
In a field theory, the equation balances time evolution, spatial gradients,
sources, and self-interactions.

For the scalar field equation
\[
    \Box\phi+\frac{dV}{d\phi}=0,
\]
the operator \(\Box\) describes propagation through spacetime, while
\(dV/d\phi\) describes the local restoring force or self-interaction coming from
the potential. For the wave equation,
\[
    \partial_t^2\phi-c^2\partial_x^2\phi=0,
\]
the acceleration of the field is determined by the curvature of its spatial
profile.

The action viewpoint also explains why field equations are so tightly connected
to symmetries. Once the dynamics is encoded in an action, transformations of the
fields and coordinates can be studied by asking whether they leave the action
invariant. This is the starting point of Noether's theorem, which is the subject
of the next chapter.

\begin{summarybox}
The action principle turns field dynamics into variational calculus. A local
Lagrangian density defines an action; varying the action gives a bulk term and a
boundary term; the vanishing of the bulk term gives the Euler--Lagrange field
equations. Boundary conditions are part of the variational problem, and total
derivatives must be interpreted relative to those boundary conditions.
\end{summarybox}
