\chapter{Hamiltonian Formulation of Field Theory}

\begin{remarkbox}
The Hamiltonian formulation rewrites field theory as dynamics on phase space.
It is less manifestly relativistic than the Lagrangian formulation, because it
chooses a time coordinate, but it reveals the canonical variables, Poisson
brackets, constraints, gauge freedom, and the role of energy as the generator of
time evolution.
\end{remarkbox}

\section{Canonical Momenta for Fields}

The Lagrangian formulation treats fields and their spacetime derivatives in a
nearly symmetric way. The Hamiltonian formulation begins by separating time from
space. We write spacetime coordinates as
\[
    x^\mu=(t,\vect{x}),
\]
and denote the time derivative of a field by
\[
    \dot{\phi}^A(t,\vect{x})=\partial_t\phi^A(t,\vect{x}).
\]
For a first-order Lagrangian density
\[
    \mathcal{L}=\mathcal{L}(\phi^A,\dot{\phi}^A,\partial_i\phi^A,t,\vect{x}),
\]
the canonical momentum conjugate to \(\phi^A\) is defined by
\[
    \pi_A(t,\vect{x})
    =\frac{\partial\mathcal{L}}{\partial\dot{\phi}^A(t,\vect{x})}.
\]
Thus each field component \(\phi^A\) has a conjugate momentum density
\(\pi_A\).

\begin{definitionbox}
The \emph{canonical momentum} conjugate to a field \(\phi^A\) is
\[
    \pi_A=\frac{\partial\mathcal{L}}{\partial\dot{\phi}^A}.
\]
It is a density on space and is evaluated at the same time and spatial point as
\(\phi^A\).
\end{definitionbox}

For a real scalar field with
\[
    \mathcal{L}
    =\frac{1}{2}(\partial_t\phi)^2
     -\frac{1}{2}|\nabla\phi|^2
     -V(\phi),
\]
the canonical momentum is simply
\[
    \pi=\frac{\partial\mathcal{L}}{\partial\dot\phi}=\dot\phi.
\]
This example is regular: the velocity \(\dot\phi\) can be recovered from the
momentum \(\pi\).

The situation is not always so simple. If the definition of the momenta cannot
be inverted to express all velocities in terms of fields and momenta, the theory
has constraints. Gauge theories and generally covariant theories almost always
have this feature. The Hamiltonian formalism is therefore not merely an
alternative notation; it is the natural language for understanding constraints
and gauge redundancy.

\section{Legendre Transform and Hamiltonian Density}

In mechanics, the Hamiltonian is obtained from the Lagrangian by a Legendre
transform:
\[
    H(q,p,t)=p_i\dot q^i-L(q,\dot q,t),
\]
where the velocities are expressed in terms of \((q,p)\). In field theory, the
same idea is applied pointwise in space. The Hamiltonian density is
\[
    \mathcal{H}
    =\pi_A\dot\phi^A-\mathcal{L}.
\]
After the velocities have been expressed in terms of \(\phi^A\), \(\pi_A\), and
spatial derivatives of \(\phi^A\), the Hamiltonian is
\[
    H=\int d^d x\,\mathcal{H}.
\]

The Legendre transform is valid in its simplest form only if the Hessian matrix
with respect to the velocities is nondegenerate:
\[
    W_{AB}
    =\frac{\partial^2\mathcal{L}}
    {\partial\dot\phi^A\partial\dot\phi^B}.
\]
If \(W_{AB}\) is invertible, the momenta determine the velocities locally. If it
is not invertible, the theory is singular and constraints appear.

\begin{definitionbox}
A field theory is \emph{regular} if the Legendre transform from velocities
\(\dot\phi^A\) to momenta \(\pi_A\) is locally invertible. It is \emph{singular}
if this transform is degenerate. Singular theories require constraints in the
Hamiltonian formulation.
\end{definitionbox}

For the real scalar field,
\[
    \pi=\dot\phi.
\]
Therefore
\[
\begin{aligned}
    \mathcal{H}
    &=\pi\dot\phi-\mathcal{L} \\
    &=\pi^2-\left[\frac{1}{2}\pi^2
      -\frac{1}{2}|\nabla\phi|^2-V(\phi)\right].
\end{aligned}
\]
Thus
\[
    \mathcal{H}
    =\frac{1}{2}\pi^2+\frac{1}{2}|\nabla\phi|^2+V(\phi).
\]
This is the energy density of the scalar field.

\begin{examplebox}
For a free massive scalar field,
\[
    V(\phi)=\frac{1}{2}m^2\phi^2,
\]
the Hamiltonian is
\[
    H=\int d^d x\left[
        \frac{1}{2}\pi^2
        +\frac{1}{2}|\nabla\phi|^2
        +\frac{1}{2}m^2\phi^2
    \right].
\]
This Hamiltonian is positive for real fields and suitable boundary conditions.
\end{examplebox}

\section{Equal-Time Poisson Brackets}

The Hamiltonian formulation requires a notion of Poisson bracket. For fields,
the canonical Poisson brackets are imposed at equal time. The fundamental
brackets are
\[
    \{\phi^A(t,\vect{x}),\phi^B(t,\vect{y})\}=0,
\]
\[
    \{\pi_A(t,\vect{x}),\pi_B(t,\vect{y})\}=0,
\]
and
\[
    \{\phi^A(t,\vect{x}),\pi_B(t,\vect{y})\}
    =\delta^A{}_B\delta^{(d)}(\vect{x}-\vect{y}).
\]
The delta distribution appears because there is one canonical degree of freedom
at each spatial point.

More generally, for two functionals \(F[\phi,\pi]\) and \(G[\phi,\pi]\), the
field-theoretic Poisson bracket is
\[
    \{F,G\}
    =\int d^d x\left(
        \frac{\delta F}{\delta\phi^A(\vect{x})}
        \frac{\delta G}{\delta\pi_A(\vect{x})}
        -
        \frac{\delta F}{\delta\pi_A(\vect{x})}
        \frac{\delta G}{\delta\phi^A(\vect{x})}
    \right).
\]
This is the infinite-dimensional analogue of the mechanical formula
\[
    \{F,G\}
    =\frac{\partial F}{\partial q^i}\frac{\partial G}{\partial p_i}
    -\frac{\partial F}{\partial p_i}\frac{\partial G}{\partial q^i}.
\]

\begin{definitionbox}
The \emph{equal-time Poisson bracket} is the canonical Poisson bracket on the
space of fields \(\phi^A(t,\vect{x})\) and conjugate momenta
\(\pi_A(t,\vect{x})\) at a fixed time \(t\).
\end{definitionbox}

Equal-time brackets are not manifestly Lorentz covariant, because one has chosen
a time coordinate and a family of spatial hypersurfaces. This is the price paid
for a canonical description. The reward is that time evolution, constraints, and
canonical quantization become transparent.

\section{Hamilton's Equations for Fields}

Once the Hamiltonian is known, time evolution is generated by Poisson brackets.
For any functional \(F[\phi,\pi,t]\), one has
\[
    \frac{dF}{dt}=\{F,H\}+\frac{\partial F}{\partial t}.
\]
Applying this to the basic fields gives Hamilton's equations:
\[
    \dot\phi^A(t,\vect{x})
    =\frac{\delta H}{\delta\pi_A(t,\vect{x})},
\]
and
\[
    \dot\pi_A(t,\vect{x})
    =-\frac{\delta H}{\delta\phi^A(t,\vect{x})}.
\]
These equations are equivalent to the Euler--Lagrange equations when the
Legendre transform is regular.

For the real scalar field,
\[
    H=\int d^d x\left[
        \frac{1}{2}\pi^2
        +\frac{1}{2}|\nabla\phi|^2
        +V(\phi)
    \right].
\]
The first Hamilton equation gives
\[
    \dot\phi=\frac{\delta H}{\delta\pi}=\pi.
\]
The second gives
\[
    \dot\pi=-\frac{\delta H}{\delta\phi}.
\]
To compute the functional derivative, vary the Hamiltonian:
\[
    \delta H=\int d^d x\left[
        \pi\delta\pi+
        \nabla\phi\cdot\nabla\delta\phi+
        V'(\phi)\delta\phi
    \right].
\]
After integrating the gradient term by parts and assuming the boundary term
vanishes,
\[
    \delta H=\int d^d x\left[
        \pi\delta\pi+
        (-\nabla^2\phi+V'(\phi))\delta\phi
    \right].
\]
Therefore
\[
    \frac{\delta H}{\delta\phi}=-\nabla^2\phi+V'(\phi),
\]
and
\[
    \dot\pi=\nabla^2\phi-V'(\phi).
\]
Using \(\pi=\dot\phi\), we obtain
\[
    \ddot\phi-\nabla^2\phi+V'(\phi)=0,
\]
which is the Euler--Lagrange equation for the scalar field.

\begin{examplebox}
For \(V(\phi)=m^2\phi^2/2\), Hamilton's equations imply
\[
    \ddot\phi-\nabla^2\phi+m^2\phi=0,
\]
that is,
\[
    (\Box+m^2)\phi=0.
\]
\end{examplebox}

\section{Phase Space of a Field Theory}

The configuration space of a field theory is a space of field configurations
\(\phi^A(\vect{x})\). The phase space is the space of pairs
\[
    (\phi^A(\vect{x}),\pi_A(\vect{x})).
\]
Thus a point of phase space specifies both the field profile and its canonical
momentum profile at one time.

This phase space is infinite-dimensional. In mechanics, a system with
\(n\) generalized coordinates has a \(2n\)-dimensional phase space with
coordinates \((q^i,p_i)\). In field theory, the spatial point \(\vect{x}\) plays
the role of a continuous label, so phase space consists of functions rather than
finite tuples of numbers.

The canonical symplectic structure is encoded in the equal-time Poisson
brackets. Equivalently, one may write the canonical symplectic form formally as
\[
    \Omega=\int d^d x\,\delta\pi_A(\vect{x})\wedge\delta\phi^A(\vect{x}).
\]
Here \(\delta\) denotes the exterior derivative on the infinite-dimensional
space of fields, not a spacetime variation. This notation is more advanced but
useful: it emphasizes that Hamiltonian field theory is symplectic geometry on
an infinite-dimensional space.

\begin{remarkbox}
A field-theoretic phase space is not the same thing as spacetime. Spacetime is
the domain on which fields are defined. Phase space is the space of all possible
instantaneous field and momentum configurations.
\end{remarkbox}

In a well-posed Hamiltonian evolution problem, initial data define a point in
phase space, and Hamilton's equations define a trajectory through phase space.
This is the canonical version of the initial-value problem for fields.

\section{Constraints: First Encounter}

A constraint is a relation among canonical variables that must hold on the
allowed phase space. Constraints arise when the canonical momenta are not all
independent. This happens when the Legendre transform is degenerate.

A simple diagnostic is the Hessian matrix
\[
    W_{AB}
    =\frac{\partial^2\mathcal{L}}
    {\partial\dot\phi^A\partial\dot\phi^B}.
\]
If \(W_{AB}\) has zero modes, then not all velocities can be solved in terms of
momenta. The momentum definitions imply relations
\[
    \chi_m(\phi,\pi)\approx 0.
\]
The symbol \(\approx\) is Dirac's weak equality. It means that the equation holds
on the constraint surface, but should not necessarily be imposed before Poisson
brackets are computed.

\begin{definitionbox}
A \emph{constraint} is a relation \(\chi(\phi,\pi)\approx 0\) that restricts the
allowed canonical data. The set of points satisfying all constraints is called
the constraint surface.
\end{definitionbox}

Electromagnetism gives the standard example. The Maxwell Lagrangian density is
\[
    \mathcal{L}=-\frac{1}{4}F_{\mu\nu}F^{\mu\nu}.
\]
The field component \(A_0\) has no time derivative in \(F_{\mu\nu}\) in the same
way as the spatial components. Its canonical momentum is
\[
    \pi^0=\frac{\partial\mathcal{L}}{\partial\dot A_0}=0.
\]
This is a constraint:
\[
    \pi^0\approx 0.
\]
It reflects the fact that \(A_0\) is not an ordinary propagating degree of
freedom. It is tied to Gauss's law and gauge freedom.

\section{Primary and Secondary Constraints}

Constraints that arise directly from the definition of the canonical momenta are
called primary constraints. If
\[
    \chi_m(\phi,\pi)\approx 0
\]
is a primary constraint, it must remain true during time evolution. Therefore
one imposes the consistency condition
\[
    \dot\chi_m\approx 0.
\]
Using the Hamiltonian, this becomes
\[
    \{\chi_m,H\}\approx 0,
\]
or, more generally, a condition involving the total Hamiltonian.

The total Hamiltonian includes the canonical Hamiltonian plus primary
constraints multiplied by arbitrary functions called Lagrange multipliers:
\[
    H_T=H_c+\int d^d x\,u^m(\vect{x})\chi_m(\vect{x}).
\]
The functions \(u^m\) are not determined by the Legendre transform. Their
presence signals that the evolution may not be unique until gauge freedom is
fixed.

The consistency condition may lead to three possibilities:
\begin{enumerate}
    \item it may be automatically satisfied;
    \item it may determine some Lagrange multipliers;
    \item it may produce new constraints, called secondary constraints.
\end{enumerate}

In electromagnetism, the primary constraint \(\pi^0\approx0\) leads, through
consistency, to Gauss's law as a secondary constraint:
\[
    \partial_i\pi^i-J^0\approx0,
\]
up to sign conventions. This is the Hamiltonian form of the electric charge
constraint.

\begin{remarkbox}
The distinction between primary and secondary constraints is procedural, not
physical. Primary constraints arise from the momentum definitions; secondary
constraints arise from demanding preservation of constraints under time
evolution.
\end{remarkbox}

The systematic treatment of constraints is known as Dirac's constraint
algorithm. This course only needs the first conceptual layer: constraints are
not optional extra equations. They are part of the canonical structure of
singular Lagrangian systems.

\section{Gauge Freedom and Degenerate Legendre Transform}

Gauge freedom and degeneracy of the Legendre transform are closely related. If a
Lagrangian has a local gauge symmetry, then some apparent field components do
not represent independent physical degrees of freedom. This redundancy appears
in the Hamiltonian formulation as constraints.

The intuitive reason is simple. Gauge-related configurations describe the same
physical state. Therefore the Lagrangian cannot assign independent physical
velocities to all field components. Some directions in velocity space correspond
not to physical motion but to motion along gauge orbits. Along such directions,
the Legendre transform is degenerate.

Constraints are often classified as first class or second class. A constraint is
first class if its Poisson bracket with all constraints vanishes on the
constraint surface. Schematically,
\[
    \{\chi_m,\chi_n\}\approx0.
\]
First-class constraints generate gauge transformations. Second-class constraints
do not generate gauge freedom; instead, they remove variables in a different
way and require replacing Poisson brackets by Dirac brackets.

\begin{definitionbox}
A \emph{first-class constraint} has weakly vanishing Poisson brackets with all
constraints. First-class constraints are associated with gauge transformations.
A \emph{second-class constraint} is a constraint that fails this condition.
\end{definitionbox}

For gauge theories, the true physical phase space is not the full constraint
surface. One must also identify points related by gauge transformations. In
words,
\[
    \text{physical phase space}
    =\frac{\text{constraint surface}}{\text{gauge transformations}}.
\]
This quotient viewpoint is essential in advanced field theory, gauge theory,
and gravity.

\section{Energy as a Generator of Time Evolution}

In Hamiltonian mechanics, the Hamiltonian generates time evolution. The same is
true in field theory. If \(F[\phi,\pi,t]\) is a phase-space functional, then
\[
    \frac{dF}{dt}=\{F,H\}+\frac{\partial F}{\partial t}
\]
for regular theories. Thus the Hamiltonian is not merely the numerical energy;
it is the generator of time translations on phase space.

For the basic fields,
\[
    \dot\phi^A=\{\phi^A,H\},
    \qquad
    \dot\pi_A=\{\pi_A,H\}.
\]
These are Hamilton's equations in bracket form. They show that the Hamiltonian
vector field associated with \(H\) gives the dynamical flow on phase space.

If the Hamiltonian has no explicit time dependence and the boundary conditions
allow the relevant integrations by parts, then
\[
    \frac{dH}{dt}=\{H,H\}=0.
\]
Thus energy is conserved. This is the Hamiltonian expression of time-translation
symmetry.

\begin{examplebox}
For the free scalar field, the Hamiltonian
\[
    H=\int d^d x\left[
        \frac{1}{2}\pi^2
        +\frac{1}{2}|\nabla\phi|^2
        +\frac{1}{2}m^2\phi^2
    \right]
\]
generates the Klein--Gordon evolution through the equal-time Poisson brackets.
\end{examplebox}

In constrained systems, the generator of time evolution is the total Hamiltonian
\(H_T\), not merely the canonical Hamiltonian. The arbitrary multipliers in
\(H_T\) encode gauge freedom. Therefore Hamiltonian evolution in a gauge theory
is unique only up to gauge transformations.

\section{Comparison with the Lagrangian Formulation}

The Lagrangian and Hamiltonian formulations describe the same classical dynamics
when the Legendre transform is regular and boundary conditions are handled
properly. However, they emphasize different structures.

The Lagrangian formulation is manifestly spacetime-oriented. It is usually the
best language for relativistic covariance, action principles, Noether's
theorems, path integrals, and coupling to geometry. A typical Lagrangian action
has the compact form
\[
    S[\phi]=\int \mathcal{L}(\phi^A,\partial_\mu\phi^A,x)\,d^{d+1}x.
\]
The Euler--Lagrange equations follow from varying \(S\).

The Hamiltonian formulation is phase-space-oriented. It requires a choice of
time coordinate and describes evolution of canonical data
\[
    (\phi^A(\vect{x}),\pi_A(\vect{x})).
\]
Its central objects are the Hamiltonian, Poisson brackets, constraints, and
canonical transformations. It is the natural starting point for canonical
quantization, constraint analysis, and the study of generators.

\begin{center}
\begin{tabular}{lll}
\toprule
Feature & Lagrangian formulation & Hamiltonian formulation \\
\midrule
Basic variables & \(\phi^A,\partial_\mu\phi^A\) & \(\phi^A,\pi_A\) \\
Central object & action \(S\) & Hamiltonian \(H\) \\
Dynamics & Euler--Lagrange equations & Hamilton's equations \\
Geometry & spacetime/covariant & phase space/canonical \\
Symmetries & Noether currents & generators via Poisson brackets \\
Gauge theories & variational identities & constraints and gauge orbits \\
\bottomrule
\end{tabular}
\end{center}

Neither formulation is intrinsically superior. The Lagrangian formulation makes
covariance and the action principle transparent. The Hamiltonian formulation
makes the phase-space structure and constraints transparent. A mature use of
classical field theory requires both.

\begin{summarybox}
The Hamiltonian formulation begins by choosing a time coordinate, defining
canonical momenta, and performing a Legendre transform. For regular theories it
leads to a Hamiltonian density, equal-time Poisson brackets, and Hamilton's
equations. For singular theories it reveals constraints, gauge freedom, and the
need for Dirac's constraint analysis. The Hamiltonian is both energy and the
generator of time evolution, provided the boundary conditions make the generator
well defined.
\end{summarybox}
